---
title: "16S SOD Preparation Steps"
output: html_document
date: "2024-06-11"
---

```{r setup}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(root.dir = "~/Documents/Postdoc/SOD_microbiome_analysis")
library(phyloseq); packageVersion("phyloseq")
library(ggplot2); packageVersion("ggplot2")
library(readr); packageVersion("readr")
library(purrr); packageVersion("purrr")
library(furrr); packageVersion("furrr")
library(dplyr); packageVersion("dplyr")
library(stringr); packageVersion("stringr")
library(metacoder); packageVersion("metacoder")
library(data.table); packageVersion("data.table")
library(decontam); packageVersion("decontam")
library(vegan); packageVersion("vegan")
```

### Load data
```{r load files, include=TRUE}
seqtab_nochim <- read.table("Data/16S/dada2_outputs/16S_seqtab_nochim.out",sep="\t",stringsAsFactors=F,header=T)

taxa <- read.table("Data/16S/dada2_outputs/16S_taxa_species_separate_minboot60.out",sep="\t",stringsAsFactors=F,header=T)

taxa_filt <- taxa %>%
  mutate(across(starts_with("tax."), ~ ifelse(get(str_replace(cur_column(), "tax.", "boot.")) < 0, NA, .))) %>%
  set_names(str_replace(names(.), "tax.", ""))

revise_species <- function(genus, species) {
  # Combine genus and species with an underscore if species is not NA
  revised_species <- ifelse(is.na(species), NA, paste(genus, species, sep = "_"))
  return(revised_species)
}

taxa$Species <- revise_species(taxa$Genus, taxa_filt$Species)

samdf <- read.csv("Data/16S/metadata_dbs/Metadata_16S_short_mod.csv")
```

### Track number of reads throughout the pipeline
```{r track reads through DADA2 analysis, include=TRUE}
track <- read.table("Data/16S/dada2_outputs/16S_trackreads.out", sep="\t", stringsAsFactors = F, header=T)

track_trim<-track[!(track$Sample %in% c("emptyextraction", "HD4390trim-S1", "HD4397trim-S15", "NTC")), ]
```

### Format dataframes to create phyloseq object
```{r format dataframes, include=TRUE}
all(rownames(seqtab_nochim) %in% samdf$Sample_name16S) 
rownames(samdf) <- samdf$Sample_name16S

samdata.ps <- sample_data(samdf)

taxa.ps <- tax_table(as.matrix(taxa))

seqtab_nochim.ps <- otu_table(seqtab_nochim, taxa_are_rows=FALSE)
ps <- phyloseq(seqtab_nochim.ps, samdata.ps, taxa.ps)

# Add refseq slot
dna <- Biostrings::DNAStringSet(taxa_names(ps))
names(dna) <- taxa_names(ps)
ps <- merge_phyloseq(ps, dna)
taxa_names(ps) <- paste0("ASV", seq(ntaxa(ps)))

# I randomly subsampled from two samples that had way more reads. For now I will remove these samples, but it is always an option to use the subsampled samples
ps_filtered<-subset_samples(ps, ps@sam_data@row.names != "HD4397trim-S15")
ps_filtered<-subset_samples(ps_filtered, ps@sam_data@row.names != "HD4390trim-S1")
```

### Decontamination step using decontam package-this was most necessary for 16S analysis.
```{r decontam, include=TRUE}
df <- as.data.frame(sample_data(ps_filtered)) # Put sample_data into a ggplot-friendly data.frame
df$LibrarySize <- sample_sums(ps_filtered)
df <- df[order(df$LibrarySize),]
df$Index <- seq(nrow(df))
ggplot(data=df, aes(x=Index, y=LibrarySize, color=Sample_or_control)) + geom_point()

sample_data(ps_filtered)$is.neg <- sample_data(ps_filtered)$Sample_or_control == "control"
contamdf.prev <- isContaminant(ps_filtered, method="prevalence", neg="is.neg")
table(contamdf.prev$contaminant)
head(which(contamdf.prev$contaminant))

contamdf.prev05 <- isContaminant(ps_filtered, method="prevalence", neg="is.neg", threshold=0.5)
table(contamdf.prev05$contaminant)
which(contamdf.prev05$contaminant)

ps.pa <- transform_sample_counts(ps_filtered, function(abund) 1*(abund>0))
ps.pa.neg <- prune_samples(sample_data(ps.pa)$Sample_or_control == "control", ps.pa)
ps.pa.pos <- prune_samples(sample_data(ps.pa)$Sample_or_control == "sample", ps.pa)

df.pa <- data.frame(pa.pos=taxa_sums(ps.pa.pos), pa.neg=taxa_sums(ps.pa.neg),
                    contaminant=contamdf.prev05$contaminant)
ggplot(data=df.pa, aes(x=pa.neg, y=pa.pos, color=contaminant)) + geom_point() +
  xlab("Prevalence (Negative Controls)") + ylab("Prevalence (True Samples)")

ps.decontam <- prune_taxa(!contamdf.prev05$contaminant, ps_filtered)

ps.decontam <- ps.decontam %>%
  subset_taxa(
    Kingdom == "Bacteria" &
      Family  != "Mitochondria" &
      Class   != "Chloroplast"
  )

save(ps.decontam, file = "Data/16S/filtered_data/16S_phyloseq.decontam.RData")
```

### Take a first glance at the data  
Helpful for deciding how to filter data. Prevalence is how many samples have the ASV, and total abundance is how many times ASV is present across all samples. 
```{r examine ASV Prevalance, include=TRUE}
prevdf = apply(X = phyloseq::otu_table(ps.decontam),
               MARGIN = ifelse(taxa_are_rows(ps.decontam), yes = 1, no = 2),
               FUN = function(x){sum(x > 0)})

prevdf = data.frame(Prevalence = prevdf,
                    TotalAbundance = taxa_sums(ps.decontam),
                    tax_table(ps.decontam))
```

### Subset SOD-specific samples
```{r Subset dataset, include=TRUE}
sod.ps<-subset_samples(ps.decontam, Experiment=="basetree_100plots")
sod.ps<- prune_taxa(taxa_sums(sod.ps) > 0, sod.ps)

sod.ps@sam_data$Treatuntreat_EP<-factor(sod.ps@sam_data$Treatuntreat_EP, levels = c(0, 1), labels = c("untreated", "treated"))
sod.ps@sam_data$YearSam_EP <- as.factor(sod.ps@sam_data$YearSam_EP)
sod.ps@sam_data$YrSinceTrt_BR <- as.factor(sod.ps@sam_data$YrSinceTrt_BR)
sod.ps@sam_data$YrSinceTrt_BR <- factor(sod.ps@sam_data$YrSinceTrt_BR, levels = c("untreated", "1to2", "3to4", "5to8", "10to12"))
sod.ps@sam_data$YrSinceTrt_Range <- as.factor(sod.ps@sam_data$YrSinceTrt_Range)
sod.ps@sam_data$YrSinceTrt_Range <- factor(sod.ps@sam_data$YrSinceTrt_Range, levels = c("untreated", "<3", ">3"))
sod.ps@sam_data$Fire_Mod <- factor(sod.ps@sam_data$Fire_Mod, levels = c("untreated", "wildfire", "SODtrt"))
sod.ps@sam_data$VegBin <- as.factor(sod.ps@sam_data$VegBin)
sod.ps@sam_data$SoilBin <- as.factor(sod.ps@sam_data$SoilBin)

readsums<-sort(sample_sums(sod.ps))
median(readsums)

sod_5asv.ps <- prune_taxa(taxa_sums(sod.ps) > 5, sod.ps)
sod_5asv.ps<- prune_taxa(taxa_sums(sod_5asv.ps) > 0, sod_5asv.ps)

save(sod.ps, file="Data/16S/filtered_data/sodexp.RData")
save(sod_5asv.ps, file="Data/16S/filtered_data/sod_5asv.RData")

readsums<-sort(sample_sums(sod.ps))
median(readsums)
summary(readsums)
```

### Subset read counts so we are just looking at sod samples
```{r subset tracked reads}
samples_to_subset <- samdf %>%
  filter(Experiment=="basetree_100plots") %>% 
  pull(Sample_name16S)
# Filter the track dataframe

#track$sample <- rownames(track)
subset_track_reads <- track[rownames(track) %in% samples_to_subset, ]
write.csv(subset_track_reads, file="Data/16S/filtered_data/16S_subset_track_reads_sod.csv",row.names = TRUE)
# Print summary of the filtered track dataframe
summary(subset_track_reads)
```

### Load data (optional) to avoid rerunning above analyses
```{r load sod data}
load("Data/16S/filtered_data/sodexp.RData")
load("Data/16S/filtered_data/sod_5asv.RData")

filtered_read_counts<-read.csv("Data/16S/filtered_data/16S_subset_track_reads_sod.csv")
summary(filtered_read_counts)

new_names <- sapply(strsplit(sod_5asv.ps@sam_data$Sample_name16S, "-"), function(x) {
  substr(x[1], 3, nchar(x[1]))
})

sod_5asv.ps@sam_data$Sample_name16S <- new_names
```

### Check prevalence after filtering
```{r check prevalence dataframes after filterig}
# Double check that the filtering makes sense by making new prevalence dataframe
prevdf_check = apply(X = phyloseq::otu_table(sod_5asv.ps),
               MARGIN = ifelse(taxa_are_rows(sod_5asv.ps), yes = 1, no = 2),
               FUN = function(x){sum(x > 0)})

prevdf_check = data.frame(Prevalence = prevdf_check,
                    TotalAbundance = taxa_sums(sod_5asv.ps),
                    tax_table(sod_5asv.ps))
```

### Create prevalance plots (Rel abundance vs. total abundance) faceted by phylum
```{r prevalence plots organized by phylum}
prevdf1 = subset(prevdf_check, Phylum %in% get_taxa_unique(sod_5asv.ps, "Phylum"))
ggplot(prevdf1, aes(TotalAbundance, Prevalence / nsamples(sod_5asv.ps),color=Phylum)) +
  geom_hline(yintercept = 0.005, alpha = 0.5, linetype = 2) +  geom_point(size = 2, alpha = 0.7) +
  scale_x_log10() +  xlab("Total Abundance") + ylab("Prevalence [Frac. Samples]") +
  facet_wrap(~Phylum) + theme(legend.position="none")
```

### Create prevalance plots (Rel abundance vs. total abundance) faceted by phylum-low abundance data not filtered
```{r prevalence plots organized by phylum-data not filtered}
prevdf_check = apply(X = phyloseq::otu_table(sod.ps),
               MARGIN = ifelse(taxa_are_rows(sod.ps), yes = 1, no = 2),
               FUN = function(x){sum(x > 0)})

prevdf_check = data.frame(Prevalence = prevdf_check,
                    TotalAbundance = taxa_sums(sod.ps),
                    tax_table(sod.ps))
```

### Create prevalance plots (Rel abundance vs. total abundance) faceted by phylum data not filtered
```{r prevalence plots organized by phylum 2}
prevdf1 = subset(prevdf_check, Phylum %in% get_taxa_unique(sod.ps, "Phylum"))

pdf("Figures/16S/exploratoryplot_abundance.pdf", width = 10, height = 8)
ggplot(prevdf1, aes(TotalAbundance, Prevalence / nsamples(sod.ps),color=Phylum)) +
  # Include a guess for parameter
  geom_hline(yintercept = 0.005, alpha = 0.5, linetype = 2) +  geom_point(size = 2, alpha = 0.7) +
  scale_x_log10() +  xlab("Total Abundance") + ylab("Prevalence [Frac. Samples]") +
  facet_wrap(~Phylum) + theme(legend.position="none")
dev.off()
```

### Let's make histograms of read counts before and after filtering
```{r Reads sequenced by experiment}
sdt = data.table::data.table(as(sample_data(sod.ps), "data.frame"),
                 TotalReads = sample_sums(sod.ps), keep.rownames = TRUE)
pseqDepth = ggplot(sdt, aes(TotalReads)) + geom_histogram(color="black", fill="lightblue") + ggtitle("Histogram of sequencing depth-before filtering samples with low read depths") + labs(x="Depth", y="Frequency")
pseqDepth

sdt_filt = data.table::data.table(as(sample_data(sod_5asv.ps), "data.frame"),
                 TotalReads = sample_sums(sod_5asv.ps), keep.rownames = TRUE)
pseqDepth_filt = ggplot(sdt_filt, aes(TotalReads)) + geom_histogram(color="black", fill="lightblue") + ggtitle("Histogram of sequencing depth after filtering samples with low read depths") + labs(x="Depth", y="Frequency")
pseqDepth_filt
```

### Let's make rarefaction curves 
```{r Rarefaction curves}
otu_data <- as(otu_table(sod.ps), "matrix")

S <- specnumber(otu_data) # observed number of species
raremax <- min(rowSums(otu_data))

pdf("Figures/16S/rarefaction_curve.pdf", width = 10, height = 8)
rarecurve(otu_data, step = 10, sample = raremax,  col = , cex = 0.6, label = FALSE, xlab = "Sequencing Depth", ylab = "Number of ASVs")
dev.off()

sample_data <- data.frame(sample_data(sod.ps))
otu_data <- as(otu_table(sod.ps), "matrix")

sample_data$Treatuntreat_EP <- factor(sample_data$Treatuntreat_EP, levels = c("Untreated", "Treated"))

raremax <- min(rowSums(otu_data))
rarecurve_data <- rarecurve(otu_data, step = 10, sample = raremax, label = FALSE)

pdf("Figures/16S/speciesaccum_curve.pdf", width = 10, height = 8)
asv_accum <- vegan::specaccum(otu_data, method = "exact")
plot(asv_accum, ci.type = "poly", col = "blue", lwd = 2, ci.lty = 0, ci.col = "lightblue", xlab = "Number of samples", ylab = "Number of Filtered ASVs")
dev.off()
```

### Software used
```{r, software used}
sessionInfo()
```
