---
title: "RPS10 Exploratory Analysis"
output: html_document
date: "2024-06-11"
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(root.dir = "~/SOD_DADS_microbiome")

library(phyloseq); packageVersion("phyloseq")
library(ggplot2); packageVersion("ggplot2")
library(readr); packageVersion("readr")
library(purrr); packageVersion("purrr")
library(furrr); packageVersion("furrr")
library(dplyr); packageVersion("dplyr")
library(stringr); packageVersion("stringr")
library(metacoder); packageVersion("metacoder")
library(data.table); packageVersion("data.table")
library(decontam); packageVersion("decontam")

```

Load data in order to make phyloseq and taxmap objects
```{r load files, include=TRUE}
#Load DADA2 files
seqtab.nochim <- read.table("Data/RPS10/dada2_outputs/rps10_seqtab_nochim.out",sep="\t",stringsAsFactors=F,header=T)

taxa <- read.table("Data/RPS10/dada2_outputs/rps10_taxa.out",sep="\t",stringsAsFactors=F,header=T)

taxa_filt <- taxa %>%
  mutate(across(starts_with("tax."), ~ ifelse(get(str_replace(cur_column(), "tax.", "boot.")) < 60, NA, .))) %>%
  set_names(str_replace(names(.), "tax.", ""))

samdf <- read.csv("Data/RPS10/Metadata_rps10_short_mod.csv")
```

Exploratory analysis-track reads throughout DADA2 analysis
Examine number of reads throughout the pipeline
```{r track reads through DADA2 analysis, include=TRUE}
track <- read.table("Data/RPS10/dada2_outputs/rps10_trackreads.out", sep="\t", stringsAsFactors = F, header=T)
```

```{r format reads, include=TRUE}
# For ease of analysis, convert the DADA2 outputs into a Phyloseq object, and filter out any necessary samples
all(rownames(seqtab.nochim) %in% samdf$Sample_nameRps10)
rownames(samdf) <- samdf$Sample_nameRps10

samdata.ps <- sample_data(samdf)

taxa.ps <- tax_table(as.matrix(taxa_filt))

seqtab.nochim.ps <- otu_table(seqtab.nochim, taxa_are_rows=FALSE)

ps <- phyloseq(seqtab.nochim.ps, samdata.ps, taxa.ps)

#Add refseq slot
dna <- Biostrings::DNAStringSet(taxa_names(ps))
names(dna) <- taxa_names(ps)
ps <- merge_phyloseq(ps, dna)
taxa_names(ps) <- paste0("ASV", seq(ntaxa(ps)))
```


Decontamination step using decontam package-this was most necessary for RPS10 analysis.
```{r decontam, include=TRUE}
#Inspect library sizes
df <- as.data.frame(sample_data(ps)) # Put sample_data into a ggplot-friendly data.frame
df$LibrarySize <- sample_sums(ps)
df <- df[order(df$LibrarySize),]
df$Index <- seq(nrow(df))
ggplot(data=df, aes(x=Index, y=LibrarySize, color=Sample_or_Control)) + geom_point()


sample_data(ps)$is.neg <- sample_data(ps)$Sample_or_Control == "Control"
contamdf.prev <- isContaminant(ps, method="prevalence", neg="is.neg")
table(contamdf.prev$contaminant)
head(which(contamdf.prev$contaminant))

contamdf.prev05 <- isContaminant(ps, method="prevalence", neg="is.neg", threshold=0.5)
table(contamdf.prev05$contaminant)
which(contamdf.prev05$contaminant)

# There don't appear to be any contaminants so we won't proceed with additional filtering described in decontam examples. 

#Let's save 
ps.pa <- transform_sample_counts(ps, function(abund) 1*(abund>0))
ps.pa.neg <- prune_samples(sample_data(ps.pa)$Sample_or_Control == "control", ps.pa)
ps.pa.pos <- prune_samples(sample_data(ps.pa)$Sample_or_Control == "sample", ps.pa)

# Make data.frame of prevalence in positive and negative samples
df.pa <- data.frame(pa.pos=taxa_sums(ps.pa.pos), pa.neg=taxa_sums(ps.pa.neg),
                     contaminant=contamdf.prev05$contaminant)
ggplot(data=df.pa, aes(x=pa.neg, y=pa.pos, color=contaminant)) + geom_point() +
  xlab("Prevalence (Negative Controls)") + ylab("Prevalence (True Samples)")

# Remove contaminants
ps.decontam <- prune_taxa(!contamdf.prev05$contaminant, ps)
save(ps.decontam, file = "Data/RPS10/RPS10_phyloseq.decontam.RData")
```


Take a first glance at the data. Helpful for deciding how to filter data. Prevelance is how many samples have the ASV, and total abundance is how many times ASV is present across all samples. I don't do any filtering until I split up samples into relevant experiment.
```{r examine ASV Prevalance, include=TRUE}
#Prevalence-number of samples in which a taxon appears at least once
#Total abundance
prevdf = apply(X = phyloseq::otu_table(ps.decontam),
               MARGIN = ifelse(taxa_are_rows(ps.decontam), yes = 1, no = 2),
               FUN = function(x){sum(x > 0)})

# Add taxonomy and total read counts to this data.frame
prevdf = data.frame(Prevalence = prevdf,
                    TotalAbundance = taxa_sums(ps.decontam),
                    tax_table(ps.decontam))
```

Now I will focus on 100 plot experiment and prepare dataset for further analysis by subsetting these samples from the rest of the sample data. I will also remove allow abundance ASVs not present in at least one sample and not present at least five times
```{r subset reads, include=TRUE}
#Experiment 1-100 plots base of tree
plot100exp.ps<-phyloseq::subset_samples(ps.decontam, Hundredplot_tree_EP=="1")
#Prune ASVs not present in at least one sample
plot100exp.ps<- phyloseq::prune_taxa(taxa_sums(plot100exp.ps) > 0, plot100exp.ps)
save(plot100exp.ps, file="Data/RPS10/rps10_ee5/plot100exp_2.RData")

#Prune ASVs not present in at least 5 times
exp1_5_filter.ps<- phyloseq::prune_taxa(taxa_sums(plot100exp.ps) > 5, plot100exp.ps)
save(exp1_5_filter.ps, file="Data/RPS10/rps10_ee5/exp1_5_filter_2.RData")

#Let's now collect stats for 100 plot experiment
samples_to_subset <- samdf %>%
  filter(Hundredplot_tree_EP == 1) %>% 
  pull(Sample_nameRps10)
track$Sample <- rownames(track)
# Filter the track dataframe

track <- track %>%
  select(Sample, everything())

subset_track_reads <- track[track$Sample %in% samples_to_subset, ]
write.csv(subset_track_reads, file="Data/RPS10/rps10_ee5/track_readsRPS10_100plot_2.csv",row.names = FALSE)
# Print summary of the filtered track dataframe
summary(subset_track_reads)
```

```{r check prevalence info within different datasets}
#Double check that the filtering makes sense by making new prevalence dataframe
prevdf_check = apply(X = phyloseq::otu_table(exp1_5_filter.ps),
               MARGIN = ifelse(taxa_are_rows(exp1_5_filter.ps), yes = 1, no = 2),
               FUN = function(x){sum(x > 0)})

# Add taxonomy and total read counts to this data.frame
prevdf_check = data.frame(Prevalence = prevdf_check,
                    TotalAbundance = taxa_sums(exp1_5_filter.ps),
                    tax_table(exp1_5_filter.ps))
```

```{r Reads sequenced by experiment}
#Before filtering reads
sdt = data.table::data.table(as(sample_data(plot100exp.ps), "data.frame"),
                 TotalReads = sample_sums(plot100exp.ps), keep.rownames = TRUE)
psDepth = ggplot(sdt, aes(TotalReads)) + geom_histogram(color="black", fill="lightblue") + ggtitle("Histogram of sequencing depth-before filtering samples with low read depths") + labs(x="Depth", y="Frequency")
psDepth

#After filtering reads
sdt_filt = data.table::data.table(as(sample_data(exp1_5_filter.ps), "data.frame"),
                 TotalReads = sample_sums(exp1_5_filter.ps), keep.rownames = TRUE)
psDepth_filt = ggplot(sdt_filt, aes(TotalReads)) + geom_histogram(color="black", fill="lightblue") + ggtitle("Histogram of sequencing depth after filtering samples with low read depths") + labs(x="Depth", y="Frequency")
psDepth_filt
```

```{r Prevalence df after filtering}
#Total abundance
prevdf_afterfilt = apply(X = otu_table(exp1_5_filter.ps),
               MARGIN = ifelse(taxa_are_rows(exp1_5_filter.ps), yes = 1, no = 2),
               FUN = function(x){sum(x > 0)})

# Add taxonomy and total read counts to this data.frame
prevdf_afterfilt = data.frame(Prevalence = prevdf_afterfilt,
                    TotalAbundance = taxa_sums(exp1_5_filter.ps),
                    tax_table(exp1_5_filter.ps))

prevdf_afterfilt
```

```{r prevalence plots organized by genus, include=TRUE}
prevdf1 = subset(prevdf_afterfilt, Genus %in% get_taxa_unique(exp1_5_filter.ps, "Genus"))
ggplot(prevdf1, aes(TotalAbundance, Prevalence / nsamples(exp1_5_filter.ps),color=Genus)) +
  # Include a guess for parameter
  geom_hline(yintercept = 0.005, alpha = 0.5, linetype = 2) +  geom_point(size = 2, alpha = 0.7) +
  scale_x_log10() +  xlab("Total Abundance") + ylab("Prevalence [Frac. Samples]") +
  facet_wrap(~Genus) + theme(legend.position="none")
```

Bar plot facet wrapped by treatment
```{r Barplot not agglomerated}
#The function "transform_sample_counts()" will perform a given function across all taxa counts
#We will use it to convert the counts to percentages
ps_transform <- transform_sample_counts(exp1_5_filter.ps,  function(x) x / sum(x))

pdf("Figures/RPS10/RelAbundance_Plot_Treatment_2.pdf", width = 22, height = 20)
barplot1<-plot_bar(ps_transform, fill="Genus", title = "Proportional Abundance Plot-ASVs Colored-Coded Family level and Faceted by Treatment") + facet_wrap(~Treatuntreat_EP, scales="free_x", nrow=1)
barplot1
dev.off()

barplot1
```

```{r barplot-Relative abundance of taxa vs. sample-agglomerated at genus level}
dataRPS10 <- exp1_5_filter.ps %>%
  tax_glom(taxrank = "Species") %>%                     # agglomerate at Genus level
  transform_sample_counts(function(x) {x/sum(x)} ) %>% # Transform to rel. abundance
  psmelt() %>%                                         # Melt to long format
  filter(Abundance > 0.005) %>%                       # Filter out low abundance taxa
  arrange(Species)                                      # Sort data frame alphabetically by Family

# Plot 
abund_plot <- ggplot(dataRPS10, aes(x = Sample, y = Abundance, fill = Species)) + 
  geom_bar(stat = "identity", position = "stack", color = "black", size = 0.2) +
  scale_fill_viridis_d() +
  scale_color_viridis_d() +
  theme_minimal() +
  labs(
    y = "Relative Abundance",
    title = "Relative abundance of taxa by sample",
    fill = "tax.Species"
  ) +
  facet_grid(~factor(Treatuntreat_EP, labels = c("Untreated", "Treated")), scales = "free_x") +
  theme(
    axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5, size = 14),
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    legend.position = "bottom",
    legend.text = element_text(size = 20),
    legend.title = element_text(size = 20),  # Adjust legend title size
    legend.key.size = unit(2, "lines"),  # Adjust the size of the legend color key
    strip.text = element_text(size = 20),
    strip.background = element_blank()
  ) +
  guides(
    fill = guide_legend(
      reverse = FALSE,
      keywidth = 1,
      keyheight = 1,
      title.position = "top",
      title.hjust = 0.5,  # Center the legend title
      label.theme = element_text(size = 18)  # Adjust the size of the legend labels
    )
  )

ggsave("Figures/RPS10/RelAbundance_Plot_Treatment_Species_2.pdf", abund_plot, width = 22, height = 20)
```

We are also curious about top taxa 
```{r Top taxa}
plot100exp.ps@sam_data$Treatuntreat_EP <- factor(plot100exp.ps@sam_data$Treatuntreat_EP, labels = c("untreated", "treated"))
new_names <- sapply(strsplit(plot100exp.ps@sam_data$Sample_nameRps10, "-"), `[`, 1)

# Rename the Sample_nameRps10 column so sample names are shorter
plot100exp.ps@sam_data$Sample_nameRps10 <- new_names
top20 <- names(sort(taxa_sums(plot100exp.ps), decreasing=TRUE))[1:20]
ps.top20 <- transform_sample_counts(plot100exp.ps, function(OTU) OTU/sum(OTU))
ps.top20<- tax_glom(ps.top20, taxrank = "Species") 
ps.top20 <- prune_taxa(top20, ps.top20)

pdf("Figures/ITS/ee5_top20_species_boot60.pdf", width = 34, height = 20)
top20_plot<-plot_bar(ps.top20, x="Sample_nameRps10", fill="Species") + facet_wrap(~Treatuntreat_EP, scales="free_x")
dev.off()
top20_plot

plot_bar(ps.top20, x="Sample_nameRps10", fill="Species") + facet_wrap(~Treatuntreat_EP, scales="free_x")

data <-
  plot100exp.ps %>%
  tax_glom("Species") %>%
  psmelt() %>%
  as_tibble()

data1 <-
  plot100exp.ps %>%
  tax_glom("Species") %>%
  transform_sample_counts(function(x) {x/sum(x)} ) %>%
  psmelt() %>%
  as_tibble()

data_summed1 <- data1 %>%
  filter(Abundance > 0) %>%
  group_by(Treatuntreat_EP, Kingdom, Phylum, Class, Order, Family, Genus, Species) %>%
  summarise(Abundance = sum(Abundance),
            Prevalence = n_distinct(Sample)) 

write.csv(data_summed1, file="Results/RPS10/mostabundanttaxa_relativeabundance_untreatedtreated_boot60.csv",row.names = TRUE)

data2 <-
  plot100exp.ps %>%
  tax_glom("Species") %>%
  psmelt() %>%
  as_tibble()

# Calculate percentages
data_summed2 <- data2 %>%
  filter(Abundance > 0) %>%
  group_by(Treatuntreat_EP, Kingdom, Phylum, Class, Order, Family, Genus, Species) %>%
  summarise(
    Abundance = sum(Abundance),
    Prevalence = n_distinct(Sample)
  ) %>%
  group_by(Treatuntreat_EP) %>%
  mutate(
    Total_Abundance = sum(Abundance),
    Percent_Abundance = (Abundance / Total_Abundance) * 100
  )

write.csv(data_summed2, file = "Results/RPS10/mostabundanttaxa_untreatedtreated_boot60.csv", row.names = TRUE)

data3 <-
  plot100exp.ps %>%
  tax_glom("Species") %>%
  psmelt() %>%
  as_tibble()

# Calculate percentages
data_summed3 <- data3 %>%
  filter(Abundance > 0) %>%
  group_by(Kingdom, Phylum, Class, Order, Family, Genus, Species) %>%
  summarise(
    Abundance = sum(Abundance),
    Total_Abundance = sum(data3$Abundance),
    Prevalence = n_distinct(Sample),
    Percent_Abundance = (sum(Abundance) / sum(data3$Abundance)) * 100
  )
write.csv(data_summed3, file="Results/RPS10/mostabundanttaxa_boot60.csv",row.names = TRUE)
```

```{r Map exploration}
new_names <- sapply(strsplit(plot100exp.ps@sam_data$Sample_nameRps10, "-"), `[`, 1)

# Rename the Sample_nameRps10 column so sample names are shorter
plot100exp.ps@sam_data$Sample_nameRps10 <- new_names

plot100exp.ps@sam_data$Treatuntreat_EP <- factor(plot100exp.ps@sam_data$Treatuntreat_EP, labels = c("untreated", "treated"))

pdf("Figures/RPS10/location_of_plots.pdf", width = 16, height = 16)

map_plot <- ggplot(plot100exp.ps@sam_data, aes(x = Longitude, y = Latitude, label = Sample_nameRps10, color = Treatuntreat_EP)) +
  geom_point(size = 5, alpha = 0.5) +  # Add transparency to points to visualize overlapping points
  geom_text(size = 4, position = position_jitter(width = 0.025, height = 0.025), check_overlap = TRUE) +  # Jitter labels
  labs(x = "Longitude", y = "Latitude", title = "Sampling Locations") +
  scale_color_manual(values = c("untreated" = "blue", "treated" = "red")) +  # Set colors for treated and untreated
  theme_minimal() +  # Remove gray background and gridlines
  theme(  # Adjust title and axis label size, and change axis marks color
    plot.title = element_text(size = 20), 
    axis.title = element_text(size = 16),
    axis.text = element_text(color = "black")
  )

print(map_plot)
dev.off()

print(map_plot)
```

Let's now visualize similar relative abundance plots as heat trees for alternative figures
First convert to taxmap object
```{r Convert Phyloseq object to taxmap object, include=FALSE}
obj<-parse_phyloseq(exp1_5_filter.ps)
```

```{r Heat tree steps}
sdt2 = data.table::data.table(as(sample_data(exp1_5_filter.ps), "data.frame"),
                 TotalReads = sample_sums(exp1_5_filter.ps), keep.rownames = TRUE)

lowest_count<-min(sdt2$TotalReads)

#$obj_dada$data$prop <- calc_obs_props(obj_dada, "otu_table", cols = #$obj$data$sample_data$sample_id)

#obj_dada$data$prop <- zero_low_counts(obj_dada, data = 'prop', min_count = 1 / #lowest_count, cols = obj$data$sample_data$sample_id)

obj$data$tax_abund <- calc_taxon_abund(obj, data = "otu_table", cols = obj$data$sample_data$sample_id)

obj$data$abund_prop <- calc_obs_props(obj, data = "otu_table", cols = obj$data$sample_data$sample_id, groups = rep("tax_prop", nrow(obj$data$sample_data)))

obj$data$tax_prop <- calc_taxon_abund(obj, data = "abund_prop", cols = "tax_prop")
obj$data$abund_prop <- NULL

# Calculate total abundance per sample
asv_total_abundance <- rowSums(obj$data$tax_abund[, -1])

obj$data$tax_data <- data.frame(lapply(obj$data$tax_data, function(x) ifelse(is.na(x), "Unknown", x)))
save(obj, file="Data/RPS10/obj.RData")
```

## Software used
```{r}
sessioninfo::session_info()
```
