---
title: "RPS10 diversity analysis"
output:
  html_document
date: "2023-12-01"
---
```{r setup, include=TRUE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(root.dir = "~/Documents/Postdoc/SOD_microbiome")

library(phyloseq); packageVersion("phyloseq")
library(ggplot2); packageVersion("ggplot2")
library(readr); packageVersion("readr")
library(purrr); packageVersion("purrr")
library(dplyr); packageVersion("dplyr")
library(metacoder); packageVersion("metacoder")
library(data.table); packageVersion("data.table")
library(agricolae);packageVersion("agricolae")
library(pander); packageVersion("pander")
library(kableExtra); packageVersion("kableExtra")
library(vegan);packageVersion("vegan")
library(heatmaply); packageVersion("heatmaply")
library(gridExtra);packageVersion("gridExtra")
```

We now need to normalize the data, as the read counts are not even across samples. This is especially important when doing beta diversity analyses. DESEQ accounts for unenven sampling, and for initial species richness and diversity (alpha diversity) analyses, it is recommended not to normalize the data first. 
```{r Normalization, include=TRUE}
#Rarefaction
#Figure out sample with lowest read depth and this is the number of reads to rarefy read counts to
load("Data/RPS10/rps10_pruned3k.RData")
load("Data/RPS10/rps10_pruned3k_5asvs.RData")

#Prune ASVs not present in least 5 times (across one more more samples)
#exp1_5_filter.ps<- phyloseq::prune_taxa(taxa_sums(exp1_prune3kreads.ps) > 5, exp1_prune3kreads.ps)
sdt = data.table::data.table(as(sample_data(exp1_5_filter.ps), "data.frame"),
                 TotalReads = sample_sums(exp1_5_filter.ps), keep.rownames = TRUE)
min_reads<-min(sdt$TotalReads)
rarefied.ps <- rarefy_even_depth(exp1_5_filter.ps, sample.size = min_reads, rngseed = 100)

#Relative abundance
strat_rel.ps<-transform_sample_counts(exp1_5_filter.ps, function(x) x / sum(x))
```

Alpha diversity
```{r alpha diversity analyses 1, include=TRUE}
exp1_prune3kreads.ps@sam_data$YearSam_EP <- as.factor(exp1_prune3kreads.ps@sam_data$YearSam_EP)
exp1_prune3kreads.ps@sam_data$Burned1.0_EP <- as.factor(exp1_prune3kreads.ps@sam_data$Burned1.0_EP)
exp1_prune3kreads.ps@sam_data$VegBin <- as.factor(exp1_prune3kreads.ps@sam_data$VegBin)
exp1_prune3kreads.ps@sam_data$SoilBin <- as.factor(exp1_prune3kreads.ps@sam_data$SoilBin)
exp1_prune3kreads.ps@sam_data$Treatuntreat_EP <- as.factor(exp1_prune3kreads.ps@sam_data$Treatuntreat_EP)
exp1_prune3kreads.ps@sam_data$YearTrt_EP <- as.factor(exp1_prune3kreads.ps@sam_data$YearTrt_EP)
exp1_prune3kreads.ps@sam_data$YrSinceTrt_EP <- as.factor(exp1_prune3kreads.ps@sam_data$YrSinceTrt_EP)
exp1_prune3kreads.ps@sam_data$YrSinceTrt_Range <- as.factor(exp1_prune3kreads.ps@sam_data$YrSinceTrt_Range)
exp1_prune3kreads.ps@sam_data$YrSinceTrt_BR <- as.factor(exp1_prune3kreads.ps@sam_data$YrSinceTrt_BR)
exp1_prune3kreads.ps@sam_data$YrSinceBurn_EP <- as.factor(exp1_prune3kreads.ps@sam_data$YrSinceBurn_EP)
exp1_prune3kreads.ps@sam_data$YrSinceTrt_BR <- as.factor(exp1_prune3kreads.ps@sam_data$YrSinceTrt_BR)
exp1_prune3kreads.ps@sam_data$Lineage_EP <- as.factor(exp1_prune3kreads.ps@sam_data$Lineage_EP)
exp1_prune3kreads.ps@sam_data$Fire_EP <- as.factor(exp1_prune3kreads.ps@sam_data$Fire_EP)

AD <- estimate_richness(exp1_prune3kreads.ps, measures = c("Observed", "InvSimpson", "Shannon"))
rownames(AD) <- gsub("\\.", "-", rownames(AD))

# Combine richness estimates with sample metadata
AD <- cbind(AD, sample_data(exp1_prune3kreads.ps)[row.names(AD), c("Treatuntreat_EP", "Lineage_EP", "YearSam_EP", "YearTrt_EP", "YrSinceTrt_EP", "YrSinceTrt_Range", "YrSinceTrt_BR", "Fire_EP", "Burned1.0_EP", "YrSinceBurn_EP", "VegBin", "SoilBin")])

# Convert to data.table and keep row names as a column
AD <- data.table(AD, keep.rownames = TRUE)

# Reshape the data frame
AD <- melt(AD, id.vars = c('rn', "Treatuntreat_EP", "Lineage_EP", "YearSam_EP", "YearTrt_EP", "YrSinceTrt_EP", "YrSinceTrt_Range", "YrSinceTrt_BR", "Fire_EP", "Burned1.0_EP", "YrSinceBurn_EP", "VegBin", "SoilBin"))

# Now conduct t-tests (assuming you have appropriate factors)
factors <- c("Treatuntreat_EP", "Burned1.0_EP")
# Iterate through each factor
ttest_list <- list()

# Iterate through each factor
for (factor_name in factors) {
  # Initialize an empty data.table to store t-test results for each alpha diversity measure
  ttest <- data.table(alpha_index = character(), pvalue = numeric())
  
  # Iterate through each alpha diversity measure
  for (alphad in c("InvSimpson", "Shannon", "Observed")) {
    # Perform t-test
    ttest_res <- t.test(value ~ get(factor_name), data = AD[variable == alphad], var.equal = TRUE)
    
    # Extract p-value from t-test results and store it in the data.table
    p_value <- ttest_res$p.value
    ttest <- rbind(ttest, data.table(alpha_index = alphad, pvalue = p_value))
  }
  
  # Add t-test results for the current factor to the list
  ttest_list[[factor_name]] <- ttest
}

# Print t-test results
for (factor_name in factors) {
  cat(paste("T-Test Results for Factor:", factor_name), "\n")
  print(knitr::kable(ttest_list[[factor_name]], format = "markdown"))
  cat("\n")
}
``` 

Let's try ANOVAs now
```{r alpha diversity analyses 2, include=TRUE}
# Initialize an empty list to store ANOVA results
anova_list <- list()

factors2 <- c("Treatuntreat_EP", "Lineage_EP", "YearSam_EP", "YearTrt_EP", "YrSinceTrt_EP", "YrSinceTrt_Range", "YrSinceTrt_BR", "Fire_EP", "Burned1.0_EP", "YrSinceBurn_EP", "VegBin", "SoilBin")

# Initialize a vector to store factor names for which a plot has been created
plotted_factors <- character()
 
for (factor_name in factors2) {
  # Initialize an empty data.table to store ANOVA results for each alpha diversity measure
  anova <- data.table(alpha_index = character(), pvalue = numeric(), df_between = numeric(), df_within = numeric(), ss_between = numeric(), ss_within = numeric(), ms_between = numeric(), ms_within = numeric(), F_value = numeric())
  
  # Iterate through each alpha diversity measure
  for (alphad in c("InvSimpson", "Shannon", "Observed")) {
    # Perform ANOVA
    anova_res <- aov(value ~ get(factor_name), data = AD[variable == alphad])
    
    # Extract relevant statistics from ANOVA results
    anova_summary <- summary(anova_res)
    
    # Extract p-value from ANOVA results and store it in the data.table
    p_value <- anova_summary[[1]]$`Pr(>F)`[1]
    
    # Extract other statistics
    df_between <- anova_summary[[1]]$"Df"[1]
    df_within <- anova_summary[[1]]$"Df"[2]
    ss_between <- anova_summary[[1]]$"Sum Sq"[1]
    ss_within <- anova_summary[[1]]$"Sum Sq"[2]
    ms_between <- anova_summary[[1]]$"Mean Sq"[1]
    ms_within <- anova_summary[[1]]$"Mean Sq"[2]
    F_value <- anova_summary[[1]]$"F value"[1]
    
    # Store statistics in the data.table
    anova <- rbind(anova, data.table(alpha_index = alphad, pvalue = p_value, df_between = df_between, df_within = df_within, ss_between = ss_between, ss_within = ss_within, ms_between = ms_between, ms_within = ms_within, F_value = F_value))
    
    # Check if p-value < 0.05 and plot has not been created for this factor yet
    if (p_value < 0.05 && !(factor_name %in% plotted_factors)) {
      # Subset data for InvSimpson and the significant factor
      data_subset <- AD[variable == "InvSimpson", .(value, factor_value = get(factor_name), variable = "InvSimpson")]
      
      # Convert factor_value to a factor
      data_subset$factor_value <- factor(data_subset$factor_value)
      
      # Plot box plot
      set.seed(123)
      plot_AD <- ggplot(data = data_subset, aes(x = factor_value, y = value, fill = factor_value)) +
        geom_boxplot(width = 0.7, outlier.colour = 'white', position = position_dodge(width = 0.75)) +
        geom_jitter(size = 1, position = position_jitter(width = 0.1)) +
        ylab('Inverse Simpson Index') +
        theme_minimal() +
        facet_grid(~variable, scales = "free_x", switch = "y") +
        theme(
          strip.text.y = element_blank(),
          strip.text.x = element_blank(),
          axis.text.y = element_text(size = 8),
          axis.ticks.y = element_blank()
        ) +
        labs(title = paste("Boxplot for", factor_name))
      
      # Save plot
      print(plot_AD)
      ggsave(paste0("Figures/RPS10/", factor_name, "alphadiversity_invsimp_boxplot.pdf"), plot_AD, width = 10, height = 12)
      
      # Add factor name to the plotted factors vector
      plotted_factors <- c(plotted_factors, factor_name)
    }
  }
  
  # Add ANOVA results for the current factor to the list
  anova_list[[factor_name]] <- anova
}

# Print ANOVA results
for (factor_name in factors2) {
  cat(paste("ANOVA Results for Factor:", factor_name), "\n")
  print(knitr::kable(anova_list[[factor_name]], format = "markdown"))
  cat("\n")
}
```

```{r compiled alpha diversity table}
# Initialize an empty data.table to store combined ANOVA results for all factors
final_anova_table <- data.table(factor_name = character(), alpha_index = character(), pvalue = numeric(), df_between = numeric(), df_within = numeric(), ss_between = numeric(), ss_within = numeric(), ms_between = numeric(), ms_within = numeric(), F_value = numeric())

for (factor_name in factors2) {
  # Initialize an empty data.table to store ANOVA results for each alpha diversity measure
  anova <- data.table(factor_name = character(), alpha_index = character(), pvalue = numeric(), df_between = numeric(), df_within = numeric(), ss_between = numeric(), ss_within = numeric(), ms_between = numeric(), ms_within = numeric(), F_value = numeric())
  
  # Iterate through each alpha diversity measure
  for (alphad in c("Observed","InvSimpson", "Shannon")) {
    # Perform ANOVA
    anova_res <- aov(value ~ get(factor_name), data = AD[variable == alphad])
    
    # Extract relevant statistics from ANOVA results
    anova_summary <- summary(anova_res)
    
    # Extract p-value from ANOVA results
    p_value <- anova_summary[[1]]$`Pr(>F)`[1]
    
    # Extract other statistics
    df_between <- anova_summary[[1]]$"Df"[1]
    df_within <- anova_summary[[1]]$"Df"[2]
    ss_between <- anova_summary[[1]]$"Sum Sq"[1]
    ss_within <- anova_summary[[1]]$"Sum Sq"[2]
    ms_between <- anova_summary[[1]]$"Mean Sq"[1]
    ms_within <- anova_summary[[1]]$"Mean Sq"[2]
    F_value <- anova_summary[[1]]$"F value"[1]
    
    # Store statistics in the data.table
    anova <- rbind(anova, data.table(factor_name = factor_name, alpha_index = alphad, pvalue = p_value, df_between = df_between, df_within = df_within, ss_between = ss_between, ss_within = ss_within, ms_between = ms_between, ms_within = ms_within, F_value = F_value))
  }
  
  # Append ANOVA results for the current factor to the final table
  final_anova_table <- rbind(final_anova_table, anova)
}

# Print final ANOVA table
print(knitr::kable(final_anova_table, format = "markdown"))
write.csv(final_anova_table, "Results/RPS10/final_anova_table.csv", row.names = FALSE)
```

```{r alpha diversity faceted plots}
# Estimate alpha diversity
AD <- estimate_richness(exp1_prune3kreads.ps, measures = c("Observed", "InvSimpson", "Shannon"))
rownames(AD) <- gsub("\\.", "-", rownames(AD))


# Combine richness estimates with sample metadata
AD <- cbind(AD, sample_data(exp1_prune3kreads.ps)[rownames(AD), c("Lineage_EP", "Fire_EP")])


variables <- c("Lineage_EP", "Fire_EP")

exp1_prune3kreads.ps@sam_data$YearSam_EP <- as.factor(exp1_prune3kreads.ps@sam_data$YearSam_EP)
# Create an empty list to store plots
plots <- list()

# Loop through each variable and create the plot
for (variable in variables) {
  p <- plot_richness(exp1_prune3kreads.ps, x = variable, measures = c("Shannon", "InvSimpson", "Observed")) + 
    geom_boxplot()
  plots[[variable]] <- p
}

# Arrange plots in a grid
plots_arrange <- gridExtra::grid.arrange(
  plots[["Lineage_EP"]], plots[["Fire_EP"]],
  nrow = length(variables)
)

# Add annotations to the top of each plot
plots_arrange_with_text <- cowplot::ggdraw() +
  cowplot::draw_plot(plots_arrange) +
  cowplot::draw_plot("Lineage_EP", x = 0.28, y = 0.98) +
  cowplot::draw_plot("Fire_EP", x = 0.73, y = 0.98)

# Save the arranged plots
ggsave(plots_arrange_with_text, path = "Figures/RPS10", filename = "alpha_diversity_each_factor.pdf", 
       width = 12, height = 16)
```

```{r alpha diversity compilation with tukey summary}
# Define your variables of interest
variables <- c("Lineage_EP", "Fire_EP")
measures <- c("Observed", "InvSimpson", "Shannon")

# Create an empty list to store ANOVA and Tukey's results
tukey_results <- list()

# Loop through each variable and measure
for (variable in variables) {
  for (measure in measures) {
    # Subset the data for the current variable and measure
    data_subset <- AD[, c(variable, measure)]
    
    # Perform ANOVA
    formula <- as.formula(paste(measure, "~ ."))
    anova_res <- aov(formula, data = data_subset)
    
    # Perform Tukey's post-hoc test
    tukey_result <- HSD.test(anova_res, variable)
    
    # Store Tukey's results
    print(tukey_result)
    tukey_results[[paste(variable, measure, sep = "_")]] <- tukey_result
  }
}
```
Linear regression analysis
```{r linear regression, include=TRUE}
AD <- estimate_richness(exp1_prune3kreads.ps, measures = c("Observed", "InvSimpson", "Shannon"))
rownames(AD) <- gsub("\\.", "-", rownames(AD))

# Combine richness estimates with sample metadata
AD <- cbind(AD, sample_data(exp1_prune3kreads.ps)[rownames(AD), c("YrSinceTrt_EP", "Fire_EP")])

AD$YrSinceTrt_EP <- as.numeric(AD$YrSinceTrt_EP)

# Plot linear regression
year_breaks <- seq(min(AD$YrSinceTrt_EP), max(AD$YrSinceTrt_EP), by = 1)
year_labels <- as.character(year_breaks)

ggplot(AD, aes(x = YrSinceTrt_EP, y = InvSimpson)) +
  geom_point() +  # Add points
  geom_smooth(method = "lm", se = FALSE) +  # Add linear regression line without confidence interval
  labs(x = "Year Since Treatment", y = "Inverse Simpson Index") +
  scale_x_continuous(breaks = year_breaks, labels = year_labels)# Label axes
  theme_minimal()  # Minimal theme

lm_model <- lm(InvSimpson ~ YrSinceTrt_EP, data = AD)

# Calculate R-squared value
r_squared <- summary(lm_model)$r.squared

# Print R-squared value
print(paste("R-squared:", round(r_squared, 3)))

# Print summary of the linear regression model
print(summary(lm_model))

#what about interaction
year_breaks <- seq(min(AD$YrSinceTrt_EP), max(AD$YrSinceTrt_EP), by = 1)
year_labels <- as.character(year_breaks)

# Plot linear regression with interaction effect and custom x-axis labels
ggplot(AD, aes(x = YrSinceTrt_EP, y = InvSimpson, color = factor(Fire_EP))) +
  geom_point() +  # Add points
  geom_smooth(method = "lm", se = FALSE, aes(group = Fire_EP)) +  # Add separate regression lines for each Fire_EP level
  labs(x = "Number of Years Since Treatment", y = "Inverse Simpson Index") +  # Label axes
  theme_minimal() +  # Minimal theme
  scale_color_manual(values = c("blue", "green", "red"), name = "Fire EP") +  # Custom color scale
  scale_x_continuous(breaks = year_breaks, labels = year_labels)  # Custom x-axis labelse EP")  # Custom color scale

lm_model <- lm(InvSimpson ~ YrSinceTrt_EP*Fire_EP, data = AD)

# Calculate R-squared value
r_squared <- summary(lm_model)$r.squared

# Print R-squared value
print(paste("R-squared:", round(r_squared, 3)))

# Print summary of the linear regression model
print(summary(lm_model))
```

Beta diversity
PERMANOVA-rarefied read counts
```{r PERMANOVA ANALYSES-rarefied,  include=TRUE}
#I used the filtered dataset (removed low abundance counts less than 2)
rarefied.ps@sam_data$YearSam_EP <- as.factor(rarefied.ps@sam_data$YearSam_EP)
rarefied.ps@sam_data$Burned1.0_EP <- as.factor(rarefied.ps@sam_data$Burned1.0_EP)
rarefied.ps@sam_data$VegBin <- as.factor(rarefied.ps@sam_data$VegBin)
rarefied.ps@sam_data$SoilBin <- as.factor(rarefied.ps@sam_data$SoilBin)
rarefied.ps@sam_data$Treatuntreat_EP <- as.factor(rarefied.ps@sam_data$Treatuntreat_EP)
rarefied.ps@sam_data$YearTrt_EP <- as.factor(rarefied.ps@sam_data$YearTrt_EP)
rarefied.ps@sam_data$YrSinceTrt_EP <- as.factor(rarefied.ps@sam_data$YrSinceTrt_EP)
rarefied.ps@sam_data$YrSinceTrt_Range <- as.factor(rarefied.ps@sam_data$YrSinceTrt_Range)
rarefied.ps@sam_data$YrSinceTrt_BR <- as.factor(rarefied.ps@sam_data$YrSinceTrt_BR)
rarefied.ps@sam_data$YrSinceBurn_EP <- as.factor(rarefied.ps@sam_data$YrSinceBurn_EP)
rarefied.ps@sam_data$YrSinceTrt_BR <- as.factor(rarefied.ps@sam_data$YrSinceTrt_BR)
rarefied.ps@sam_data$Lineage_EP <- as.factor(rarefied.ps@sam_data$Lineage_EP)
rarefied.ps@sam_data$Fire_EP <- as.factor(rarefied.ps@sam_data$Fire_EP)

mapping <- as.data.frame(sample_data(rarefied.ps))

mapping2=mapping[, c( "Treatuntreat_EP", "Lineage_EP", "YearSam_EP", "YearTrt_EP", "YrSinceTrt_EP", "YrSinceTrt_Range", "YrSinceTrt_BR", "Fire_EP", "Burned1.0_EP", "YrSinceBurn_EP", "VegBin", "SoilBin")]

# use an sapply loop as it's much more efficient than a for loop! sapply is a function that goes through each element of the first parameter (in our case colnames(mapping)), and passes each one to a custom function.
# You can declare the custom function in the sapply function RPS10elf, using the following syntax:

ASV_tables_bray <- phyloseq::distance(rarefied.ps, method = "bray")

df_metadata <- data.frame(sample_data(rarefied.ps))

df_metadata<-df_metadata[, c( "Treatuntreat_EP", "Lineage_EP", "YearSam_EP", "YearTrt_EP", "YrSinceTrt_EP", "YrSinceTrt_Range", "YrSinceTrt_BR", "Fire_EP", "Burned1.0_EP", "YrSinceBurn_EP", "VegBin", "SoilBin")]

pvals <- sapply(colnames(mapping2), function(variable){
  
  #Create formula for adonis() 
  form1<-as.formula(paste("ASV_tables_bray", variable, sep="~"))
  
  #run adonis with formula (ASV distance table ~ variable of interest) with sample information, for 9999 permutaions
  res <- adonis2(form1, data = df_metadata, permutations = 9999)
  
  # extract pvalue from adonis output
  pval <- res$`Pr(>F)`[[1]] # we take the first p value which corresponds to the variable of interest
  return(pval)
})

#Adjust p_values with fdr correction
pval_adj <- p.adjust(pvals, method = "fdr")

#Find significant variables with lower than 0.05 for their p values
save.pval.adonis0.05_braycurtis <- pval_adj[pval_adj < 0.05]
save.pval.adonis0.05_braycurtis
print(pval_adj)
```

Let's now examine beta diversity through ordination plots.
PERMANOVA analyses-ASV counts transformed to rel. abundance
```{r PERMANOVA ANALYSES-relative abundance, include=TRUE}
strat_rel.ps@sam_data$YearSam_EP <- as.factor(strat_rel.ps@sam_data$YearSam_EP)
strat_rel.ps@sam_data$Burned1.0_EP <- as.factor(strat_rel.ps@sam_data$Burned1.0_EP)
strat_rel.ps@sam_data$VegBin <- as.factor(strat_rel.ps@sam_data$VegBin)
strat_rel.ps@sam_data$SoilBin <- as.factor(strat_rel.ps@sam_data$SoilBin)
strat_rel.ps@sam_data$Treatuntreat_EP <- as.factor(strat_rel.ps@sam_data$Treatuntreat_EP)
strat_rel.ps@sam_data$YearTrt_EP <- as.factor(strat_rel.ps@sam_data$YearTrt_EP)
strat_rel.ps@sam_data$YrSinceTrt_EP <- as.factor(strat_rel.ps@sam_data$YrSinceTrt_EP)
strat_rel.ps@sam_data$YrSinceTrt_Range <- as.factor(strat_rel.ps@sam_data$YrSinceTrt_Range)
strat_rel.ps@sam_data$YrSinceTrt_BR <- as.factor(strat_rel.ps@sam_data$YrSinceTrt_BR)
strat_rel.ps@sam_data$YrSinceBurn_EP <- as.factor(strat_rel.ps@sam_data$YrSinceBurn_EP)
strat_rel.ps@sam_data$YrSinceTrt_BR <- as.factor(strat_rel.ps@sam_data$YrSinceTrt_BR)
strat_rel.ps@sam_data$Lineage_EP <- as.factor(strat_rel.ps@sam_data$Lineage_EP)
strat_rel.ps@sam_data$Fire_EP <- as.factor(strat_rel.ps@sam_data$Fire_EP)

#I used the filtered dataset (removed low abundance counts less than 2)
mapping <- as.data.frame(sample_data(strat_rel.ps))

mapping2=mapping[, c( "Treatuntreat_EP", "Lineage_EP", "YearSam_EP", "YearTrt_EP", "YrSinceTrt_EP", "YrSinceTrt_Range", "YrSinceTrt_BR", "Fire_EP", "Burned1.0_EP", "YrSinceBurn_EP", "VegBin", "SoilBin")]

# use an sapply loop as it's much more efficient than a for loop! sapply is a function that goes through each element of the first parameter (in our case colnames(mapping)), and passes each one to a custom function.
# You can declare the custom function in the sapply function RPS10elf, using the following syntax:

ASV_tables_bray <- phyloseq::distance(strat_rel.ps, method = "bray")

df_metadata <- data.frame(sample_data(strat_rel.ps))

df_metadata<-df_metadata[, c( "Treatuntreat_EP", "Lineage_EP", "YearSam_EP", "YearTrt_EP", "YrSinceTrt_EP", "YrSinceTrt_Range", "YrSinceTrt_BR", "Fire_EP", "Burned1.0_EP", "YrSinceBurn_EP", "VegBin", "SoilBin")]

pvals <- sapply(colnames(mapping2), function(variable){
  
  #Create formula for adonis() 
  form1<-as.formula(paste("ASV_tables_bray", variable, sep="~"))
  
  #run adonis with formula (ASV distance table ~ variable of interest) with sample information, for 9999 permutaions
  res <- adonis2(form1, data = df_metadata, permutations = 9999)
  
  # extract pvalue from adonis output
  pval <- res$`Pr(>F)`[[1]] # we take the first p value which corresponds to the variable of interest
  return(pval)
})

#Adjust p_values with fdr correction
pval_adj <- p.adjust(pvals, method = "fdr")

#Find significant variables with lower than 0.05 for their p values
save.pval.adonis0.05_braycurtis <- pval_adj[pval_adj < 0.05]
save.pval.adonis0.05_braycurtis

print(pval_adj)
```

```{r compile the stats}
ASV_tables_bray <- phyloseq::distance(strat_rel.ps, method = "bray")
df_metadata <- data.frame(sample_data(strat_rel.ps))
df_metadata <- df_metadata[, c("Treatuntreat_EP", "Lineage_EP", "YearSam_EP", "YearTrt_EP", "YrSinceTrt_EP", "YrSinceTrt_Range", "YrSinceTrt_BR", "Fire_EP", "Burned1.0_EP", "YrSinceBurn_EP", "VegBin", "SoilBin")]

# Perform PERMANOVA and beta dispersion analysis, and extract statistics
permanova_results <- lapply(colnames(df_metadata), function(variable) {
  # PERMANOVA
  res_permanova <- adonis2(ASV_tables_bray ~ ., data = df_metadata[, variable, drop = FALSE], permutations = 9999)
  r_squared <- res_permanova$R2[1] # Extract R-squared value for PERMANOVA
  f_value <- res_permanova$F[1] # Extract F-value for PERMANOVA
  pval_permanova <- res_permanova$`Pr(>F)`[[1]] # Extract p-value for PERMANOVA
  
  # Beta dispersion analysis
  res_betadisper <- betadisper(d = ASV_tables_bray, group = df_metadata[, variable])
  perm_test_res <- permutest(res_betadisper)
  pval_betadisper <- perm_test_res$tab$`Pr(>F)`[[1]] # Extract p-value for beta dispersion
  
  # Combine results
  results <- data.frame(Factor = variable, F_Value = f_value, R_Squared = r_squared, P_Value_BetaDispersion = pval_betadisper, P_Value_PERMANOVA = pval_permanova)
  return(results)
})

# Combine results into a data frame
permanova_results_df <- do.call(rbind, permanova_results)

# Adjust p-values for PERMANOVA
permanova_results_df$Adjusted_P_Value_PERMANOVA_fdr <- p.adjust(permanova_results_df$P_Value_PERMANOVA, method = "fdr")

permanova_results_df$Adjusted_P_Value_PERMANOVA_bonferroni <- p.adjust(permanova_results_df$P_Value_PERMANOVA, method = "bonferroni")

# Save as CSV file
write.csv(permanova_results_df, file = "Results/RPS10/PERMANOVA_results.csv", row.names = FALSE)
```

What about interactions between factors?
```{r pairwise comparisons}
# Calculate PERMANOVA
ASV_tables_bray <- phyloseq::distance(strat_rel.ps, method = "bray")
df_metadata <- data.frame(sample_data(strat_rel.ps))
df_metadata <- df_metadata[, c( "Treatuntreat_EP", "Lineage_EP", "YearTrt_EP", "YearSam_EP", "YrSinceTrt_Range", "YrSinceTrt_EP","Fire_EP", "Burned1.0_EP", "YrSinceBurn_EP", "VegBin", "SoilBin")]

# Define function to perform PERMANOVA for pairwise combinations of factors
perform_permutation <- function(factor1, factor2) {
  interaction <- paste(factor1, factor2, sep = " * ")
  res <- adonis2(ASV_tables_bray ~ ., data = df_metadata[, c(factor1, factor2)], permutations = 9999)
  return(list(interaction = interaction, p_value = res$`Pr(>F)`[[1]]))
}

# Perform pairwise comparisons between factors
significant_interactions <- sapply(names(df_metadata), function(factor1) {
  sapply(names(df_metadata), function(factor2) {
    if (factor1 != factor2) {
      perform_permutation(factor1, factor2)
    } else {
      NULL
    }
  })
}, simplify = "array")

# Filter out NULLs and combine results
significant_interactions <- significant_interactions[!sapply(significant_interactions, is.null)]

# Print significant interactions
print(significant_interactions)

interactions_df <- do.call(rbind.data.frame, significant_interactions)
print(significant_interactions)

interactions_df <- do.call(rbind.data.frame, significant_interactions)


adjusted_p_values <- p.adjust(interactions_df$p_value, method = "fdr")
# Add adjusted p-values to the data frame
interactions_df$p_value <- adjusted_p_values


# Extract unique factors
unique_factors <- unique(unlist(strsplit(interactions_df$interaction, " \\* ")))

# Create an empty matrix to store p-values
interaction_matrix <- matrix(NA, nrow = length(unique_factors), ncol = length(unique_factors), dimnames = list(unique_factors, unique_factors))

# Fill the matrix with p-values
for (i in 1:nrow(interactions_df)) {
  factors <- unlist(strsplit(interactions_df$interaction[i], " \\* "))
  interaction_matrix[factors[1], factors[2]] <- interactions_df$p_value[i]
  interaction_matrix[factors[2], factors[1]] <- interactions_df$p_value[i]
}

rownames(interaction_matrix) <- unique_factors 
colnames(interaction_matrix) <- unique_factors 


# Convert the matrix to a data frame for ggplot
interaction_df <- as.data.frame(as.table(interaction_matrix))
interaction_df <- as.data.frame(as.table(interaction_matrix))
interaction_df$color <- ifelse(is.na(interaction_df$Freq), "gray",
                               ifelse(interaction_df$Freq <= 0.01, "darkblue",
                                      ifelse(interaction_df$Freq <= 0.05, "lightblue", "white")))

# Plot heatmap with continuous color scale
heatmap <- ggplot(data = interaction_df, aes(x = Var1, y = Var2, fill = color)) +
  geom_tile(color = "black") +  # Add border for each tile
  scale_fill_manual(values = c("gray", "darkblue", "white"),
                    name = "P-value",
                    labels = c("NA", "P < 0.01", "P > 0.05"),
                    na.value = "gray") +  # Set manual color scale and legend labels
  scale_fill_identity(name = NULL) +  # Use identity scale for custom colors
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 10),  # Adjust x-axis text angle and size
        axis.text.y = element_text(size = 10),  # Adjust y-axis text size
        panel.grid = element_blank(),  # Remove gridlines
        panel.border = element_blank(),  # Remove border around the plot
        axis.line = element_blank(),  # Remove axis lines
        legend.position = "bottom",  # Position legend at the bottom
        legend.direction = "horizontal",  # Horizontal legend
        legend.key.size = unit(0.5, "cm"),  # Adjust legend key size
        legend.title = element_blank(),  # Remove legend title
        legend.text = element_text(size = 8))  # Adjust legend text font size

# Assign row and column names
heatmap <- heatmap + labs(x = NULL, y = NULL)

# Print heatmap
print(heatmap)

# Save the heatmap as a PDF file
ggsave("Results/RPS10/ee7/permanova_interactions_heatmap.pdf", plot = heatmap, width = 8, height = 6)
```

```{r PERMANOVA interactions}
res <- adonis2(ASV_tables_bray~Treatuntreat_EP*Fire_EP*YrSinceTrt_Range, data = df_metadata, permutations = 9999)
res
pval=res$`Pr(>F)`[2]
pval_adj <- p.adjust(pval, method = "fdr")
print(pval_adj)

res2 <- adonis2(ASV_tables_bray~Treatuntreat_EP*YrSinceTrt_Range*Fire_EP*YearSam_EP, data = df_metadata, permutations = 9999)
res2
pval=res2$`Pr(>F)`[2]
pval_adj <- p.adjust(pval, method = "fdr")
print(pval_adj)

res3 <- adonis2(ASV_tables_bray~Fire_EP*YrSinceTrt_EP*Lineage_EP*YrSinceBurn_EP, data = df_metadata, permutations = 9999)
res3
pval=res3$`Pr(>F)`[2]
pval_adj <- p.adjust(pval, method = "fdr")
print(pval_adj)
```

Let's now examine beta diversity through ordination plots. We'll will examine outputs for rarefied data as well as data adjusted by relative abundance

Beta diversity-not normalized to account for uneven read depth 
```{r Beta diversity}
#Not normalized

ps_PCoA <- ordinate(exp1_5_filter.ps, 
                    method = "PCoA", 
                    distance = "bray")

plot_1 <- plot_ordination(exp1_5_filter.ps, 
                          ps_PCoA, 
                          color = "Treatuntreat_EP") +
  ggtitle("PCoA ordination plot - read counts not normalized") +
  labs(color = "Factor") +  # Rename legend title
  theme(axis.text = element_text(size = 14),
        axis.title = element_text(size = 14),
        title = element_text(size = 13),
        legend.text = element_text(size = 18))

plot_1
```

We'll now make ordination plots on rarefied data
```{r Beta diversity-Rarefy read counts}

ps_PCoA2 = ordinate(rarefied.ps, 
                    method = "PCoA", 
                    distance = "bray")

plot_2<- plot_ordination(rarefied.ps, 
                         ps_PCoA2, 
                         labs(color = "Factor"),
                         color= "Treatuntreat_EP") +
  ggtitle("PCoA ordination plot - rarefied counts") + theme(axis.text=element_text(size=14), axis.title=element_text(size=14), title =  element_text(size=13), legend.text = element_text(size = 18))

plot_2
```

We'll now make ordination plots on data for which we've calculated relative abundance
```{r Beta diversity-Rel. abundance treated and untreated samples}
#Relative abundance
ps_PCoA3 = ordinate(strat_rel.ps, 
                   method="PCoA", 
                   distance="bray")

plot_3<- plot_ordination(strat_rel.ps, 
                         ps_PCoA3,
                         color= "Treatuntreat_EP") +
  ggtitle("PCoA ordination plot - relative abundance") +
  labs(color = "Factor") + 
  theme(axis.text=element_text(size=14), axis.title=element_text(size=14), title =  element_text(size=13), legend.text = element_text(size = 18))

plot_ellipses<-plot_3 +
   stat_ellipse(type = "norm", linetype = 2) +
  theme_bw()

ggsave("Figures/RPS10/PCoA_ordination_treatuntreat.pdf", plot_ellipses, width = 10, height = 10)

plot_ellipses

#Export PCoA coordinates per sample
write.csv(ps_PCoA3[["points"]], file="Results/RPS10/RPS10_PCoA_coordinates_persample.csv",row.names = TRUE)
```

Now, we'll look at ordination plot (rel abundance dataset), with fire factor
```{r Beta diversity-Rel. abundance by fire factor}
#Relative abundance

ps_PCoA3 = ordinate(strat_rel.ps, 
                   method="PCoA",
                   distance="bray")

plot_4<- plot_ordination(strat_rel.ps, 
                         ps_PCoA3,
                         labs(color = "Factor"),
                         color= "Fire_EP") +
  ggtitle("PCoA ordination plot - relative abundance") +
  theme(axis.text=element_text(size=14), axis.title=element_text(size=14), title =  element_text(size=13), legend.text = element_text(size = 18))


plot_ellipses<-plot_4 +
   stat_ellipse(type = "norm", linetype = 2) +
  theme_bw()

plot_ellipses

res <- adonis2(ASV_tables_bray~Treatuntreat_EP*Fire_EP, data = df_metadata, permutations = 9999)
res
pval=res$`Pr(>F)`[2]
pval_adj <- p.adjust(pval, method = "fdr")
print(pval_adj)

```

Now, we'll look at ordination plot (rel abundance dataset), with year since trt range factor
```{r Beta diversity-Rel. abundance by YrSinceTrt_Range}
#Relative abundance
ps_PCoA3 = ordinate(strat_rel.ps, 
                   method="PCoA", 
                   distance="bray")

plot_5<- plot_ordination(strat_rel.ps, 
                         ps_PCoA3,
                         labs(color = "Factor"),
                         color= "YrSinceTrt_Range") +
  ggtitle("PCoA ordination plot - relative abundance") +
  theme(axis.text=element_text(size=14), axis.title=element_text(size=14), title =  element_text(size=13), legend.text = element_text(size = 18))
plot_5


plot_ellipses<-plot_5 +
   stat_ellipse(type = "norm", linetype = 2) +
  theme_bw()

plot_ellipses

res <- adonis2(ASV_tables_bray~Treatuntreat_EP*YrSinceTrt_Range, data = df_metadata, permutations = 9999)
res
pval=res$`Pr(>F)`[2]
pval_adj <- p.adjust(pval, method = "fdr")
print(pval_adj)

```

Now, we'll look at ordination plot (rel abundance dataset), with year since trt broad range factor
```{r Beta diversity-Rel. abundance by yr since trt factor BR}
#Relative abundance

strat_rel.ps@sam_data$YrSinceTrt_BR <- factor(strat_rel.ps@sam_data$YrSinceTrt_BR, levels = c("untreated", "1to2", "3to4", "5to8", "10to12"))


ps_PCoA3 = ordinate(strat_rel.ps, 
                   method="PCoA", 
                   distance="bray")

plot_6<- plot_ordination(strat_rel.ps, 
                         ps_PCoA3,
                         labs(color = "Factor"),
                         color= "YrSinceTrt_BR") +
  ggtitle("PCoA ordination plot - relative abundance") +
  theme(axis.text=element_text(size=14), axis.title=element_text(size=14), title =  element_text(size=13), legend.text = element_text(size = 18))
plot_6


plot_ellipses<-plot_6 +
   stat_ellipse(type = "norm", linetype = 2) +
  theme_bw()

plot_ellipses

ggsave("Figures/RPS10/PCoA_ordination_yrsincetrtBR.pdf", plot_ellipses, width = 10, height = 10)
```

Let's simplify ordination plot figure, so we have a pdf figure with all ordination plots, if permanova p value was less than 0.05 after being adjusted-using less conservative fdr approach.

```{r}
# Define a custom labeller function that returns an empty string
custom_labeller <- function(variable, value) {
  return("")
}

# Create a list to store plots
plot_list <- list()

# Loop through each factor
for (factor in c("Fire_EP", "Lineage_EP", "YrSinceTrt_EP", "YrSinceBurn_EP")) {
  # Perform PCoA ordination
  ps_PCoA <- ordinate(strat_rel.ps, 
                      method = "PCoA", 
                      distance = "bray")
  
  # Plot ordination
  ordination_plot <- plot_ordination(strat_rel.ps, 
                                      ps_PCoA,
                                      color = factor) +
    labs(title = NULL,
         color = factor) +
    theme(axis.text = element_text(size = 14), 
          axis.title = element_text(size = 14), 
          title = element_text(size = 13), 
          legend.text = element_text(size = 18)) +
    stat_ellipse(type = "norm", linetype = 2) + # Add ellipses
    theme_bw() +
    facet_wrap(~ ., scales = "free", labeller = custom_labeller)  # Use custom labeller
  
  # Add the plot to the list
  plot_list[[factor]] <- ordination_plot
}

# Arrange plots side by side in rows of three
plots_arrange <- grid.arrange(
  plot_list$Lineage_EP, plot_list$YrSinceTrt_EP, plot_list$Fire_EP, plot_list$YrSinceBurn_EP,
  ncol = 3)

# Save the arranged plots
ggsave("Figures/RPS10/PCoA_ordination_facetted.pdf", plots_arrange, width = 15, height = 7)

```

## Software used
```{r}
sessioninfo::session_info()
```