---
title: "RPS10 core microbiome followup"
output: html_document
date: "2023-11-06"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(root.dir = "~/Documents/Postdoc/SOD_microbiome/")

library(phyloseq); packageVersion("phyloseq")
library(ggplot2); packageVersion("ggplot2")
library(readr); packageVersion("readr")
library(purrr); packageVersion("purrr")
library(furrr); packageVersion("furrr")
library(dplyr); packageVersion("dplyr")
library(stringr); packageVersion("stringr")
library(metacoder); packageVersion("metacoder")
library(data.table); packageVersion("data.table")
library(Biostrings); packageVersion("Biostrings")
library(phylosmith); packageVersion("phylosmith")
```

Load filtered phyloseq objects
```{r Make additional taxmap objects asvs not present 5 percent of time filtered}
load("Data/RPS10/rps10_ee7/rps10_pruned3k_5asvs.RData")
load("Data/RPS10/rps10_ee7/exp1_5percent_filter.RData")
load("Data/RPS10/rps10_ee7/exp1_10percent_filter.RData")
```

These subsequent analysis are an attempt to understand what my constitute the core microbiome of treated vs. untreated samples. This is in addition to analysis in the first script. 

Towards a core microbiome analysis

Let's look at common taxa in both treated and untreated samples and taxa that are unique to each
```{r common taxa}
#Samples filtered if less than 3k reads and ASVs not present at least 5 times
phylosmith::common_taxa(exp1_5_filter.ps, treatment = "treatuntreat_EP", subset = NULL, n = 'all')

#Samples filtered above, and any additional ASVs filtered if not present at least 5% of time
phylosmith::common_taxa(exp1_5percent_filter.ps, treatment = "treatuntreat_EP", subset = NULL, n = 'all')

#Samples filtered above, and any additional ASVs filtered if not present at least 10% of time
phylosmith::common_taxa(exp1_10percent_filter.ps, treatment = "treatuntreat_EP", subset = NULL, n = 'all')
```


Let's look at core taxa-it doesn't seem there are any core taxa across treatments with these particular frequency and abundance threshold values
```{r core taxa}
coretaxa.ps<-taxa_core(exp1_5_filter.ps, subset = NULL, frequency = 0.05, abundance_threshold = 0.01)

taxa_core_graph(exp1_5_filter.ps, treatment = NULL, subset = NULL,
abundance_thresholds = seq(0.01, 1, 0.01), frequencies = seq(0.05, 0.3, 0.01), colors = 'default', treatment_labels = "treatuntreat_EP", sample_labels = NULL, classification_labels= NULL)
```

```{r unique taxa}
#Let's identify taxa only present in treated vs. untreated samples
uniquetaxa<-phylosmith::unique_taxa(exp1_5_filter.ps, "treatuntreat_EP", subset = NULL)
data.frame(lapply(uniquetaxa, "length<-", max(lengths(uniquetaxa))))

#Let's identify taxa only present in treated vs. untreated samples in dataset where ASVs filtered if not at least 5% of samples
uniquetaxa_5percent<-phylosmith::unique_taxa(exp1_5percent_filter.ps, "treatuntreat_EP", subset = NULL)
data.frame(lapply(uniquetaxa_5percent, "length<-", max(lengths(uniquetaxa_5percent))))
```

```{r Abundance heatmaps, fig.width=22, fig.height=15}
#Let's create a tidier vector of sample names
samp_names<-exp1_5_filter.ps@sam_data$Sample_nameRps10
# Function to extract the first part before underscore
extract_first_part <- function(name) {
  parts <- unlist(strsplit(name, "[-_]"))
  return(parts[1])
}

# Applying the function to each name in the list
first_part_names <- sapply(samp_names, extract_first_part)

pdf("Figures/RPS10/phylosmith_abundance_heatmp.pdf", width = 40, height = 30)
#There aren't any taxa only in treated vs. untreated samples
abundance_heatmap(exp1_5_filter.ps, classification = "Species", "treatuntreat_EP", subset = NULL, treatment_labels=c("Untreated", "Treated"), sample_labels=first_part_names, transformation = 'log2', colors = 'default')
dev.off()

abundance_heatmap(exp1_5percent_filter.ps, classification = "Species", "treatuntreat_EP", subset = NULL, treatment_labels=c("Untreated", "Treated"), sample_labels=first_part_names, transformation = 'log2', colors = 'default')

abundance_heatmap(exp1_10percent_filter.ps, classification = "Species", "treatuntreat_EP", subset = NULL, treatment_labels=c("Untreated", "Treated"), sample_labels=first_part_names, transformation = 'log2', colors = 'default')
```


```{r}
sessioninfo::session_info()
```