---
title: "06_preposttrtexp"
output: html_document
date: "2024-05-21"
---
```{r setup, include=TRUE}

knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(root.dir = "~/Documents/Postdoc/SOD_microbiome")
library(phyloseq); packageVersion("phyloseq")
library(ggplot2); packageVersion("ggplot2")
library(readr); packageVersion("readr")
library(purrr); packageVersion("purrr")
library(furrr); packageVersion("furrr")
library(dplyr); packageVersion("dplyr")
library(stringr); packageVersion("stringr")
library(metacoder); packageVersion("metacoder")
library(data.table); packageVersion("data.table")
library(Biostrings); packageVersion("Biostrings")
library(vegan);packageVersion("vegan")
library(agricolae);packageVersion("agricolae")
library(gt);packageVersion("gt")
library(magrittr);packageVersion("magrittr")
library(data.table);packageVersion("data.table")
library(DESeq2);packageVersion("DESeq2")
library(cplm);packageVersion("cplm")
library(cowplot);packageVersion("cowplot")
theme_set(theme_cowplot())
library(ggbeeswarm);packageVersion("ggbeeswarm")
library(doParallel);packageVersion("doParallel")
library(foreach);packageVersion("foreach")
```

```{r subset data, include=TRUE}
load("Data/RPS10/rps10_pruned3k.RData")

pre_post.ps<-subset_samples(exp1_prune3kreads.ps, PrePost_sam_EP=="1")
pre_post.ps<- prune_taxa(taxa_sums(pre_post.ps) > 0, pre_post.ps)

readsums<-sort(sample_sums(pre_post.ps))
median(readsums)

#Let's also load data where any ASVs present less than 5 times are filtered
#load("Data/RPS10/rps10_pruned3k_5asvs.RData")
pre_post_5asv.ps<- prune_taxa(taxa_sums(pre_post.ps) > 5, pre_post.ps)
```

```{r Normalization experiment 1, include=TRUE}
#Could rarefy read counts, but will focus just on normalizing using the methods below. 

#Relative abundance
strat_rel_prepost.ps<-transform_sample_counts(pre_post_5asv.ps, function(x) x / sum(x))
```


Bar plot conglomerated by Genus-using filtered dataset and facet wrapped by treatment
```{r Barplot conglomerated by Genus}
#The function "transform_sample_counts()" will perform a given function across all taxa counts
#We will use it to convert the counts to percentages
ps_transform <- transform_sample_counts(pre_post_5asv.ps,  function(x) x / sum(x))
pdf("Figures/RPS10/relabundance_plot_treatment_prepost.pdf", width = 16, height = 16)
bar_plot<-plot_bar(ps_transform, fill="Genus", title = "Proportional Abundance Plot-ASVs Agglomerated at Family level and Faceted by Treatment") + facet_wrap(~Treatuntreat_EP, scales="free_x", nrow=1)
dev.off()

bar_plot
```

Another relative abundance barplot (must view PDF figure since it is a larger fig)
```{r barplot-Relative abundance of taxa vs. sample-not agglomerated}
rps10data <- pre_post_5asv.ps %>%
  tax_glom(taxrank = "Species") %>%                     # agglomerate at Genus level
  transform_sample_counts(function(x) {x/sum(x)} ) %>% # Transform to rel. abundance
  psmelt() %>%                                         # Melt to long format
  #filter(Abundance > 0.05) %>%                       # Filter out low abundance taxa
  arrange(desc(Abundance))
                                # Sort data frame alphabetically by phylum

# Plot 
abund_plot <- ggplot(rps10data, aes(x = Sample, y = Abundance, fill = Species)) + 
  geom_bar(stat = "identity", position = "stack", color = "black", size = 0.2) +
  scale_fill_viridis_d() +
  scale_color_viridis_d() +
  theme_minimal() +
  labs(
    y = "Relative Abundance",
    title = "Relative abundance of taxa by sample",
    fill = "Species"
  ) +
  facet_grid(~factor(Treatuntreat_EP, labels = c("untreated", "treated")), scales = "free_x") +
  theme(
    axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5, size = 14),
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    legend.position = "bottom",
    legend.text = element_text(size = 20),
    legend.title = element_text(size = 20),  # Adjust legend title size
    legend.key.size = unit(2, "lines"),  # Adjust the size of the legend color key
    strip.text = element_text(size = 20),
    strip.background = element_blank()
  ) +
  guides(
    fill = guide_legend(
      reverse = TRUE,
      keywidth = 1,
      keyheight = 1,
      title.position = "top",
      title.hjust = 3,  # Center the legend title
      label.theme = element_text(size = 18)  # Adjust the size of the legend labels
    )
  )

abund_plot

ggsave("Figures/RPS10/relabundance_plot_treatment_Species_prepost.pdf", abund_plot, width = 22, height = 16)
```

We are also curious about top taxa 
```{r Top taxa}
new_names <- sapply(strsplit(pre_post.ps@sam_data$Sample_nameRPS10, "-"), `[`, 1)

# Rename the Sample_nameRPS10 column
pre_post.ps@sam_data$Sample_nameRPS10 <- new_names

pre_post.ps@sam_data$Treatuntreat_EP <- factor(pre_post.ps@sam_data$Treatuntreat_EP, labels = c("untreated", "treated"))

top20 <- names(sort(taxa_sums(pre_post.ps), decreasing=TRUE))[1:20]
ps.top20 <- transform_sample_counts(pre_post.ps, function(OTU) OTU/sum(OTU))
ps.top20<- tax_glom(ps.top20, taxrank = "Species") 
ps.top20 <- prune_taxa(top20, ps.top20)

pdf("Figures/RPS10/ee5_top20_species_boot60_prepost.pdf", width = 34, height = 20)
plot_bar(ps.top20, x="Sample_nameRPS10", fill="Species") + facet_wrap(~Treatuntreat_EP, scales="free_x")
dev.off()

plot_bar(ps.top20, x="Sample_nameRPS10", fill="Species") + facet_wrap(~Treatuntreat_EP, scales="free_x")

data <-
  pre_post.ps %>%
  tax_glom("Species") %>%
  psmelt() %>%
  as_tibble()

data1 <-
  pre_post.ps %>%
  tax_glom("Species") %>%
  transform_sample_counts(function(x) {x/sum(x)} ) %>%
  psmelt() %>%
  as_tibble()

data_summed1 <- data1 %>%
  filter(Abundance > 0) %>%
  group_by(Treatuntreat_EP, Kingdom, Phylum, Class, Order, Family, Genus, Species) %>%
  summarise(Abundance = sum(Abundance),
            Prevalence = n_distinct(Sample)) 

write.csv(data_summed1, file="Results/RPS10/mostabundanttaxa_relativeabundance_untreatedtreated_boot60_prepost.csv",row.names = TRUE)

data2 <-
  pre_post.ps %>%
  tax_glom("Species") %>%
  psmelt() %>%
  as_tibble()

# Calculate percentages
data_summed2 <- data2 %>%
  filter(Abundance > 0) %>%
  group_by(Treatuntreat_EP, Kingdom, Phylum, Class, Order, Family, Genus, Species) %>%
  summarise(
    Abundance = sum(Abundance),
    Prevalence = n_distinct(Sample)
  ) %>%
  group_by(Treatuntreat_EP) %>%
  mutate(
    Total_Abundance = sum(Abundance),
    Percent_Abundance = (Abundance / Total_Abundance) * 100
  )

write.csv(data_summed2, file = "Results/RPS10/mostabundanttaxa_untreatedtreated_boot60_prepost.csv", row.names = TRUE)

data3 <-
  pre_post.ps %>%
  tax_glom("Species") %>%
  psmelt() %>%
  as_tibble()

# Calculate percentages
data_summed3 <- data3 %>%
  filter(Abundance > 0) %>%
  group_by(Kingdom, Phylum, Class, Order, Family, Genus, Species) %>%
  summarise(
    Abundance = sum(Abundance),
    Total_Abundance = sum(data3$Abundance),
    Prevalence = n_distinct(Sample),
    Percent_Abundance = (sum(Abundance) / sum(data3$Abundance)) * 100
  )
write.csv(data_summed3, file="Results/RPS10/mostabundanttaxa_boot60_prepost.csv",row.names = TRUE)
```


```{r Map exploration}
new_names <- sapply(strsplit(pre_post_5asv.ps@sam_data$Sample_nameRps10, "-"), `[`, 1)

# Rename the Sample_nameRps10 column so sample names are shorter
pre_post_5asv.ps@sam_data$Sample_nameRps10 <- new_names

pre_post_5asv.ps@sam_data$Treatuntreat_EP <- factor(pre_post_5asv.ps@sam_data$Treatuntreat_EP, labels = c("untreated", "treated"))

pdf("Figures/RPS10/location_of_plots_prepost.pdf", width = 16, height = 16)

map_plot <- ggplot(pre_post_5asv.ps@sam_data, aes(x = Longitude, y = Latitude, label = Sample_nameRps10, color = Treatuntreat_EP)) +
  geom_point(size = 5, alpha = 0.5) +  # Add transparency to points to visualize overlapping points
  geom_text(size = 4, position = position_jitter(width = 0.025, height = 0.025), check_overlap = TRUE) +  # Jitter labels
  labs(x = "Longitude", y = "Latitude", title = "Sampling Locations") +
  scale_color_manual(values = c("untreated" = "blue", "treated" = "red")) +  # Set colors for treated and untreated
  theme_minimal() +  # Remove gray background and gridlines
  theme(  # Adjust title and axis label size, and change axis marks color
    plot.title = element_text(size = 20), 
    axis.title = element_text(size = 16),
    axis.text = element_text(color = "black")
  )

print(map_plot)
dev.off()

print(map_plot)
```

Let's now visualize similar relative abundance plots as heat trees for alternative figures
First convert to taxmap object
```{r Convert Phyloseq object to taxmap object, include=FALSE}
obj<-parse_phyloseq(pre_post_5asv.ps)
```

```{r Heat tree steps}
sdt2 = data.table::data.table(as(sample_data(pre_post_5asv.ps), "data.frame"),
                 TotalReads = sample_sums(pre_post_5asv.ps), keep.rownames = TRUE)

lowest_count<-min(sdt2$TotalReads)

obj$data$tax_abund <- calc_taxon_abund(obj, data = "otu_table", cols = obj$data$sample_data$sample_id)

obj$data$abund_prop <- calc_obs_props(obj, data = "otu_table", cols = obj$data$sample_data$sample_id, groups = rep("tax_prop", nrow(obj$data$sample_data)))

obj$data$tax_prop <- calc_taxon_abund(obj, data = "abund_prop", cols = "tax_prop")
obj$data$abund_prop <- NULL

# Calculate total abundance per sample
asv_total_abundance <- rowSums(obj$data$tax_abund[, -1])

obj$data$tax_data <- data.frame(lapply(obj$data$tax_data, function(x) ifelse(is.na(x), "Unknown", x)))

save(obj, file="Data/RPS10/obj_prepost.RData")
```

```{r Heat tree for all samples}
set.seed(1)
heat_tree(obj,
          node_label = taxon_names,
          node_size = n_obs,
          node_color = n_obs,
          node_color_axis_label = "ASV count",
          node_size_axis_label = "Total Abundance of Taxa",
          layout = "da", initial_layout = "re",
          output_file = "Figures/RPS10/plot_heattree_rps10_prepost.pdf")
```

Now I will make more trees that better shows differences between treated and untreated samples
Let's first look at taxa present in treated samples
```{r Heat tree treated samples}
#help me make the heat tree in below code chunk but just treated samples
obj$data$tax_occ <- calc_n_samples(obj, "tax_abund", groups = obj$data$sample_data$Treatuntreat_EP, cols = obj$data$sample_data$sample_id)

set.seed(1) # This makes the plot appear the same each time it is run 
heat_tree(obj, 
          node_label = taxon_names,
          node_size = n_obs,
          node_color = obj$data$tax_occ$"treated", 
          node_size_axis_label = "ASV count",
          node_color_axis_label = "Treated samples with reads",
          layout = "davidson-harel",
          initial_layout = "reingold-tilford",
          output_file = "Figures/RPS10/plot_heattree_treated_plots_rps10_prepost.pdf")
```

Let's look at taxa present in untreated samples
```{r Heat tree untreated samples}
#help me make the heat tree in below code chunk but just treated samples

set.seed(1) # This makes the plot appear the same each time it is run 
heat_tree(obj, 
          node_label = taxon_names,
          node_size = n_obs,
          node_color = obj$data$tax_occ$"untreated", 
          node_size_axis_label = "ASV count",
          node_color_axis_label = "Untreated samples with reads",
          layout = "davidson-harel", # The primary layout algorithm
          initial_layout = "reingold-tilford",
          output_file = "Figures/RPS10/plot_heattree_untreated_plots_rps10_prepost.pdf")
```

Let's look at taxon abundance in treated samples
We first need to add more information to our taxmap object
```{r Heat tree untreated samples relative abundance}
obj$data$otu_props <- calc_obs_props(obj, "otu_table", cols = obj$data$sample_data$sample_id)

obj$data$tax_abund_prop <- calc_taxon_abund(obj, "otu_props")

obj$data$type_abund <- calc_group_mean(obj, "tax_abund_prop",
                                       cols =obj$data$sample_data$sample_id,
                                       groups = obj$data$sample_data$Treatuntreat_EP)
```

Let's look at taxon relative abundance in untreated samples
```{r Heat tree illustrating taxa in untreated samples}
set.seed(1)
obj %>%
  heat_tree(node_label = taxon_names,
            node_size = obj$data$type_abund$"untreated", 
            node_color = obj$data$type_abund$"untreated", 
            layout = "da", initial_layout = "re", 
            node_color_axis_label = "Untreated samples",
            title = "Taxa in untreated samples",
            output_file = "Figures/RPS10/plot_heattree_rel_abund_taxa_untreated_plots_rps10_prepost.pdf")
```

Let's look at taxon relative abundance in treated samples
```{r Heat tree illustrating taxa in treated samples}
set.seed(1)
obj %>%
  heat_tree(node_label = taxon_names,
            node_size = obj$data$type_abund$"treated", 
            node_color = obj$data$type_abund$"treated", 
            layout = "da", initial_layout = "re", 
            node_color_axis_label = "Treated samples",
            title = "Taxa in treated samples",
            output_file = "Figures/RPS10/plot_heattree_rel_abund_taxa_treated_plots_rps10_prepost.pdf")
```

```{r Normalization, include=TRUE}
#Prune ASVs not present in least 5 times (across one more more samples)
#pre_post_5asv.ps<- phyloseq::prune_taxa(taxa_sums(pre_post.ps) > 5, pre_post.ps)
sdt = data.table::data.table(as(sample_data(pre_post_5asv.ps), "data.frame"),
                 TotalReads = sample_sums(pre_post_5asv.ps), keep.rownames = TRUE)
min_reads<-min(sdt$TotalReads)
rarefied.ps <- rarefy_even_depth(pre_post_5asv.ps, sample.size = min_reads, rngseed = 100)

#Relative abundance
strat_rel_prepost.ps<-transform_sample_counts(pre_post_5asv.ps, function(x) x / sum(x))
```

Alpha diversity
```{r alpha diversity analyses 1, include=TRUE}
AD <- estimate_richness(pre_post.ps, measures = c("Observed", "InvSimpson", "Shannon"))
rownames(AD) <- gsub("\\.", "-", rownames(AD))

pre_post.ps@sam_data$YearSam_EP <- as.factor(pre_post.ps@sam_data$YearSam_EP)

pre_post.ps@sam_data$Burned1.0_EP <- as.factor(pre_post.ps@sam_data$Burned1.0_EP)

pre_post.ps@sam_data$VegBin <- as.factor(pre_post.ps@sam_data$VegBin)

pre_post.ps@sam_data$SoilBin <- as.factor(pre_post.ps@sam_data$SoilBin)

# Combine richness estimates with sample metadata
AD <- cbind(AD, sample_data(pre_post.ps)[row.names(AD), c("Treatuntreat_EP", "YearSam_EP", "Fire_EP", "Burned1.0_EP", "VegBin", "SoilBin")])

# Convert to data.table and keep row names as a column
AD <- data.table(AD, keep.rownames = TRUE)

# Reshape the data frame
AD <- melt(AD, id.vars = c('rn', c("Treatuntreat_EP", "YearSam_EP", "Fire_EP", "Burned1.0_EP", "VegBin", "SoilBin")))

# Now conduct t-tests (assuming you have appropriate factors)
factors <- c("Treatuntreat_EP", "Fire_EP", "Burned1.0_EP", "VegBin", "SoilBin")
# Iterate through each factor
ttest_list <- list()

# Iterate through each factor
for (factor_name in factors) {
  # Initialize an empty data.table to store t-test results for each alpha diversity measure
  ttest <- data.table(alpha_index = character(), pvalue = numeric())
  
  # Iterate through each alpha diversity measure
  for (alphad in c("InvSimpson", "Shannon", "Observed")) {
    # Perform t-test
    ttest_res <- t.test(value ~ get(factor_name), data = AD[variable == alphad], var.equal = TRUE)
    
    # Extract p-value from t-test results and store it in the data.table
    p_value <- ttest_res$p.value
    ttest <- rbind(ttest, data.table(alpha_index = alphad, pvalue = p_value))
  }
  
  # Add t-test results for the current factor to the list
  ttest_list[[factor_name]] <- ttest
}

# Print t-test results
for (factor_name in factors) {
  cat(paste("T-Test Results for Factor:", factor_name), "\n")
  print(knitr::kable(ttest_list[[factor_name]], format = "markdown"))
  cat("\n")
}
``` 

Let's try ANOVAs now
```{r alpha diversity analyses 2, include=TRUE}
# Initialize an empty list to store ANOVA results
anova_list <- list()

factors2 <- c("Treatuntreat_EP", "YearSam_EP", "Fire_EP", "Burned1.0_EP", "VegBin", "SoilBin")

# Initialize a vector to store factor names for which a plot has been created
plotted_factors <- character()
 
for (factor_name in factors2) {
  # Initialize an empty data.table to store ANOVA results for each alpha diversity measure
  anova <- data.table(alpha_index = character(), pvalue = numeric(), df_between = numeric(),
                      df_within = numeric(), ss_between = numeric(), ss_within = numeric(),
                      ms_between = numeric(), ms_within = numeric(), F_value = numeric())
  
  # Iterate through each alpha diversity measure
  for (alphad in c("InvSimpson", "Shannon", "Observed")) {
    # Perform ANOVA
    anova_res <- aov(value ~ get(factor_name), data = AD[variable == alphad])
    
    # Extract relevant statistics from ANOVA results
    anova_summary <- summary(anova_res)
    
    # Extract p-value from ANOVA results and store it in the data.table
    p_value <- anova_summary[[1]]$`Pr(>F)`[1]
    
    # Extract other statistics
    df_between <- anova_summary[[1]]$"Df"[1]
    df_within <- anova_summary[[1]]$"Df"[2]
    ss_between <- anova_summary[[1]]$"Sum Sq"[1]
    ss_within <- anova_summary[[1]]$"Sum Sq"[2]
    ms_between <- anova_summary[[1]]$"Mean Sq"[1]
    ms_within <- anova_summary[[1]]$"Mean Sq"[2]
    F_value <- anova_summary[[1]]$"F value"[1]
    
    # Store statistics in the data.table
    anova <- rbind(anova, data.table(alpha_index = alphad, pvalue = p_value, df_between = df_between, df_within = df_within, ss_between = ss_between, ss_within = ss_within, ms_between = ms_between, ms_within = ms_within, F_value = F_value))
    
    # Check if p-value < 0.05 and plot has not been created for this factor yet
    if (p_value < 0.05 && !(factor_name %in% plotted_factors)) {
      # Subset data for InvSimpson and the significant factor
      data_subset <- AD[variable == "InvSimpson", .(value, factor_value = get(factor_name), variable = "InvSimpson")]
      
      # Convert factor_value to a factor
      data_subset$factor_value <- factor(data_subset$factor_value)
      
      # Plot box plot
      set.seed(123)
      plot_AD <- ggplot(data = data_subset, aes(x = factor_value, y = value, fill = factor_value)) +
        geom_boxplot(width = 0.7, outlier.colour = 'white', position = position_dodge(width = 0.75)) +
        geom_jitter(size = 1, position = position_jitter(width = 0.1)) +
        ylab('Inverse Simpson Index') +
        theme_minimal() +
        facet_grid(~variable, scales = "free_x", switch = "y") +
        theme(
          strip.text.y = element_blank(),
          strip.text.x = element_blank(),
          axis.text.y = element_text(size = 8),
          axis.ticks.y = element_blank()
        ) +
        labs(title = paste("Boxplot for", factor_name))
      
      # Save plot
      print(plot_AD)
      ggsave(paste0("Figures/RPS10/", factor_name, "alphadiversity_invsimp_boxplot_prepost.pdf"), plot_AD, width = 10, height = 12)
      
      # Add factor name to the plotted factors vector
      plotted_factors <- c(plotted_factors, factor_name)
    }
  }
  
  # Add ANOVA results for the current factor to the list
  anova_list[[factor_name]] <- anova
}

# Print ANOVA results
for (factor_name in factors2) {
  cat(paste("ANOVA Results for Factor:", factor_name), "\n")
  print(knitr::kable(anova_list[[factor_name]], format = "markdown"))
  cat("\n")
}
```

```{r compiled alpha diversity table}
# Initialize an empty data.table to store combined ANOVA results for all factors
final_anova_table <- data.table(factor_name = character(), alpha_index = character(), pvalue = numeric(), df_between = numeric(), df_within = numeric(), ss_between = numeric(), ss_within = numeric(), ms_between = numeric(), ms_within = numeric(), F_value = numeric())

for (factor_name in factors2) {
  # Initialize an empty data.table to store ANOVA results for each alpha diversity measure
  anova <- data.table(factor_name = character(), alpha_index = character(), pvalue = numeric(), df_between = numeric(), df_within = numeric(), ss_between = numeric(), ss_within = numeric(), ms_between = numeric(), ms_within = numeric(), F_value = numeric())
  
  # Iterate through each alpha diversity measure
  for (alphad in c("Observed","InvSimpson", "Shannon")) {
    # Perform ANOVA
    anova_res <- aov(value ~ get(factor_name), data = AD[variable == alphad])
    
    # Extract relevant statistics from ANOVA results
    anova_summary <- summary(anova_res)
    
    # Extract p-value from ANOVA results
    p_value <- anova_summary[[1]]$`Pr(>F)`[1]
    
    # Extract other statistics
    df_between <- anova_summary[[1]]$"Df"[1]
    df_within <- anova_summary[[1]]$"Df"[2]
    ss_between <- anova_summary[[1]]$"Sum Sq"[1]
    ss_within <- anova_summary[[1]]$"Sum Sq"[2]
    ms_between <- anova_summary[[1]]$"Mean Sq"[1]
    ms_within <- anova_summary[[1]]$"Mean Sq"[2]
    F_value <- anova_summary[[1]]$"F value"[1]
    
    # Store statistics in the data.table
    anova <- rbind(anova, data.table(factor_name = factor_name, alpha_index = alphad, pvalue = p_value, df_between = df_between, df_within = df_within, ss_between = ss_between, ss_within = ss_within, ms_between = ms_between, ms_within = ms_within, F_value = F_value))
  }
  
  # Append ANOVA results for the current factor to the final table
  final_anova_table <- rbind(final_anova_table, anova)
}

# Print final ANOVA table
print(knitr::kable(final_anova_table, format = "markdown"))
write.csv(final_anova_table, "Results/RPS10/final_anova_table_prepost.csv", row.names = FALSE)
```

```{r alpha diversity faceted plots}
# Estimate alpha diversity

AD <- estimate_richness(pre_post.ps, measures = c("Observed", "InvSimpson", "Shannon"))
rownames(AD) <- gsub("\\.", "-", rownames(AD))

# Combine richness estimates with sample metadata
AD <- cbind(AD, sample_data(pre_post.ps)[rownames(AD), c("YearSam_EP")])


variables <- c("YearSam_EP")

pre_post.ps@sam_data$YearSam_EP <- as.factor(pre_post.ps@sam_data$YearSam_EP)

# Create an empty list to store plots
plots <- list()

# Loop through each variable and create the plot
for (variable in variables) {
  p <- plot_richness(pre_post.ps, x = variable, measures = c("Shannon", "InvSimpson", "Observed")) + 
    geom_boxplot()
  plots[[variable]] <- p
}

# Arrange plots in a grid
plots_arrange <- gridExtra::grid.arrange(
  plots[["YearSam_EP"]],
  nrow = length(variables)
)

# Add annotations to the top of each plot
plots_arrange_with_text <- cowplot::ggdraw() +
  cowplot::draw_plot(plots_arrange) +
  cowplot::draw_plot("YearSam_EP", x = 0.28, y = 0.98)

# Save the arranged plots
ggsave(plots_arrange_with_text, path = "Figures/RPS10", filename = "alpha_diversity_each_factor_prepost.pdf", 
       width = 12, height = 16)
```

```{r alpha diversity compilation with tukey summary}
# Define your variables of interest
# Define your variables of interest
variables <- c("YearSam_EP")
measures <- c("Observed", "InvSimpson", "Shannon")

# Create an empty list to store ANOVA and Tukey's results
tukey_results <- list()

# Loop through each variable and measure
for (variable in variables) {
  for (measure in measures) {
    # Subset the data for the current variable and measure
    data_subset <- AD[, c(variable, measure)]
    
    # Perform ANOVA
    formula <- as.formula(paste(measure, "~ ."))
    anova_res <- aov(formula, data = data_subset)
    
    # Perform Tukey's post-hoc test
    tukey_result <- HSD.test(anova_res, variable)
    
    # Store Tukey's results
    print(tukey_result)
  }
}
```

Beta diversity

Let's now examine beta diversity through ordination plots.
PERMANOVA analyses-ASV counts transformed to rel. abundance
```{r PERMANOVA ANALYSES-relative abundance, include=TRUE}
#I used the filtered dataset (removed low abundance counts less than 2)
mapping <- as.data.frame(sample_data(strat_rel_prepost.ps))

mapping2=mapping[, c("Treatuntreat_EP", "YearSam_EP", "Fire_EP", "Burned1.0_EP", "VegBin", "SoilBin")]

# use an sapply loop as it's much more efficient than a for loop! sapply is a function that goes through each element of the first parameter (in our case colnames(mapping)), and passes each one to a custom function.
# You can declare the custom function in the sapply function RPS10elf, using the following syntax:

ASV_tables_bray <- phyloseq::distance(strat_rel_prepost.ps, method = "bray")

df_metadata <- data.frame(sample_data(strat_rel_prepost.ps))

df_metadata<-df_metadata[, c("Treatuntreat_EP", "YearSam_EP", "Fire_EP", "Burned1.0_EP", "VegBin", "SoilBin")]

pvals <- sapply(colnames(mapping2), function(variable){
  
  #Create formula for adonis() 
  form1<-as.formula(paste("ASV_tables_bray", variable, sep="~"))
  
  #run adonis with formula (ASV distance table ~ variable of interest) with sample information, for 9999 permutaions
  res <- adonis2(form1, data = df_metadata, permutations = 9999)
  
  # extract pvalue from adonis output
  pval <- res$`Pr(>F)`[[1]] # we take the first p value which corresponds to the variable of interest
  return(pval)
})

#Adjust p_values with fdr correction
pval_adj <- p.adjust(pvals, method = "fdr")

#Find significant variables with lower than 0.05 for their p values
save.pval.adonis0.05_braycurtis <- pval_adj[pval_adj < 0.05]
save.pval.adonis0.05_braycurtis

print(pval_adj)
```

```{r compile the stats}
ASV_tables_bray <- phyloseq::distance(strat_rel_prepost.ps, method = "bray")
df_metadata <- data.frame(sample_data(strat_rel_prepost.ps))
df_metadata <- df_metadata[, c("Treatuntreat_EP", "YearSam_EP", "Fire_EP", "Burned1.0_EP", "VegBin", "SoilBin")]

# Perform PERMANOVA and beta dispersion analysis, and extract statistics
permanova_results <- lapply(colnames(df_metadata), function(variable) {
  # PERMANOVA
  res_permanova <- adonis2(ASV_tables_bray ~ ., data = df_metadata[, variable, drop = FALSE], permutations = 9999)
  r_squared <- res_permanova$R2[1] # Extract R-squared value for PERMANOVA
  f_value <- res_permanova$F[1] # Extract F-value for PERMANOVA
  pval_permanova <- res_permanova$`Pr(>F)`[[1]] # Extract p-value for PERMANOVA
  
  # Beta dispersion analysis
  res_betadisper <- betadisper(d = ASV_tables_bray, group = df_metadata[, variable])
  perm_test_res <- permutest(res_betadisper)
  pval_betadisper <- perm_test_res$tab$`Pr(>F)`[[1]] # Extract p-value for beta dispersion
  
  # Combine results
  results <- data.frame(Factor = variable, F_Value = f_value, R_Squared = r_squared, P_Value_BetaDispersion = pval_betadisper, P_Value_PERMANOVA = pval_permanova)
  return(results)
})

# Combine results into a data frame
permanova_results_df <- do.call(rbind, permanova_results)

# Adjust p-values for PERMANOVA
permanova_results_df$Adjusted_P_Value_PERMANOVA_fdr <- p.adjust(permanova_results_df$P_Value_PERMANOVA, method = "fdr")

permanova_results_df$Adjusted_P_Value_PERMANOVA_bonferroni <- p.adjust(permanova_results_df$P_Value_PERMANOVA, method = "bonferroni")

# Save as CSV file
write.csv(permanova_results_df, file = "Results/RPS10/PERMANOVA_results_prepost.csv", row.names = FALSE)
```

What about interactions between factors?
```{r pairwise comparisons}
# Calculate PERMANOVA
ASV_tables_bray <- phyloseq::distance(strat_rel_prepost.ps, method = "bray")
df_metadata <- data.frame(sample_data(strat_rel_prepost.ps))
df_metadata <- df_metadata[, c("Treatuntreat_EP", "YearSam_EP", "Fire_EP", "Burned1.0_EP", "VegBin", "SoilBin")]

# Define function to perform PERMANOVA for pairwise combinations of factors
perform_permutation <- function(factor1, factor2) {
  interaction <- paste(factor1, factor2, sep = " * ")
  res <- adonis2(ASV_tables_bray ~ ., data = df_metadata[, c(factor1, factor2)], permutations = 9999)
  return(list(interaction = interaction, p_value = res$`Pr(>F)`[[1]]))
}

# Perform pairwise comparisons between factors
significant_interactions <- sapply(names(df_metadata), function(factor1) {
  sapply(names(df_metadata), function(factor2) {
    if (factor1 != factor2) {
      perform_permutation(factor1, factor2)
    } else {
      NULL
    }
  })
}, simplify = "array")

# Filter out NULLs and combine results
significant_interactions <- significant_interactions[!sapply(significant_interactions, is.null)]

# Print significant interactions
print(significant_interactions)

interactions_df <- do.call(rbind.data.frame, significant_interactions)
print(significant_interactions)

interactions_df <- do.call(rbind.data.frame, significant_interactions)


adjusted_p_values <- p.adjust(interactions_df$p_value, method = "fdr")
# Add adjusted p-values to the data frame
interactions_df$p_value <- adjusted_p_values


# Extract unique factors
unique_factors <- unique(unlist(strsplit(interactions_df$interaction, " \\* ")))

# Create an empty matrix to store p-values
interaction_matrix <- matrix(NA, nrow = length(unique_factors), ncol = length(unique_factors), dimnames = list(unique_factors, unique_factors))

# Fill the matrix with p-values
for (i in 1:nrow(interactions_df)) {
  factors <- unlist(strsplit(interactions_df$interaction[i], " \\* "))
  interaction_matrix[factors[1], factors[2]] <- interactions_df$p_value[i]
  interaction_matrix[factors[2], factors[1]] <- interactions_df$p_value[i]
}

rownames(interaction_matrix) <- unique_factors 
colnames(interaction_matrix) <- unique_factors 


# Convert the matrix to a data frame for ggplot
interaction_df <- as.data.frame(as.table(interaction_matrix))
interaction_df <- as.data.frame(as.table(interaction_matrix))
interaction_df$color <- ifelse(is.na(interaction_df$Freq), "gray",
                               ifelse(interaction_df$Freq <= 0.01, "darkblue",
                                      ifelse(interaction_df$Freq <= 0.05, "lightblue", "white")))

# Plot heatmap with continuous color scale
heatmap <- ggplot(data = interaction_df, aes(x = Var1, y = Var2, fill = color)) +
  geom_tile(color = "black") +  # Add border for each tile
  scale_fill_manual(values = c("gray", "darkblue", "white"),
                    name = "P-value",
                    labels = c("NA", "P < 0.01", "P > 0.05"),
                    na.value = "gray") +  # Set manual color scale and legend labels
  scale_fill_identity(name = NULL) +  # Use identity scale for custom colors
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 10),  # Adjust x-axis text angle and size
        axis.text.y = element_text(size = 10),  # Adjust y-axis text size
        panel.grid = element_blank(),  # Remove gridlines
        panel.border = element_blank(),  # Remove border around the plot
        axis.line = element_blank(),  # Remove axis lines
        legend.position = "bottom",  # Position legend at the bottom
        legend.direction = "horizontal",  # Horizontal legend
        legend.key.size = unit(0.5, "cm"),  # Adjust legend key size
        legend.title = element_blank(),  # Remove legend title
        legend.text = element_text(size = 8))  # Adjust legend text font size

# Assign row and column names
heatmap <- heatmap + labs(x = NULL, y = NULL)

# Print heatmap
print(heatmap)

# Save the heatmap as a PDF file
ggsave("Results/RPS10/permanova_interactions_heatmap_prepost.pdf", plot = heatmap, width = 8, height = 6)
```

Let's now examine beta diversity through ordination plots. We'll will examine outputs for rarefied data as well as data adjusted by relative abundance

We'll now make ordination plots on data for which we've calculated relative abundance
```{r Beta diversity-Rel. abundance treated and untreated samples}
#Relative abundance
ps_NMDS3 = ordinate(strat_rel_prepost.ps, 
                   method="NMDS", 
                   distance="bray")

plot_3<- plot_ordination(strat_rel_prepost.ps, 
                         ps_NMDS3,
                         color= "Treatuntreat_EP") +
  ggtitle("NMDS ordination plot - relative abundance") +
  labs(color = "Factor") + 
  theme(axis.text=element_text(size=14), axis.title=element_text(size=14), title =  element_text(size=13), legend.text = element_text(size = 18))

plot_ellipses<-plot_3 +
   stat_ellipse(type = "norm", linetype = 2) +
  theme_bw()

ggsave("Figures/RPS10/NMDS_ordination_treatuntreat_prepost.pdf", plot_ellipses, width = 10, height = 10)

plot_ellipses

#Export NMDS coordinates per sample
write.csv(ps_NMDS3[["points"]], file="Results/RPS10/RPS10_NMDS_coordinates_persample.csv",row.names = TRUE)
```

Now, we'll look at ordination plot (rel abundance dataset), with fire factor
```{r Beta diversity-Rel. abundance by YrSinceTrt_EP factor}
#Relative abundance
ps_NMDS3 = ordinate(strat_rel_prepost.ps, 
                   method="NMDS",
                   distance="bray")

plot_4<- plot_ordination(strat_rel_prepost.ps, 
                         ps_NMDS3,
                         labs(color = "Factor"),
                         color= "Fire_EP") +
  ggtitle("NMDS ordination plot - relative abundance") +
  theme(axis.text=element_text(size=14), axis.title=element_text(size=14), title =  element_text(size=13), legend.text = element_text(size = 18))


plot_ellipses<-plot_4 +
   stat_ellipse(type = "norm", linetype = 2) +
  theme_bw()

plot_ellipses
```

```{r ad-hoc-functions}
sample.data.table <- function(ps) {
  as(sample_data(ps), "data.frame") %>%
    as.data.table(keep.rownames = "Sample") %>% 
    setkey(Sample) %>%
    return()
}
asv.data.table <- function(ps) {
  if (taxa_are_rows(ps)) {
    otu.mat <- t(as(otu_table(ps), "matrix"))
  } else {
    otu.mat <- as(otu_table(ps), "matrix")
  }
  return(as.data.table(otu.mat, keep.rownames = "Sample") %>% setkey(Sample))
}
cat.n <- function(x) cat(x, sep = "\n")
```

Below, we import the data we'll be using for the rest of the practical, and put it into a format that makes working with it easier. This section shouldn't need to be modified.
```{r data-import-and-prep}
pre_post.data <- sample.data.table(pre_post_5asv.ps)
pre_post.asvs <- asv.data.table(pre_post_5asv.ps)
pre_post.asv.ids <- taxa_names(pre_post_5asv.ps)
pre_post.dt <- pre_post.data[pre_post.asvs]
# Assuming you have the factor as "Untreated" and "Treated" currently
pre_post_5asv.ps@sam_data$Treatuntreat_EP <- factor(pre_post_5asv.ps@sam_data$Treatuntreat_EP, levels = c("untreated", "treated"))

pre_post_5asv.ps@sam_data$Fire_EP <- factor(pre_post_5asv.ps@sam_data$Fire_EP, levels = c("Unburned", "SODTrt"))
```

For differential abundance analyses of treated vs. untreated plots: we will serially conduct Wilcoxon tests on the pre_post data set. This test is the non-parametric (distribution-free) version of the Student's *t*-test and will tell us if the abundances of each taxon different significantly between our two groups: Unexposed vs Exposed. After running the tests, we correct for multiple tests by adjusting the p-values. There are two examples of methods for adjusting the p-values provided.
```{r wilcoxon-tests}
wilcox.tests <- lapply(pre_post.asv.ids, function(asv) {
  frm <- paste(asv, "~ Fire_EP") # this combined with the as.formula() function below makes it possible to generate formulas in R programmatically
  return(
    wilcox.test(as.formula(frm), data = pre_post.dt)
  )
})
pre_post.raw.pvals <- sapply(wilcox.tests, function(mod) {return(mod$p.value)})

pre_post.bon.pvals <- p.adjust(pre_post.raw.pvals, method = "bonferroni")
# Printing the number of associations with adjusted p-values below our 0.05 cutoff. This value can be adjusted depending on your data and expectations
paste("Significant associations:", length(pre_post.bon.pvals[pre_post.bon.pvals <= 0.05])) %>%
  cat.n()

pre_post.fdr.pvals <- p.adjust(pre_post.raw.pvals, method = "fdr")
paste("Significant associations:", length(pre_post.fdr.pvals[pre_post.fdr.pvals <= 0.05])) %>%
  cat.n()
```

We will now plot the results
```{r plot}
pre_post.melt.dt <- melt(
  pre_post.dt, 
  measure.vars = pre_post.asv.ids, 
  variable.name = "ASV", 
  value.name = "Abundance"
)
sig.asv <- pre_post.asv.ids[pre_post.bon.pvals <= 0.05]

pre_post.plot.dt <- pre_post.melt.dt[ASV %in% sig.asv]
pre_post.plot.dt[, Treatuntreat_EP := factor(Treatuntreat_EP, levels = c("untreated", "treated"))] # the default is for levels to be in alphanumeric order, this way the control group, Unexposed, will be first along the x-axes

pre_post.plot <- ggplot(pre_post.plot.dt, aes(x = Fire_EP, y = Abundance)) + 
  geom_beeswarm(alpha = 0.6) + 
  stat_summary(
    fun.data = "mean_cl_boot", # This summary statistical calculates the 95% C.I. of the mean via non-parametric methods
    color = "red", 
    geom = "errorbar", 
    width = 0.8

  ) + 
  facet_wrap(~ ASV, scales = "free_y") +
  theme(axis.text.x = element_text(angle = 30, hjust = 1))

#Cannot plot as now sig associations
#pre_post.plot 
```

Now we will look at an example where we have more than two factors and use kruskal wallis test
```{r kw test1}
kw.tests <- lapply(pre_post.asv.ids, function(asv) {
  frm <- paste(asv, "~ Fire_EP")
  return(
    kruskal.test(as.formula(frm), data = pre_post.dt)
  )
})

pre_post.raw.pvals <- sapply(kw.tests, function(mod) {return(mod$p.value)})

pre_post.bon.pvals <- p.adjust(pre_post.raw.pvals, method = "bonferroni")
paste(
  "Significant associations:", 
  length(pre_post.bon.pvals[pre_post.bon.pvals <= 0.05])
) %>% 
  cat.n()


pre_post.fdr.pvals <- p.adjust(pre_post.raw.pvals, method = "fdr")
paste(
  "Significant associations:", 
  length(pre_post.fdr.pvals[pre_post.fdr.pvals <= 0.05])
) %>% 
  cat.n()
```

```{r plotting-kruskal-wallis-tests}
#pre_post.melt.dt <- melt(
  #pre_post.dt, 
  #measure.vars = pre_post.asv.ids, 
  #variable.name = "ASV", 
  #value.name = "Abundance"
#)
#sig.asvs <- pre_post.asv.ids[pre_post.fdr.pvals <= 0.05] # or pre_post.fdr.pvals, Their choice
#pre_post.plot.dt <- pre_post.melt.dt[ASV %in% sig.asvs]

#ggplot(pre_post.plot.dt, aes(x = YearSam_EP, y = Abundance)) + 
  #geom_beeswarm(alpha = 0.6) + 
  #stat_summary(
    #fun.data = "mean_cl_boot", 
    #color = "red", 
    #geom = "errorbar", 
    #width = 0.8
  #) + 
  #facet_wrap(~ ASV, scales = "free_y")
```

```{r deseq2-treatuntreat}
pre_post_5asv.ps@sam_data$Treatuntreat_EP <- factor(pre_post_5asv.ps@sam_data$Treatuntreat_EP, labels = c("untreated", "treated"))

pre_post.deseq <- phyloseq_to_deseq2(pre_post_5asv.ps, ~ Treatuntreat_EP) # this function does all the heavy lifting for us to turn a phyloseq object into a DESeq2 object


#This was required so DESEQ doesn't error out (there are several approaches, but the deseq developer recommended this most recently, if going from phyloseq to deseq option)
pre_post.dds <- estimateSizeFactors(pre_post.deseq, type = 'poscounts')
pre_post.dds = DESeq(pre_post.dds)
pre_post.res.dt <- results(pre_post.dds) %>% as.data.table(keep.rownames = "ASV")

pre_post.sig.dt <- pre_post.res.dt[padj <= 0.05]

pre_post.sig.dt = cbind(as(pre_post.sig.dt, "data.frame"), as(tax_table(pre_post_5asv.ps)[pre_post.sig.dt$ASV, ], "matrix"))

deseqplot2 <- ggplot(pre_post.sig.dt,  aes(y = reorder(Species, log2FoldChange), x = log2FoldChange, fill=Family)) +
  geom_vline(xintercept = 0.0, color = "gray", size = 0.5) +
  geom_bar(stat = "identity", position = "dodge", width = 0.8) +  
  labs(x = "ASV", y = "Log2 Fold Change", title = "Treated vs Untreated Plots") + 
  theme(axis.text.x = element_text(angle = -90, hjust = 0, vjust = 0.5))
ggsave("Figures/RPS10/deseq_trtuntrt_barplot_prepost.pdf", deseqplot2, width = 10, height = 12)

deseqplot2
```
```{r deseq2-treatuntreat_2}
pre_post_5asv.ps@sam_data$Treatuntreat_EP <- factor(pre_post_5asv.ps@sam_data$Fire_EP, labels = c("Unburned", "SODTrt"))

pre_post.deseq <- phyloseq_to_deseq2(pre_post_5asv.ps, ~ Fire_EP) # this function does all the heavy lifting for us to turn a phyloseq object into a DESeq2 object


#This was required so DESEQ doesn't error out (there are several approaches, but the deseq developer recommended this most recently, if going from phyloseq to deseq option)
pre_post.dds <- estimateSizeFactors(pre_post.deseq, type = 'poscounts')
pre_post.dds = DESeq(pre_post.dds)
pre_post.res.dt <- results(pre_post.dds) %>% as.data.table(keep.rownames = "ASV")

pre_post.sig.dt <- pre_post.res.dt[padj <= 0.05]

pre_post.sig.dt = cbind(as(pre_post.sig.dt, "data.frame"), as(tax_table(pre_post_5asv.ps)[pre_post.sig.dt$ASV, ], "matrix"))

deseqplot3 <- ggplot(pre_post.sig.dt,  aes(y = reorder(Species, log2FoldChange), x = log2FoldChange, fill=Family)) +
  geom_vline(xintercept = 0.0, color = "gray", size = 0.5) +
  geom_bar(stat = "identity", position = "dodge", width = 0.8) +  
  labs(x = "ASV", y = "Log2 Fold Change", title = "Unburned vs SODtrt plots") + 
  theme(axis.text.x = element_text(angle = -90, hjust = 0, vjust = 0.5))
ggsave("Figures/RPS10/deseq_unburned_sodtrt_barplot_prepost.pdf", deseqplot3, width = 10, height = 12)

deseqplot3
```
## Software used
```{r}
sessioninfo::session_info()
```