---
title: "SOD compiled figure and tables--alpha and beta diversity"
output:
  html_document: default
  pdf_document: default
date: "2025-11-17"
---

```{r setup}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(root.dir = "~/SOD_Microbiome")
library(phyloseq); packageVersion("phyloseq")
library(ggplot2); packageVersion("ggplot2")
library(readr); packageVersion("readr")
library(purrr); packageVersion("purrr")
library(furrr); packageVersion("furrr")
library(dplyr); packageVersion("dplyr")
library(stringr); packageVersion("stringr")
library(metacoder); packageVersion("metacoder")
library(data.table); packageVersion("data.table")
library(Biostrings); packageVersion("Biostrings")
library(vegan);packageVersion("vegan")
library(agricolae);packageVersion("agricolae")
library(magrittr);packageVersion("magrittr")
library(data.table);packageVersion("data.table")
library(cowplot);packageVersion("cowplot")
library(ggpubr);packageVersion("ggpubr")
library(viridis);packageVersion("viridis")

```

### Load data 16S

```{r subset data 16S, include=TRUE}
load("Data/16S/filtered_data/sod_5asv_nowildfire.RData")
sod_5asv_16S_nowf.ps<-sod_5asv_nowildfire.ps

sod_5asv_16S_nowf.ps@sam_data$Treatuntreat_EP <- factor(sod_5asv_16S_nowf.ps@sam_data$Treatuntreat_EP, 
                                            levels = c("untreated", "treated"), 
                                                  labels=c("SOD-UT", "SOD-T"))

sod_5asv_16S_nowf.ps@sam_data$YrSinceTrt_BR <- factor(sod_5asv_16S_nowf.ps@sam_data$YrSinceTrt_BR, 
                                            levels = c("untreated", "1to2", "3to4", "5to8", "10to12"), 
                                                  labels=c("SOD-UT", "1-2 PT", "3-4 PT", "5-8 PT", "10-12 PT"))


load("Data/16S/filtered_data/sodexp.RData")
sod_16S.ps<-sod.ps
sod_16S.ps@sam_data$Fire_Mod <- factor(sod_16S.ps@sam_data$Fire_Mod, 
                                              levels = c("untreated","SODtrt","wildfire"), 
                                                  labels=c("SOD-UT","SOD-T","Wildfire"))

load("Data/16S/filtered_data/sod_5asv.RData")
sod_5asv_16S.ps<-sod_5asv.ps
sod_5asv_16S.ps@sam_data$Fire_Mod <- factor(sod_5asv_16S.ps@sam_data$Fire_Mod,
                                            levels = c("untreated","SODtrt","wildfire"), 
                                                  labels=c("SOD-UT","SOD-T","Wildfire"))

load("Data/16S/filtered_data/sodexp_nowildfire.RData")
sod_16S_nowf.ps<-sod_nowildfire.ps
sod_16S_nowf.ps@sam_data$Fire_Mod <- factor(sod_16S_nowf.ps@sam_data$Fire_Mod, 
                                       levels = c("untreated", "SODtrt","wildfire"), 
                                                  labels=c("SOD-UT", "SOD-T","Wildfire"))
```

### Load data ITS

```{r subset data ITS, include=TRUE}
load("Data/ITS/filtered_data/sod_5asv_nowildfire.RData")
sod_5asv_its_nowf.ps<-sod_5asv_nowildfire.ps

sod_5asv_its_nowf.ps@sam_data$Treatuntreat_EP <- factor(sod_5asv_its_nowf.ps@sam_data$Treatuntreat_EP, 
                                            levels = c("untreated", "treated"), 
                                                  labels=c("SOD-UT", "SOD-T"))

sod_5asv_its_nowf.ps@sam_data$YrSinceTrt_BR <- factor(sod_5asv_its_nowf.ps@sam_data$YrSinceTrt_BR, 
                                            levels = c("untreated", "1to2", "3to4", "5to8", "10to12"), 
                                                  labels=c("SOD-UT", "1-2 PT", "3-4 PT", "5-8 PT", "10-12 PT"))

load("Data/ITS/filtered_data/sodexp.RData")
sod_its.ps<-sod.ps
sod_its.ps@sam_data$Fire_Mod <- factor(sod_its.ps@sam_data$Fire_Mod, 
                                       levels = c("untreated", "SODtrt","wildfire"), 
                                                  labels=c("SOD-UT", "SOD-T","Wildfire"))

load("Data/ITS/filtered_data/sod_5asv.RData")
sod_5asv_its.ps<-sod_5asv.ps
sod_5asv_its.ps@sam_data$Fire_Mod <- factor(sod_5asv_its.ps@sam_data$Fire_Mod, 
                                            levels = c("untreated","SODtrt","wildfire"), 
                                                  labels=c("SOD-UT","SOD-T","Wildfire"))

load("Data/ITS/filtered_data/sodexp_nowildfire.RData")
sod_its_nowf.ps<-sod_nowildfire.ps
sod_its_nowf.ps@sam_data$Fire_Mod <- factor(sod_its_nowf.ps@sam_data$Fire_Mod, 
                                       levels = c("untreated", "SODtrt","wildfire"), 
                                                  labels=c("SOD-UT", "SOD-T","Wildfire"))
```

### Load data rps10

```{r load data rps10, include=TRUE}
load("Data/RPS10/filtered_data/sod_5asv_nowildfire.RData")
sod_5asv_rps10_nowf.ps<-sod_5asv_nowildfire.ps

sod_5asv_rps10_nowf.ps@sam_data$Treatuntreat_EP <- factor(sod_5asv_rps10_nowf.ps@sam_data$Treatuntreat_EP, 
                                            levels = c("untreated", "treated"), 
                                                  labels=c("SOD-UT", "SOD-T"))

sod_5asv_rps10_nowf.ps@sam_data$YrSinceTrt_BR <- factor(sod_5asv_rps10_nowf.ps@sam_data$YrSinceTrt_BR, 
                                            levels = c("untreated", "1to2", "3to4", "5to8", "10to12"), 
                                                  labels=c("SOD-UT", "1-2 PT", "3-4 PT", "5-8 PT", "10-12 PT"))

load("Data/RPS10/filtered_data/sod_pruned.RData")
sod_rps10.ps<-sod_prunereads.ps
sod_rps10.ps@sam_data$Fire_Mod <- factor(sod_rps10.ps@sam_data$Fire_Mod, 
                                                levels = c("untreated","SODtrt","wildfire"), 
                                                  labels=c("SOD-UT","SOD-T","Wildfire"))


load("Data/RPS10/filtered_data/sod_prunereads_5asv.RData")
sod_5asv_rps10.ps<-sod_prunereads_5asv.ps
sod_5asv_rps10.ps@sam_data$Fire_Mod <- factor(sod_5asv_rps10.ps@sam_data$Fire_Mod, 
                                                     levels=
                                                c("untreated","SODtrt","wildfire"),
                                              labels=c("SOD-UT","SOD-T","Wildfire"))

load("Data/RPS10/filtered_data/sodexp_nowildfire.RData")
sod_rps10_nowf.ps<-sod_nowildfire.ps
sod_rps10_nowf.ps@sam_data$Fire_Mod <- factor(sod_rps10_nowf.ps@sam_data$Fire_Mod, 
                                       levels = c("untreated", "SODtrt","wildfire"), 
                                                  labels=c("SOD-UT", "SOD-T","Wildfire"))
```

### Let's calculate alpha diversity metrics for ITS dataset

```{r, combine alpha diversity boxplots its}
set.seed(1)

estimate_and_clean <- function(ps_obj) {
  richness <- estimate_richness(ps_obj, measures = c("InvSimpson"))
  rownames(richness) <- gsub("\\.", "-", rownames(richness))
  return(richness)
}

AD_its_nowf <- estimate_and_clean(sod_its_nowf.ps)
AD_its <- estimate_and_clean(sod_its.ps)

AD_its_nowf <- cbind(AD_its_nowf, sample_data(sod_5asv_its_nowf.ps)[row.names(AD_its_nowf), c("Treatuntreat_EP", "YrSinceTrt_BR")])
AD_its <- cbind(AD_its, sample_data(sod_its.ps)[row.names(AD_its), "Fire_Mod"])

variables <- list(nowf = c("Treatuntreat_EP", "YrSinceTrt_BR"), wf = c("Fire_Mod"))
measures <- c("InvSimpson")

create_plot <- function(ps_obj, variable) {
  phyloseq::plot_richness(ps_obj, x = variable, measures = c("InvSimpson")) + 
    geom_boxplot() +
    theme_classic() +  
    theme(
      panel.border = element_rect(color = "black", fill = NA, size = 1),  
      axis.line = element_blank(),  
      plot.margin = unit(c(1, 1, 1, 1), "lines"),  
      plot.title = element_text(margin = margin(b = 10)),  
      axis.title.x = element_text(margin = margin(t = 10)),  
      axis.title.y = element_text(margin = margin(r = 10)), 
      axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1)
    )
}

plots_its <- list()
for (variable in variables$nowf) {
  plots_its[[variable]] <- create_plot(sod_5asv_its_nowf.ps, variable)
}
for (variable in variables$wf) {
  plots_its[[variable]] <- create_plot(sod_its.ps, variable)
}

get_tukey_and_anova <- function(data, measure, variable) {
  formula <- as.formula(paste(measure, "~", variable))
  aov_result <- aov(formula, data = data)
  tukey_result <- HSD.test(aov_result, variable, group = TRUE)
  letters <- tukey_result$groups$groups
  names(letters) <- rownames(tukey_result$groups)  
  return(list(letters = letters, anova = summary(aov_result)))
}

perform_tukey_tests <- function(data, variables, measures) {
  results <- list()
  for (variable in variables) {
    for (measure in measures) {
      tryCatch({
        result <- get_tukey_and_anova(data, measure, variable)
        results[[paste(measure, variable, sep = "_")]] <- result
      }, error = function(e) {
        message("Error in Tukey test for ", measure, " by ", variable, ": ", e$message)
      })
    }
  }
  return(results)
}

tukey_results_ITS <- perform_tukey_tests(AD_its, variables$wf, measures)
tukey_results_ITS_nowf <- perform_tukey_tests(AD_its_nowf, variables$nowf, measures)

print(tukey_results_ITS)
print(tukey_results_ITS_nowf)

# Function to print ANOVA stats
print_anova_stats <- function(results) {
  for (key in names(results)) {
    cat("ANOVA results for", key, ":\n")
    print(results[[key]]$anova)
    cat("\n")
  }
}

print_anova_stats(tukey_results_ITS)
print_anova_stats(tukey_results_ITS_nowf)
```

### Let's calculate alpha diversity metrics for 16S dataset

```{r, combine alpha diversity boxplots 16S}
set.seed(1)

estimate_and_clean <- function(ps_obj) {
  richness <- estimate_richness(ps_obj, measures = c("InvSimpson"))
  rownames(richness) <- gsub("\\.", "-", rownames(richness))
  return(richness)
}

AD_16S_nowf <- estimate_and_clean(sod_16S_nowf.ps)
AD_16S <- estimate_and_clean(sod_16S.ps)

AD_16S_nowf <- cbind(AD_16S_nowf, sample_data(sod_5asv_16S_nowf.ps)[row.names(AD_16S_nowf), c("Treatuntreat_EP", "YrSinceTrt_BR")])
AD_16S <- cbind(AD_16S, sample_data(sod_16S.ps)[row.names(AD_16S), "Fire_Mod"])

variables <- list(nowf = c("Treatuntreat_EP", "YrSinceTrt_BR"), wf = c("Fire_Mod"))
measures <- c("InvSimpson")

create_plot <- function(ps_obj, variable) {
  phyloseq::plot_richness(ps_obj, x = variable, measures = c("InvSimpson")) + 
    geom_boxplot() +
    theme_classic() +  
    theme(
      panel.border = element_rect(color = "black", fill = NA, size = 1),  
      axis.line = element_blank(),  
      plot.margin = unit(c(1, 1, 1, 1), "lines"),  
      plot.title = element_text(margin = margin(b = 10)),  
      axis.title.x = element_text(margin = margin(t = 10)),  
      axis.title.y = element_text(margin = margin(r = 10)), 
      axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1)
    )
}

plots_16S <- list()
for (variable in variables$nowf) {
  plots_16S[[variable]] <- create_plot(sod_5asv_16S_nowf.ps, variable)
}
for (variable in variables$wf) {
  plots_16S[[variable]] <- create_plot(sod_16S.ps, variable)
}

get_tukey_and_anova <- function(data, measure, variable) {
  formula <- as.formula(paste(measure, "~", variable))
  aov_result <- aov(formula, data = data)
  tukey_result <- HSD.test(aov_result, variable, group = TRUE)
  letters <- tukey_result$groups$groups
  names(letters) <- rownames(tukey_result$groups) 
  return(list(letters = letters, anova = summary(aov_result)))
}


perform_tukey_tests <- function(data, variables, measures) {
  results <- list()
  for (variable in variables) {
    for (measure in measures) {
      tryCatch({
        result <- get_tukey_and_anova(data, measure, variable)
        results[[paste(measure, variable, sep = "_")]] <- result
      }, error = function(e) {
        message("Error in Tukey test for ", measure, " by ", variable, ": ", e$message)
      })
    }
  }
  return(results)
}

tukey_results_16S <- perform_tukey_tests(AD_16S, variables$wf, measures)
tukey_results_16S_nowf <- perform_tukey_tests(AD_16S_nowf, variables$nowf, measures)

print(tukey_results_16S)
print(tukey_results_16S_nowf)

# Function to print ANOVA stats
print_anova_stats <- function(results) {
  for (key in names(results)) {
    cat("ANOVA results for", key, ":\n")
    print(results[[key]]$anova)
    cat("\n")
  }
}

print_anova_stats(tukey_results_16S)
print_anova_stats(tukey_results_16S_nowf)
```

### Let's calculate alpha diversity metrics for rps10 dataset

```{r, combine alpha diversity boxplots rps10}
set.seed(1)

estimate_and_clean <- function(ps_obj) {
  richness <- estimate_richness(ps_obj, measures = c("InvSimpson"))
  rownames(richness) <- gsub("\\.", "-", rownames(richness))
  return(richness)
}

AD_rps10_nowf <- estimate_and_clean(sod_rps10_nowf.ps)
AD_rps10 <- estimate_and_clean(sod_rps10.ps)

AD_rps10_nowf <- cbind(AD_rps10_nowf, sample_data(sod_5asv_rps10_nowf.ps)[row.names(AD_rps10_nowf), c("Treatuntreat_EP", "YrSinceTrt_BR")])
AD_rps10 <- cbind(AD_rps10, sample_data(sod_rps10.ps)[row.names(AD_rps10), "Fire_Mod"])

variables <- list(nowf = c("Treatuntreat_EP", "YrSinceTrt_BR"), wf = c("Fire_Mod"))
measures <- c("InvSimpson")

create_plot <- function(ps_obj, variable) {
  phyloseq::plot_richness(ps_obj, x = variable, measures = c("InvSimpson")) + 
    geom_boxplot() +
    theme_classic() +  
    theme(
      panel.border = element_rect(color = "black", fill = NA, size = 1),  
      axis.line = element_blank(),  
      plot.margin = unit(c(1, 1, 1, 1), "lines"),  
      plot.title = element_text(margin = margin(b = 10)),  
      axis.title.x = element_text(margin = margin(t = 10)),  
      axis.title.y = element_text(margin = margin(r = 10)), 
      axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1)
    )
}

plots_rps10 <- list()
for (variable in variables$nowf) {
  plots_rps10[[variable]] <- create_plot(sod_5asv_rps10_nowf.ps, variable)
}
for (variable in variables$wf) {
  plots_rps10[[variable]] <- create_plot(sod_rps10.ps, variable)
}

get_tukey_and_anova <- function(data, measure, variable) {
  formula <- as.formula(paste(measure, "~", variable))
  aov_result <- aov(formula, data = data)
  tukey_result <- HSD.test(aov_result, variable, group = TRUE)
  letters <- tukey_result$groups$groups
  names(letters) <- rownames(tukey_result$groups) 
  return(list(letters = letters, anova = summary(aov_result)))
}

perform_tukey_tests <- function(data, variables, measures) {
  results <- list()
  for (variable in variables) {
    for (measure in measures) {
      tryCatch({
        result <- get_tukey_and_anova(data, measure, variable)
        results[[paste(measure, variable, sep = "_")]] <- result
      }, error = function(e) {
        message("Error in Tukey test for ", measure, " by ", variable, ": ", e$message)
      })
    }
  }
  return(results)
}

tukey_results_rps10 <- perform_tukey_tests(AD_rps10, variables$wf, measures)
tukey_results_rps10_nowf <- perform_tukey_tests(AD_rps10_nowf, variables$nowf, measures)

print(tukey_results_rps10)
print(tukey_results_rps10_nowf)

# Function to print ANOVA stats
print_anova_stats <- function(results) {
  for (key in names(results)) {
    cat("ANOVA results for", key, ":\n")
    print(results[[key]]$anova)
    cat("\n")
  }
}

print_anova_stats(tukey_results_rps10)
print_anova_stats(tukey_results_rps10_nowf)
```

### Let's make a more publication ready figure
```{r, combine all alpha trt box plots}
colors_treatuntreat <- c(
  "SOD-UT" = "#2271B2",  # deep blue
  "SOD-T"  = "#e69f00"   # rusty orange
)

colors_yrsince <- c(
  "SOD-UT"   = "#2271B2", # deep blue (anchor)
  "1-2 PT"   = "#D55E00", # rusty orange
  "3-4 PT"   = "#FF54ED", # cyan
  "5-8 PT"   = "#00B19F", # green
  "10-12 PT" = "#9EFF37"  # yellow (swap to #E69F00 if you need darker)
)


colors_fire <- c(
  "SOD-UT"   = "#2271B2", # deep blue
  "SOD-T"    = "#e69f00", # rusty orange
  "Wildfire" = "#F8071D"  # pink/magenta
)


modify_plot <- function(plot, remove_y_title = FALSE, colors, x_var, tukey_result = NULL, ylim = NULL, panel_label = NULL) {
  p <- plot +
    geom_boxplot(aes(fill = .data[[x_var]]), alpha = 0.8) +
    scale_fill_manual(values = colors) +
    theme_classic() +
    theme(
      panel.border = element_rect(color = "black", fill = NA, size = 0.5),
      axis.line = element_blank(),
      plot.margin = unit(c(0.5, 0.3, 0.3, 0.8), "lines"), # reduced bottom margin
      axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1, size = 12), # changed vjust to 1
      axis.text.y = element_text(size = 12),
      axis.title.x = element_blank(),
      axis.title.y = element_blank(),
      legend.position = "none",
      plot.title = element_blank(),
      strip.text = element_blank(),
      strip.background = element_blank()
    )
  
  if (remove_y_title) {
    p <- p + theme(axis.text.y = element_blank(), axis.ticks.y = element_blank())
  }
  
  if (!is.null(tukey_result)) {
    letters_df <- data.frame(group = names(tukey_result$letters), letter = tukey_result$letters)
    y_max <- if (!is.null(ylim)) ylim[2] else max(layer_data(p, 2)$ymax, na.rm = TRUE)
    range_span <- if (!is.null(ylim)) diff(ylim) else y_max
    if (range_span <= 12) {
      upper_limit <- y_max + 2.5
      letters_y <- upper_limit - 0.8
    } else {
      upper_limit <- y_max * 1.25
      letters_y <- upper_limit * 0.96
    }
    p <- p + coord_cartesian(ylim = c(ylim[1], upper_limit), clip = "off") + # added clip = "off"
      geom_text(data = letters_df, aes(x = group, y = letters_y, label = letter), inherit.aes = FALSE, size = 4)
  } else if (!is.null(ylim)) {
    p <- p + coord_cartesian(ylim = ylim, clip = "off") # added clip = "off"
  }
  
  if (!is.null(ylim) && ylim[2] == 9) {
    p <- p + scale_y_continuous(breaks = c(0, 5, 10))
  }
  
  # Add panel label inside plot
  if (!is.null(panel_label)) {
    label_y <- if (!is.null(ylim)) ylim[2] * 1.20 else y_max * 1.20
    p <- p + annotate("text", x = -Inf, y = label_y, label = panel_label,
                      fontface = "bold", size = 6, hjust = -0.2)
  }
  
  return(p)
}

ylim_16S <- c(0, 230)
ylim_ITS <- c(0, 65)
ylim_RPS10 <- c(0, 9)

plots_16S_mod <- list(
  modify_plot(plots_16S[["Treatuntreat_EP"]], FALSE, colors_treatuntreat, "Treatuntreat_EP",
              tukey_result = tukey_results_16S_nowf$InvSimpson_Treatuntreat_EP, ylim = ylim_16S, panel_label = "A"),
  modify_plot(plots_16S[["YrSinceTrt_BR"]], TRUE, colors_yrsince, "YrSinceTrt_BR",
              tukey_result = tukey_results_16S_nowf$InvSimpson_YrSinceTrt_BR, ylim = ylim_16S, panel_label = "B"),
  modify_plot(plots_16S[["Fire_Mod"]], TRUE, colors_fire, "Fire_Mod",
              tukey_result = tukey_results_16S$InvSimpson_Fire_Mod, ylim = ylim_16S, panel_label = "C")
)

plots_its_mod <- list(
  modify_plot(plots_its[["Treatuntreat_EP"]], FALSE, colors_treatuntreat, "Treatuntreat_EP",
              tukey_result = tukey_results_ITS_nowf$InvSimpson_Treatuntreat_EP, ylim = ylim_ITS, panel_label = "D"),
  modify_plot(plots_its[["YrSinceTrt_BR"]], TRUE, colors_yrsince, "YrSinceTrt_BR",
              tukey_result = tukey_results_ITS_nowf$InvSimpson_YrSinceTrt_BR, ylim = ylim_ITS, panel_label = "E"),
  modify_plot(plots_its[["Fire_Mod"]], TRUE, colors_fire, "Fire_Mod",
              tukey_result = tukey_results_ITS$InvSimpson_Fire_Mod, ylim = ylim_ITS, panel_label = "F")
)

plots_rps10_mod <- list(
  modify_plot(plots_rps10[["Treatuntreat_EP"]], FALSE, colors_treatuntreat, "Treatuntreat_EP",
              tukey_result = tukey_results_rps10_nowf$InvSimpson_Treatuntreat_EP, ylim = ylim_RPS10, panel_label = "G"),
  modify_plot(plots_rps10[["YrSinceTrt_BR"]], TRUE, colors_yrsince, "YrSinceTrt_BR",
              tukey_result = tukey_results_rps10_nowf$InvSimpson_YrSinceTrt_BR, ylim = ylim_RPS10, panel_label = "H"),
  modify_plot(plots_rps10[["Fire_Mod"]], TRUE, colors_fire, "Fire_Mod",
              tukey_result = tukey_results_rps10$InvSimpson_Fire_Mod, ylim = ylim_RPS10, panel_label = "I")
)

create_header <- function(text) {
  ggdraw() + draw_label(text, fontface = 'bold', x = 0.5, y = 0.5, hjust = 0.5, vjust = 0.5, size = 14)
}
header_1 <- create_header("SOD Treatment")
header_2 <- create_header("Years Since SOD Treatment")
header_3 <- create_header("Chetco Bar Fire")

header_row <- plot_grid(NULL, header_1, header_2, header_3, ncol = 4, rel_widths = c(0.05, 1, 1, 1))

row_16S <- plot_grid(
  plots_16S_mod[[1]], plots_16S_mod[[2]], plots_16S_mod[[3]],
  ncol = 3, rel_widths = c(1, 1, 1), align = "h", axis = "lr"
)

row_ITS <- plot_grid(
  plots_its_mod[[1]], plots_its_mod[[2]], plots_its_mod[[3]],
  ncol = 3, rel_widths = c(1, 1, 1), align = "h", axis = "lr"
)

row_RPS10 <- plot_grid(
  plots_rps10_mod[[1]], plots_rps10_mod[[2]], plots_rps10_mod[[3]],
  ncol = 3, rel_widths = c(1, 1, 1), align = "h", axis = "lr"
)

plot_panel <- plot_grid(
  header_row, row_16S, row_ITS, row_RPS10,
  nrow = 4, rel_heights = c(0.08, 1, 1, 1),
  align = "hv", axis = "tblr"
)

facetted_figure_final <- plot_grid(
  ggdraw() + draw_label("Inverse Simpson Index", angle = 90, fontface = 'bold', size = 14),
  plot_panel, ncol = 2, rel_widths = c(0.03, 1), align = "hv", axis = "tblr"
)

ggsave("Figures/Figure1_alpha_diversity_faceted_colorB.png", facetted_figure_final, width = 9, height = 9, dpi = 800)
ggsave("Figures/Figure1_alpha_diversity_faceted_colorB.pdf", facetted_figure_final, width = 9, height = 9, dpi = 800)
ggsave("Figures/Figure1_alpha_diversity_faceted_colorB.svg", facetted_figure_final, width = 9, height = 9, dpi = 800)

facetted_figure_final
```

### Let's proceed with ordination for the ITS dataset-NMDS

```{r, ordination plots its NMDS}
set.seed(1)
strat_rel_sod_its_nowf.ps<-transform_sample_counts(sod_5asv_its_nowf.ps, function(x) x / sum(x))

strat_rel_sod_its.ps<-transform_sample_counts(sod_5asv_its.ps, function(x) x / sum(x))

ps_NMDS_its = ordinate(strat_rel_sod_its_nowf.ps, 
                   method="NMDS", 
                   distance="bray")

plot_its_trt<- plot_ordination(strat_rel_sod_its_nowf.ps, 
                         ps_NMDS_its ,
                         color= "Treatuntreat_EP") +
  #ggtitle("NMDS ordination plot - relative abundance") +
  labs(color = "Untreated vs. Treated") + 
  theme(axis.text=element_text(size=14), axis.title=element_text(size=14), title =  element_text(size=13), legend.text = element_text(size = 18))

plot_ellipses_its_trt<-plot_its_trt +
   stat_ellipse(linetype = 2) +
  theme_bw()

plot_ellipses_its_trt



plot_its_yrtrt<- plot_ordination(strat_rel_sod_its_nowf.ps, 
                         ps_NMDS_its ,
                         color= "YrSinceTrt_BR") +
  #ggtitle("NMDS ordination plot - relative abundance") +
  labs(color = "Year Since Treatment") + 
  theme(axis.text=element_text(size=14), axis.title=element_text(size=14), title =  element_text(size=13), legend.text = element_text(size = 18))

plot_ellipses_its_yrtrt<-plot_its_yrtrt +
   stat_ellipse(linetype = 2) +
  theme_bw()

plot_ellipses_its_yrtrt

# Now we use dataset with wildfire samples
ps_NMDS_its_wf = ordinate(strat_rel_sod_its.ps, 
                   method="NMDS", 
                   distance="bray")

plot_its_fire<- plot_ordination(strat_rel_sod_its.ps, 
                         ps_NMDS_its_wf,
                         color= "Fire_Mod") +
  #ggtitle("NMDS ordination plot - relative abundance") +
  labs(color = "Chetco Bar Fire") + 
  theme(axis.text=element_text(size=14), axis.title=element_text(size=14), title =  element_text(size=13), legend.text = element_text(size = 18))

plot_ellipses_its_fire<-plot_its_fire +
   stat_ellipse(linetype =2) +
  theme_bw()

plot_ellipses_its_fire
```

### Let's proceed with ordination for the 16S dataset-NMDS

```{r, ordination plots 16s NMDS}
set.seed(1)
strat_rel_sod_16S_nowf.ps<-transform_sample_counts(sod_5asv_16S_nowf.ps, function(x) x / sum(x))

strat_rel_sod_16S.ps<-transform_sample_counts(sod_5asv_16S.ps, function(x) x / sum(x))


ps_NMDS_16S = ordinate(strat_rel_sod_16S_nowf.ps, 
                   method="NMDS", 
                   distance="bray")

plot_16S_trt<- plot_ordination(strat_rel_sod_16S_nowf.ps, 
                         ps_NMDS_16S,
                         color= "Treatuntreat_EP") +
  #ggtitle("NMDS ordination plot - relative abundance") +
  labs(color = "Treatment") + 
  theme(axis.text=element_text(size=14), axis.title=element_text(size=14), title =  element_text(size=13), legend.text = element_text(size = 18))

plot_ellipses_16S_trt<-plot_16S_trt +
   stat_ellipse(linetype = 2) +
  theme_bw()

plot_ellipses_16S_trt

plot_16S_yrtrt<- plot_ordination(strat_rel_sod_16S_nowf.ps, 
                         ps_NMDS_16S,
                         color= "YrSinceTrt_BR") +
  #ggtitle("NMDS ordination plot - relative abundance") +
  labs(color = "Year Since Treatment") + 
  theme(axis.text=element_text(size=14), axis.title=element_text(size=14), title =  element_text(size=13), legend.text = element_text(size = 18))

plot_ellipses_16S_yrtrt<-plot_16S_yrtrt +
   stat_ellipse(linetype = 2) +
  theme_bw()

plot_ellipses_16S_yrtrt

ps_NMDS_16S_wf = ordinate(strat_rel_sod_16S.ps, 
                   method="NMDS", 
                   distance="bray")

plot_16S_fire<- plot_ordination(strat_rel_sod_16S.ps, 
                         ps_NMDS_16S_wf,
                         color= "Fire_Mod") +
  #ggtitle("NMDS ordination plot - relative abundance") +
  labs(color = "Chetco Bar Fire") + 
  theme(axis.text=element_text(size=14), axis.title=element_text(size=14), title =  element_text(size=13), legend.text = element_text(size = 18))

plot_ellipses_16S_fire<-plot_16S_fire +
   stat_ellipse(linetype = 2) +
  theme_bw()

plot_ellipses_16S_fire
```

### Let's proceed with ordination for the RPS10 dataset-NMDS
```{r, ordination plots rps10 NMDS}
set.seed(1)
strat_rel_sod_rps10_nowf.ps<-transform_sample_counts(sod_5asv_rps10_nowf.ps, function(x) x / sum(x))

strat_rel_sod_rps10.ps<-transform_sample_counts(sod_5asv_rps10.ps, function(x) x / sum(x))


ps_NMDS_rps10 = ordinate(strat_rel_sod_rps10_nowf.ps, 
                   method="NMDS", 
                   distance="bray")

plot_rps10_trt<- plot_ordination(strat_rel_sod_rps10_nowf.ps, 
                         ps_NMDS_rps10,
                         color= "Treatuntreat_EP") +
  #ggtitle("NMDS ordination plot - relative abundance") +
  labs(color = "Treatment") + 
  theme(axis.text=element_text(size=14), axis.title=element_text(size=14), title =  element_text(size=13), legend.text = element_text(size = 18))

plot_ellipses_rps10_trt<-plot_rps10_trt +
   stat_ellipse(linetype = 2) +
  theme_bw()

plot_ellipses_rps10_trt

plot_rps10_yrtrt<- plot_ordination(strat_rel_sod_rps10_nowf.ps, 
                         ps_NMDS_rps10,
                         color= "YrSinceTrt_BR") +
  #ggtitle("NMDS ordination plot - relative abundance") +
  labs(color = "Year Since Treatment") + 
  theme(axis.text=element_text(size=14), axis.title=element_text(size=14), title =  element_text(size=13), legend.text = element_text(size = 18))

plot_ellipses_rps10_yrtrt<-plot_rps10_yrtrt +
   stat_ellipse(linetype = 2) +
  theme_bw()

plot_ellipses_rps10_yrtrt

ps_NMDS_rps10_wf = ordinate(strat_rel_sod_rps10.ps, 
                   method="NMDS", 
                   distance="bray")

plot_rps10_fire<- plot_ordination(strat_rel_sod_rps10.ps, 
                         ps_NMDS_rps10_wf,
                         color= "Fire_Mod") +
  #ggtitle("NMDS ordination plot - relative abundance") +
  labs(color = "Chetco Bar Fire") + 
  theme(axis.text=element_text(size=14), axis.title=element_text(size=14), title =  element_text(size=13), legend.text = element_text(size = 18))

plot_ellipses_rps10_fire<-plot_rps10_fire +
   stat_ellipse(linetype = 2) +
  theme_bw()

plot_ellipses_rps10_fire
```

### Let's combine plots into facetted plot

```{r, plots ordination pcoa}
### Let's combine plots into facetted plot
create_column <- function(ord_plot) {
  plot_grid(
    ord_plot,
    nrow = 1,
    align = 'v',
    axis = 'l',
    rel_heights = c(1)
  )
}

col_16S_treatment <- create_column(plot_ellipses_16S_trt)
col_16S_yrtrt <- create_column(plot_ellipses_16S_yrtrt)
col_16S_fire <- create_column(plot_ellipses_16S_fire)
col_its_treatment <- create_column(plot_ellipses_its_trt)
col_its_yrtrt <- create_column(plot_ellipses_its_yrtrt)
col_its_fire <- create_column(plot_ellipses_its_fire)
col_rps10_treatment <- create_column(plot_ellipses_rps10_trt)
col_rps10_yrtrt <- create_column(plot_ellipses_rps10_yrtrt)
col_rps10_fire <- create_column(plot_ellipses_rps10_fire)

create_header <- function(text) {
  ggplot() +
    geom_hline(yintercept = 0.8, color = "black", size = 0.5) +
    geom_text(aes(x = 0.02, y = 0.4, label = text), size = 5, fontface = "bold", hjust = 0) +
    theme_void() +
    theme(plot.margin = margin(5, 0, 0, 10, "pt")) +
    xlim(0, 1) + ylim(0, 1)
}

row_16S <- plot_grid(col_16S_treatment, col_16S_yrtrt, col_16S_fire,
                     ncol = 3, labels = "A.", label_size = 14, hjust = -0.5, vjust = 2)
row_its <- plot_grid(col_its_treatment, col_its_yrtrt, col_its_fire,
                     ncol = 3, labels = "B.", label_size = 14, hjust = -0.5, vjust = 2)
row_rps10 <- plot_grid(col_rps10_treatment, col_rps10_yrtrt, col_rps10_fire,
                       ncol = 3, labels = "C.", label_size = 14, hjust = -0.5, vjust = 2)

facetted_figure <- plot_grid(
  row_16S,
  row_its,
  row_rps10,
  nrow = 3,
  align = 'v',
  axis = 'l',
  rel_heights = c(1.1, 1, 1)  # Slightly more height for the first row
)

facetted_figure

# Save the figure
#ggsave("Figures/Figure2_beta_diversity_ordination_NMDS.svg", facetted_figure, width = 16, height = 12)
#ggsave("Figures/Figure2_beta_diversity_ordination_NMDS.pdf", facetted_figure, width = 16, height = 12)
```

### Revised figure
```{r revised figures}
colors_treatuntreat <- c(
  "SOD-UT" = "#2271B2",  # deep blue
  "SOD-T"  = "#e69f00"   # rusty orange
)

colors_yrsince <- c(
  "SOD-UT"   = "#2271B2", # deep blue (anchor)
  "1-2 PT"   = "#D55E00", # rusty orange
  "3-4 PT"   = "#FF54ED", # cyan
  "5-8 PT"   = "#00B19F", # green
  "10-12 PT" = "#9EFF37"  # yellow (swap to #E69F00 if you need darker)
)


colors_fire <- c(
  "SOD-UT"   = "#2271B2", # deep blue
  "SOD-T"    = "#e69f00", # rusty orange
  "Wildfire" = "#F8071D"  # pink/magenta
)

modify_ordination_plot <- function(plot, colors, remove_y_label = FALSE, remove_x_label = FALSE,
                                   show_legend = FALSE, xlim = NULL, ylim = NULL, panel_label = NULL, legend_nrow = 1) {
  p <- plot +
    scale_color_manual(values = colors) +
    scale_fill_manual(values = colors) +
    stat_ellipse(aes(fill = NULL), linetype = 2, size = 0.8) +
    theme_classic() +
    theme(
      panel.border = element_rect(color = "black", fill = NA, size = 0.5),
      axis.line = element_blank(),
      plot.margin = unit(c(0.2, 0.3, 0.2, 0.3), "lines"),
      axis.text.x = element_text(size = 12),
      axis.text.y = element_text(size = 12),
      axis.title.x = element_text(size = 12, face = "bold", margin = margin(t = 8)),
      axis.title.y = element_text(size = 12, face = "bold", margin = margin(r = 8)),
      legend.position = if(show_legend) "bottom" else "none",
      legend.title = element_blank(),
      legend.text = element_text(size = 9),
      legend.key.size = unit(0.4, "cm"),
      plot.title = element_blank(),
      strip.text = element_blank(),
      strip.background = element_blank()
    )

  if (remove_y_label) p <- p + theme(axis.title.y = element_blank())
  if (remove_x_label) p <- p + theme(axis.title.x = element_blank())

  if (!is.null(xlim) || !is.null(ylim)) {
    p <- p + coord_cartesian(xlim = xlim, ylim = ylim, clip = "off") +
             scale_x_continuous(expand = expansion(mult = c(0.02, 0.02))) +
             scale_y_continuous(expand = expansion(mult = c(0.02, 0.02)))
  }

  if (identical(xlim, c(-1, 1)) && identical(ylim, c(-1, 1))) {
    p <- p + scale_x_continuous(breaks = seq(-1, 1, 1), labels = seq(-1, 1, 1)) +
             scale_y_continuous(breaks = seq(-1, 1, 1), labels = seq(-1, 1, 1))
  }

  if (show_legend) {
    p <- p + guides(
      color = guide_legend(nrow = legend_nrow, byrow = TRUE),
      fill = guide_legend(nrow = legend_nrow, byrow = TRUE)
    ) +
    theme(
      legend.box = "horizontal",
      legend.spacing.x = unit(1, "pt"),
      legend.spacing.y = unit(0, "pt"),
      legend.margin = margin(t = 3, r = 0, b = 0, l = 0),
      legend.box.margin = margin(t = 0, r = 0, b = 0, l = 0)
    )
  }

  if (!is.null(panel_label)) {
    label_y <- if (!is.null(ylim)) ylim[2] * 0.98 else Inf
    p <- p + annotate("text", x = -Inf, y = label_y, label = panel_label,
                      fontface = "bold", size = 6, hjust = -0.2, vjust = 1)
  }

  return(p)
}

xlim_16S <- c(-2, 4); ylim_16S <- c(-3, 2)
xlim_ITS <- c(-1, 1); ylim_ITS <- c(-1, 1)
xlim_RPS10 <- c(-3, 3); ylim_RPS10 <- c(-3, 3)

plot_16S_trt_mod    <- modify_ordination_plot(plot_16S_trt, colors_treatuntreat, TRUE, TRUE, FALSE, xlim_16S, ylim_16S, panel_label = "A")
plot_16S_yrtrt_mod  <- modify_ordination_plot(plot_16S_yrtrt, colors_yrsince, TRUE, TRUE, FALSE, xlim_16S, ylim_16S, panel_label = "B")
plot_16S_fire_mod   <- modify_ordination_plot(plot_16S_fire, colors_fire, TRUE, TRUE, FALSE, xlim_16S, ylim_16S, panel_label = "C")

plot_its_trt_mod    <- modify_ordination_plot(plot_its_trt, colors_treatuntreat, TRUE, TRUE, FALSE, xlim_ITS, ylim_ITS, panel_label = "D")
plot_its_yrtrt_mod  <- modify_ordination_plot(plot_its_yrtrt, colors_yrsince, TRUE, TRUE, FALSE, xlim_ITS, ylim_ITS, panel_label = "E")
plot_its_fire_mod   <- modify_ordination_plot(plot_its_fire, colors_fire, TRUE, TRUE, FALSE, xlim_ITS, ylim_ITS, panel_label = "F")

plot_rps10_trt_mod  <- modify_ordination_plot(plot_rps10_trt, colors_treatuntreat, TRUE, TRUE, TRUE, xlim_RPS10, ylim_RPS10, panel_label = "G", legend_nrow = 1)
plot_rps10_yrtrt_mod<- modify_ordination_plot(plot_rps10_yrtrt, colors_yrsince, TRUE, TRUE, TRUE, xlim_RPS10, ylim_RPS10, panel_label = "H", legend_nrow = 2)
plot_rps10_fire_mod <- modify_ordination_plot(plot_rps10_fire, colors_fire, TRUE, TRUE, TRUE, xlim_RPS10, ylim_RPS10, panel_label = "I", legend_nrow = 1)

create_header <- function(text) {
  ggdraw() + draw_label(text, fontface = 'bold', x = 0.5, y = 0.5, hjust = 0.5, vjust = 0.5, size = 14)
}
header_1 <- create_header("SOD Treatment")
header_2 <- create_header("Years Since SOD Treatment")
header_3 <- create_header("Chetco Bar Fire")

header_row <- plot_grid(NULL, header_1, header_2, header_3, ncol = 4, rel_widths = c(0.05, 1, 1, 1))

row_16S   <- plot_grid(plot_16S_trt_mod, plot_16S_yrtrt_mod, plot_16S_fire_mod, ncol = 3, rel_widths = c(1,1,1), align = "h", axis = "tb")
row_ITS   <- plot_grid(plot_its_trt_mod, plot_its_yrtrt_mod, plot_its_fire_mod, ncol = 3, rel_widths = c(1,1,1), align = "h", axis = "tb")
row_RPS10 <- plot_grid(plot_rps10_trt_mod, plot_rps10_yrtrt_mod, plot_rps10_fire_mod, ncol = 3, rel_widths = c(1,1,1), align = "h", axis = "tb")

plot_panel <- plot_grid(header_row, row_16S, row_ITS, row_RPS10,
                        nrow = 4, rel_heights = c(0.08, 1, 1, 1.12),  # more space for 2-row legend
                        align = "v", axis = "lr")

y_axis_label <- ggdraw() + draw_label("NMDS2", angle = 90, fontface = 'bold', size = 14)
x_axis_label <- ggdraw() + draw_label("NMDS1", fontface = 'bold', size = 14)

plot_with_x_label <- plot_grid(plot_panel, x_axis_label, nrow = 2, rel_heights = c(1, 0.04))

facetted_figure_final <- plot_grid(y_axis_label, plot_with_x_label,
                                   ncol = 2, rel_widths = c(0.03, 1))

ggsave("Figures/Figure2_beta_diversity_NMDS_faceted_colorB.png", facetted_figure_final, width = 9, height = 9, dpi = 800)
ggsave("Figures/Figure2_beta_diversity_NMDS_faceted_colorB.pdf", facetted_figure_final, width = 9, height = 9, dpi = 800)
ggsave("Figures/Figure2_beta_diversity_NMDS_faceted_colorB.svg", facetted_figure_final, width = 9, height = 9, dpi = 800)

facetted_figure_final
```

### Software used

```{r software used}
sessioninfo::session_info()
```
