---
output:
  html_document: default
  pdf_document: default
---
---
title: "ITS exploratory analysis IDtaxa"
output: html_document
date: "2023-11-06"
---
```{r load packages, include=TRUE}
library(phyloseq); packageVersion("phyloseq")
library(ggplot2); packageVersion("ggplot2")
library(readr); packageVersion("readr")
library(purrr); packageVersion("purrr")
library(dplyr); packageVersion("dplyr")
library(metacoder); packageVersion("metacoder")
library(data.table); packageVersion("data.table")
```
#To do, save csv file of taxonomic assignments and ASVs to save space
#Load data
```{r load files, include=TRUE}
#Load DADA2 files
load("~/Documents/Postdoc/SOD_microbiome/Data/ITS/my_work_space_ITS_ee2.RData")
#taxa<- read.table("~/Documents/Postdoc/SOD_microbiome/Data/ITS/ITSee2_taxonomy_unite7_18_23_bootstraps_60.out",sep="#\t",stringsAsFactors=F,header=T)
taxa<- read.table("~/ITSee2_taxonomy_unite_july2023_decipher_idtaxa.out2",sep="\t",stringsAsFactors=F,header=T)
#Load Metadata file
samdf <- read.csv("~/Documents/Postdoc/SOD_microbiome/Data/ITS/Metadata_ITS.csv")
```

Exploratory analysis-track reads throughout DADA2 analysis
Track number of reads throughout the pipeline
```{r track reads through DADA2 analysis, include=TRUE}
# #MAXEE was 2. I could always make this even less stringent, since I am losing reads in the first step
# getN <- function(x) sum(dada2::getUniques(x))
# 
track <- cbind(out, sapply(dadaFs, getN), sapply(dadaRs, getN), sapply(mergers, getN), rowSums(seqtab.nochim))
# 
# # If processing a single s ample, remove the sapply calls: e.g. replace sapply(dadaFs, getN) with getN(dadaFs)
colnames(track) <- c("input", "filtered", "denoisedF", "denoisedR", "merged", "nonchim")
rownames(track) <- sample.names
track<-as.data.frame(track)
print(track)
write.csv(track, file="Results/ITS/track_readsITS.csv",row.names = TRUE)
#track <- read.csv(file = "Results/ITS/track_readsITS.csv", row.names = 1)
```

For ease of analysis, convert the DADA2 outputs into a Phyloseq object, and filter out any necessary samples
```{r Phyloseq object, include=TRUE}
all(rownames(seqtab.nochim) %in% samdf$Sample_nameITS) #This needs to be TRUE
rownames(samdf) <- samdf$Sample_nameITS
samdata.ps <- sample_data(samdf)
taxa.ps <- tax_table(as.matrix(taxa))

seqtab.nochim.ps <- otu_table(seqtab.nochim, taxa_are_rows=FALSE)
ps <- phyloseq(seqtab.nochim.ps, samdata.ps, taxa.ps)

#Add refseq slot
dna <- Biostrings::DNAStringSet(taxa_names(ps))
names(dna) <- taxa_names(ps)
ps <- merge_phyloseq(ps, dna)
taxa_names(ps) <- paste0("ASV", seq(ntaxa(ps)))
```

Decontamination step using decontam package-this was most necessary for 16S analysis. ITS control samples don't have reads, so I skipped this.
```{r Decontam, include=TRUE}
#Inspect library sizes
df <- as.data.frame(sample_data(ps)) # Put sample_data into a ggplot-friendly data.frame
df$LibrarySize <- sample_sums(ps)
df <- df[order(df$LibrarySize),]
df$Index <- seq(nrow(df))
ggplot(data=df, aes(x=Index, y=LibrarySize, color=Sample_or_Control)) + geom_point()

sample_data(ps)$is.neg <- sample_data(ps)$Sample_or_Control == "Control"
contamdf.prev <- isContaminant(ps, method="prevalence", neg="is.neg")
table(contamdf.prev$contaminant)
# head(which(contamdf.prev$contaminant))
#
contamdf.prev05 <- isContaminant(ps, method="prevalence", neg="is.neg", threshold=0.5)
table(contamdf.prev05$contaminant)
which(contamdf.prev05$contaminant)

ps.pa <- transform_sample_counts(ps, function(abund) 1*(abund>0))
ps.pa.neg <- prune_samples(sample_data(ps.pa)$Sample_or_Control == "Control", ps.pa)
ps.pa.pos <- prune_samples(sample_data(ps.pa)$Sample_or_Control == "Sample", ps.pa)

# Make data.frame of prevalence in positive and negative samples
df.pa <- data.frame(pa.pos=taxa_sums(ps.pa.pos), pa.neg=taxa_sums(ps.pa.neg),
                    contaminant=contamdf.prev05$contaminant)
ggplot(data=df.pa, aes(x=pa.neg, y=pa.pos, color=contaminant)) + geom_point() +
  xlab("Prevalence (Negative Controls)") + ylab("Prevalence (True Samples)")

#Remove contaminants
ps.decontam <- prune_taxa(!contamdf.prev05$contaminant, ps)

#Save for use with other analyses
save(ps.decontam, file = "Data/ITS/ITS_phyloseq.decontam.RData")
```

Take a first glance at the data. Helpful for deciding how to filter data. Prevelance is how many samples have the ASV, and total abundance is how many times ASV is present across all samples. I don't do any filtering until I split up samples into relevant experiment.
```{r A look at prevalence}
#Prevalence-number of samples in which a taxon appears at least once
#Total abundance
prevdf = apply(X = phyloseq::otu_table(ps.decontam),
               MARGIN = ifelse(taxa_are_rows(ps.decontam), yes = 1, no = 2),
               FUN = function(x){sum(x > 0)})

# Add taxonomy and total read counts to this data.frame
prevdf = data.frame(Prevalence = prevdf,
                    TotalAbundance = taxa_sums(ps.decontam),
                    tax_table(ps.decontam))
```

Now I will focus on 100 plot experiment and prepare dataset for further analysis by subsetting these samples from the rest of the sample data. I will also remove allow abundance ASVs not present in at least one sample and not present at least five times
```{r Subset dataset, include=TRUE}
#Experiment 1-100 plots base of tree
plot100exp.ps<-phyloseq::subset_samples(ps.decontam, hundredplot_tree_EP=="1")
#Prune ASVs not present in at least one sample
plot100exp.ps<- phyloseq::prune_taxa(taxa_sums(plot100exp.ps) > 0, plot100exp.ps)
save(plot100exp.ps, file="Data/ITS/plot100exp.RData")
#Prune ASVs not present in at least 5 times
exp1_5_filter.ps<- phyloseq::prune_taxa(taxa_sums(plot100exp.ps) > 5, plot100exp.ps)
save(exp1_5_filter.ps, file="Data/ITS/exp1_5_filter.RData")
```

```{r Filter datasets}
#Let's make an addtional dataset that is pruned to only include ASVs that are present in at least 5% of samples (and present at least five times) (Remove ASVs present in less than or equal to 5 samples)
exp1_5percent_filter.ps = phyloseq::filter_taxa(exp1_5_filter.ps, function(x) sum(x > 5) >= (0.05*length(x)), TRUE)

#Let's make an addtiional dataset that is pruned to only include ASVs that are present in at least 10% of samples (Remove ASVs present in less than or equal to 10 samples)
exp1_10percent_filter.ps = phyloseq::filter_taxa(exp1_5_filter.ps, function(x) sum(x > 5) >= (0.10*length(x)), TRUE)
```

```{r check prevalence dataframes after filterig}
#Double check that the filtering makes sense by making new prevalence dataframe
prevdf_check = apply(X = phyloseq::otu_table(exp1_5_filter.ps),
               MARGIN = ifelse(taxa_are_rows(exp1_5_filter.ps), yes = 1, no = 2),
               FUN = function(x){sum(x > 0)})

# Add taxonomy and total read counts to this data.frame
prevdf_check = data.frame(Prevalence = prevdf_check,
                    TotalAbundance = taxa_sums(exp1_5_filter.ps),
                    tax_table(exp1_5_filter.ps))

prevdf_afterfilt2 = apply(X = otu_table(exp1_5percent_filter.ps),
               MARGIN = ifelse(taxa_are_rows(exp1_5percent_filter.ps), yes = 1, no = 2),
               FUN = function(x){sum(x > 0)})

# Add taxonomy and total read counts to this data.frame
prevdf_afterfilt2 = data.frame(Prevalence = prevdf_afterfilt2,
                    TotalAbundance = taxa_sums(exp1_5percent_filter.ps),
                    tax_table(exp1_5percent_filter.ps))

prevdf_afterfilt3 = apply(X = otu_table(exp1_10percent_filter.ps),
               MARGIN = ifelse(taxa_are_rows(exp1_10percent_filter.ps), yes = 1, no = 2),
               FUN = function(x){sum(x > 0)})

# Add taxonomy and total read counts to this data.frame
prevdf_afterfilt3 = data.frame(Prevalence = prevdf_afterfilt3,
                    TotalAbundance = taxa_sums(exp1_10percent_filter.ps),
                    tax_table(exp1_10percent_filter.ps))
```

```{r subset tracked reads}
samples_to_subset <- samdf %>%
  filter(hundredplot_tree_EP == 1) %>% 
  pull(Sample_nameITS)

# Filter the track dataframe
subset_track_reads <- track[row.names(track) %in% samples_to_subset, ]
write.csv(track, file="Results/ITS/track_readsITS_100plot.csv",row.names = TRUE)
# Print summary of the filtered track dataframe
summary(subset_track_reads)
```


```{r prevalence plots organized by Genus}
prevdf1 = subset(prevdf_check, Phylum %in% get_taxa_unique(exp1_5_filter.ps, "Phylum"))
ggplot(prevdf1, aes(TotalAbundance, Prevalence / nsamples(exp1_5_filter.ps),color=Phylum)) +
  # Include a guess for parameter
  geom_hline(yintercept = 0.005, alpha = 0.5, linetype = 2) +  geom_point(size = 2, alpha = 0.7) +
  scale_x_log10() +  xlab("Total Abundance") + ylab("Prevalence [Frac. Samples]") +
  facet_wrap(~Phylum) + theme(legend.position="none")
```

```{r Reads sequenced by experiment}
#Before filtering reads
sdt = data.table::data.table(as(sample_data(plot100exp.ps), "data.frame"),
                 TotalReads = sample_sums(plot100exp.ps), keep.rownames = TRUE)
ps2eqDepth = ggplot(sdt, aes(TotalReads)) + geom_histogram(color="black", fill="lightblue") + ggtitle("Histogram of sequencing depth-before filtering samples with low read depths") + labs(x="Depth", y="Frequency")
ps2eqDepth

#After filtering reads
sdt_filt = data.table::data.table(as(sample_data(exp1_5_filter.ps), "data.frame"),
                 TotalReads = sample_sums(exp1_5_filter.ps), keep.rownames = TRUE)
ps2eqDepth_filt = ggplot(sdt_filt, aes(TotalReads)) + geom_histogram(color="black", fill="lightblue") + ggtitle("Histogram of sequencing depth after filtering samples with low read depths") + labs(x="Depth", y="Frequency")
ps2eqDepth_filt
```

Bar plot conglomerated by Genus-using filtered dataset and facet wrapped by treatment
```{r Barplot Phyloseq barplot}
#The function "transform_sample_counts()" will perform a given function across all taxa counts
#We will use it to convert the counts to percentages
ps_transform <- transform_sample_counts(exp1_5_filter.ps,  function(x) x / sum(x))
pdf("Figures/ITS/RelAbundance_Plot_Treatment.pdf", width = 40, height = 30)
plot_bar(ps_transform, fill="Phylum", title = "Proportional abundance plot-ASVs color-coded at Phylum level and faceted by treatment") + facet_wrap(~treatuntreat_EP, scales="free_x", nrow=1)
dev.off()
```


```{r barplot-Relative abundance of taxa vs. sample-not agglomerated}
dataITS <- exp1_5_filter.ps %>%
  tax_glom(taxrank = "Genus") %>%                     # agglomerate at Genus level
  transform_sample_counts(function(x) {x/sum(x)} ) %>% # Transform to rel. abundance
  psmelt() %>%                                         # Melt to long format
  filter(Abundance > 0.02) %>%                       # Filter out low abundance taxa
  arrange(Family)

dataITS<- data.frame(lapply(dataITS, function(x) ifelse(is.na(x), "Unknown", x)))

# Plot 
abund_plot <- ggplot(dataITS, aes(x = Sample, y = Abundance, fill = Family)) + 
  geom_bar(stat = "identity", position = "stack", color = "black", size = 0.2) +
  scale_fill_viridis_d() +
  scale_color_viridis_d() +
  theme_minimal() +
  labs(
    y = "Relative Abundance",
    title = "Relative abundance of taxa by sample",
    fill = "Family"
  ) +
  facet_grid(~factor(treatuntreat_EP, labels = c("Untreated", "Treated")), scales = "free_x") +
  theme(
    axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5, size = 14),
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    legend.position = "top",
    legend.text = element_text(size = 20),
    legend.title = element_text(size = 20),  # Adjust legend title size
    legend.key.size = unit(2, "lines"),  # Adjust the size of the legend color key
    strip.text = element_text(size = 20),
    strip.background = element_blank()
  ) +
  guides(
    fill = guide_legend(
      reverse = TRUE,
      keywidth = 1,
      keyheight = 1,
      title.position = "top",
      title.hjust = 0.5,  # Center the legend title
      label.theme = element_text(size = 18)  # Adjust the size of the legend labels
    )
  )

ggsave("Figures/ITS/RelAbundance_Plot_Treatment2_0.02abundFamily.pdf", abund_plot, width = 30, height = 20)
```

Let's now visualize similar relative abundance plots as heat trees for alternative figures
First convert to taxmap object
```{r Convert Phyloseq object to taxmap object, include=FALSE}
obj<-parse_phyloseq(exp1_5_filter.ps)

obj$data$tax_abund <- calc_taxon_abund(obj, data = "otu_table", cols = obj$data$sample_data$Sample_nameITS)

obj$data$abund_prop <- calc_obs_props(obj, data = "otu_table", cols = obj$data$sample_data$Sample_nameITS, groups = rep("tax_prop", nrow(obj$data$sample_data)))

obj$data$tax_prop <- calc_taxon_abund(obj, data = "abund_prop", cols = "tax_prop")
obj$data$abund_prop <- NULL

# Calculate total abundance per sample
asv_total_abundance <- rowSums(obj$data$tax_abund[, -1])

obj$data$tax_data <- data.frame(lapply(obj$data$tax_data, function(x) ifelse(is.na(x), "Unknown", x)))
save(obj, file="Data/ITS/obj.RData")
```

```{r Heat tree steps}
set.seed(2)
obj %>%
  #added this filters because otherwise hard to decipher the tree
  #ilter_taxa(grepl(pattern = "^[a-zA-Z]+$", taxon_names)) %>% # remove "odd" taxa
  #filter_taxa(taxon_ranks == "s", supertaxa=TRUE) %>% # subset to the genus rank
  heat_tree(node_label = taxon_names,
          node_size = n_obs,
          node_color = n_obs,
          node_color_axis_label = "ASV count",
          node_size_axis_label = "Total Abundance of Taxa",
          layout = "da", initial_layout = "re",
          output_file = "Figures/ITS/plot_heattree_ITS.pdf")
```

Now I will make more trees that better shows differences between treated and untreated samples
Let's first look at taxa present in treated samples
```{r Heat tree treated samples}
#help me make the heat tree in below code chunk but just treated samples
obj$data$tax_abund <- calc_taxon_abund(obj, data = "otu_table", cols = obj$data$sample_data$Sample_nameITS)
obj$data$tax_occ <- calc_n_samples(obj, "tax_abund", groups = as.factor(obj$data$sample_data$treatuntreat_EP), cols = obj$data$sample_data$Sample_nameITS)

set.seed(4)
heat_tree(obj, 
          node_label = taxon_names,
          node_size = n_obs,
          node_color = obj$data$tax_occ$"1", 
          node_size_axis_label = "ASV count",
          node_color_axis_label = "Treated samples with reads",
          layout = "da", initial_layout = "re",
          output_file = "Figures/ITS/plot_heattree_treated_plots_ITS.pdf")
```

Let's look at taxa present in untreated samples
```{r Heat tree untreated samples}
set.seed(4)
heat_tree(obj,
          node_label = taxon_names,
          node_size = n_obs,
          node_color = obj$data$tax_occ$"0",
          node_size_axis_label = "ASV count",
          node_color_axis_label = "Unreated samples with reads",
          layout = "da", initial_layout = "re",
          output_file = "Figures/ITS/plot_heattree_untreated_plots_ITS.pdf")

```


Let's look at taxon abundance in treated samples
We first need to add more information to our taxmap object
```{r Heat tree untreated samples relative abundance}
obj$data$otu_props <- calc_obs_props(obj, "otu_table", cols = obj$data$sample_data$Sample_nameITS)

obj$data$tax_abund_prop <- calc_taxon_abund(obj, "otu_props")

obj$data$type_abund <- calc_group_mean(obj, "tax_abund_prop",
                                       cols =obj$data$sample_data$Sample_nameITS,
                                       groups = obj$data$sample_data$treatuntreat_EP)

print(obj$data$type_abund)
```

```{r Heat tree illustrating taxa in untreated samples}
set.seed(1)
obj %>%
  heat_tree(node_label = taxon_names,
            node_size = obj$data$type_abund$"0", 
            node_color = obj$data$type_abund$"0", 
            initial_layout = "re", layout = "da",
            node_color_axis_label = "Untreated samples",
            title = "Taxa in untreated samples",
            output_file = "Figures/ITS/plot_heattree_rel_abund_taxa_untreated_plots_ITS.pdf")
```

Let's look at taxon abundance in untreated samples
```{r Heat tree illustrating taxa in treated samples}
set.seed(1)
obj %>%
  heat_tree(node_label = taxon_names,
            node_size = obj$data$type_abund$"1", 
            node_color = obj$data$type_abund$"1", 
            layout = "da", initial_layout = "re", 
            node_color_axis_label = "Treated samples",
            title = "Taxa in treated samples",
            output_file = "Figures/ITS/plot_heattree_rel_abund_taxa_treated_plots_ITS.pdf")
```

Towards a core microbiome analysis
```{r Make additional taxmap objects 5 percent filter}
#Now make taxmap object for dataset where ASVs filtered if not present also in at leaset 5% of samples
obj_5<-parse_phyloseq(exp1_5percent_filter.ps)

obj_5$data$tax_abund <- calc_taxon_abund(obj_5, data = "otu_table", cols = obj_5$data$sample_data$Sample_nameITS)

obj_5$data$abund_prop <- calc_obs_props(obj_5, data = "otu_table", cols = obj_5$data$sample_data$Sample_nameITS, groups = rep("tax_prop", nrow(obj_5$data$sample_data)))

obj_5$data$tax_prop <- calc_taxon_abund(obj_5, data = "abund_prop", cols = "tax_prop")
obj_5$data$abund_prop <- NULL

# Calculate total abundance per sample
asv_total_abundance <- rowSums(obj_5$data$tax_abund[, -1])

obj_5$data$tax_data <- data.frame(lapply(obj_5$data$tax_data, function(x) ifelse(is.na(x), "Unknown", x)))

obj_5$data$otu_props <- calc_obs_props(obj_5, "otu_table", cols = obj_5$data$sample_data$Sample_nameITS)

obj_5$data$tax_abund_prop <- calc_taxon_abund(obj_5, "otu_props")

obj_5$data$type_abund <- calc_group_mean(obj_5, "tax_abund_prop",
                                       cols =obj_5$data$sample_data$Sample_nameITS,
                                       groups = obj_5$data$sample_data$treatuntreat_EP)

print(obj_5$data$type_abund)
```


What happens if we look at taxon abundance when we remove any taxa not present in at least 5 samples?
```{r Heat tree illustrating taxa in untreated samples 5 percent filter}
set.seed(1)
obj_5 %>%
  heat_tree(node_label = taxon_names,
            node_size = obj_5$data$type_abund$"0", 
            node_color = obj_5$data$type_abund$"0", 
            layout = "da", initial_layout = "re", 
            node_color_axis_label = "Untreated samples",
            title = "Taxa in untreated samples",
            output_file = "Figures/ITS/plot_heattree_rel_abund_taxa_untreated_plots_ITS_5perc.pdf")
```

Let's look at taxon abundance in untreated samples
```{r Heat tree illustrating taxa in treated samples 5 percent filter}
set.seed(1)
obj_5%>%
  heat_tree(node_label = taxon_names,
            node_size = obj_5$data$type_abund$"1", 
            node_color = obj_5$data$type_abund$"1", 
            layout = "da", initial_layout = "re", 
            node_color_axis_label = "Treated samples",
            title = "Taxa in treated samples",
            output_file = "Figures/ITS/plot_heattree_rel_abund_taxa_treated_plots_ITS_5perc.pdf")
```

```{r Make additional taxmap objects 10 percent filter}
#Now make taxmap object for dataset where ASVs filtered if not present also in at leaset 5% of samples
obj_10<-parse_phyloseq(exp1_10percent_filter.ps)

obj_10$data$tax_abund <- calc_taxon_abund(obj_10, data = "otu_table", cols = obj_10$data$sample_data$Sample_nameITS)

obj_10$data$abund_prop <- calc_obs_props(obj_10, data = "otu_table", cols = obj_10$data$sample_data$Sample_nameITS, groups = rep("tax_prop", nrow(obj_10$data$sample_data)))

obj_10$data$tax_prop <- calc_taxon_abund(obj_10, data = "abund_prop", cols = "tax_prop")
obj_10$data$abund_prop <- NULL

# Calculate total abundance per sample
asv_total_abundance <- rowSums(obj_10$data$tax_abund[, -1])

obj_10$data$tax_data <- data.frame(lapply(obj_10$data$tax_data, function(x) ifelse(is.na(x), "Unknown", x)))

obj_10$data$otu_props <- calc_obs_props(obj_10, "otu_table", cols = obj_10$data$sample_data$Sample_nameITS)

obj_10$data$tax_abund_prop <- calc_taxon_abund(obj_10, "otu_props")

obj_10$data$type_abund <- calc_group_mean(obj_10, "tax_abund_prop",
                                       cols =obj_10$data$sample_data$Sample_nameITS,
                                       groups = obj_10$data$sample_data$treatuntreat_EP)

print(obj_10$data$type_abund)
```

What happens if we look at taxon abundance when we remove any taxa not present in at least 10 samples?
```{r Heat tree illustrating taxa in untreated samples 10 percent filter}
set.seed(1)
obj_10 %>%
  heat_tree(node_label = taxon_names,
            node_size = obj_10$data$type_abund$"0", 
            node_color = obj_10$data$type_abund$"0", 
            layout = "da", initial_layout = "re", 
            node_color_axis_label = "Untreated samples",
            title = "Taxa in untreated samples",
            output_file = "Figures/ITS/plot_heattree_rel_abund_taxa_untreated_plots_ITS_10perc.pdf")
```

Let's look at taxon abundance in untreated samples
```{r Heat tree illustrating taxa in treated samples 10 percent filter}
set.seed(1)
obj_10 %>%
  heat_tree(node_label = taxon_names,
            node_size = obj_10$data$type_abund$"1", 
            node_color = obj_10$data$type_abund$"1", 
            layout = "da", initial_layout = "re", 
            node_color_axis_label = "Treated samples",
            title = "Taxa in treated samples",
            output_file = "Figures/ITS/plot_heattree_rel_abund_taxa_treated_plots_ITS_10perc.pdf")
```

Software used
```{r, software used}
sessionInfo()
```
