---
title: "01_ITS_exploratory_analysis"
output: html_document
date: "2023-12-01"
---

```{r setup, include=TRUE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(root.dir = "~/SOD_DADS_microbiome")

library(phyloseq); packageVersion("phyloseq")
library(ggplot2); packageVersion("ggplot2")
library(readr); packageVersion("readr")
library(purrr); packageVersion("purrr")
library(stringr); packageVersion("stringr")
library(dplyr); packageVersion("dplyr")
library(igraph); packageVersion("igraph")
library(metacoder); packageVersion("metacoder")
library(decontam); packageVersion("decontam")
library(data.table); packageVersion("data.table")
library(dada2); packageVersion("dada2")
```

### Exploratory analysis
Load data
```{r load files, include=TRUE}
seqtab.nochim <- read.table("Data/ITS/dada2_outputs/its_seqtab_nochim.out",sep="\t",stringsAsFactors=F,header=T)
rownames(seqtab.nochim) <- gsub("__", "_", rownames(seqtab.nochim))

taxa<- read.table("Data/ITS/dada2_outputs/its_taxa_unitefungi.out", sep="\t", stringsAsFactors=F,header=T)

taxa_filt <- taxa %>%
  mutate(across(starts_with("tax."), ~ ifelse(get(str_replace(cur_column(), "tax.", "boot.")) < 60, NA, .))) %>%
  set_names(str_replace(names(.), "tax.", ""))


#I want to rename tax input so more in line with other two datasets for final figures
remove_prefix <- function(data) {
  data <- gsub("[a-z]__", "", data)  
  return(data)
}

# Apply the function to each column in the table
taxa_filt[] <- lapply(taxa_filt, remove_prefix)

revise_species <- function(genus, species) {
  # Combine genus and species with an underscore if species is not NA
  revised_species <- ifelse(is.na(species), NA, paste(genus, species, sep = "_"))
  return(revised_species)
}

# Apply the function to revise Species column
taxa_filt$Species <- revise_species(taxa_filt$Genus, taxa_filt$Species)

samdf <- read.csv("Data/ITS/Metadata_ITS_short_mod.csv")
```

Exploratory analysis-track reads throughout DADA2 analysis
Track number of reads throughout the pipeline
```{r track reads through DADA2 analysis, include=TRUE}
track <- read.table("Data/ITS/dada2_outputs/its_trackreads.out", sep="\t", stringsAsFactors = F, header=T)
rownames(track) <- gsub("__", "_", rownames(track))
```

For ease of analysis, convert the DADA2 outputs into a Phyloseq object, and filter out any necessary samples
```{r Phyloseq object, include=TRUE}
all(rownames(seqtab.nochim) %in% samdf$Sample_nameITS)

rownames(samdf) <- samdf$Sample_nameITS

samdata.ps <- sample_data(samdf)

taxa.ps <- tax_table(as.matrix(taxa_filt))

seqtab.nochim.ps <- otu_table(seqtab.nochim, taxa_are_rows=FALSE)
ps <- phyloseq(seqtab.nochim.ps, samdata.ps, taxa.ps)

#Add refseq slot
dna <- Biostrings::DNAStringSet(taxa_names(ps))
names(dna) <- taxa_names(ps)
ps <- merge_phyloseq(ps, dna)
taxa_names(ps) <- paste0("ASV", seq(ntaxa(ps)))
```

Decontamination step using decontam package-this was most necessary for 16S analysis. ITS control samples don't have reads, so I skipped this.
```{r Decontam, include=TRUE}
#Inspect library sizes
df <- as.data.frame(sample_data(ps)) # Put sample_data into a ggplot-friendly data.frame
df$LibrarySize <- sample_sums(ps)
df <- df[order(df$LibrarySize),]
df$Index <- seq(nrow(df))
ggplot(data=df, aes(x=Index, y=LibrarySize, color=Sample_or_control)) + geom_point()

sample_data(ps)$is.neg <- sample_data(ps)$Sample_or_control == "control"
contamdf.prev <- isContaminant(ps, method="prevalence", neg="is.neg")
table(contamdf.prev$contaminant)
# head(which(contamdf.prev$contaminant))

contamdf.prev05 <- isContaminant(ps, method="prevalence", neg="is.neg", threshold=0.5)
table(contamdf.prev05$contaminant)
which(contamdf.prev05$contaminant)

ps.pa <- transform_sample_counts(ps, function(abund) 1*(abund>0))
ps.pa.neg <- prune_samples(sample_data(ps.pa)$Sample_or_control == "control", ps.pa)
ps.pa.pos <- prune_samples(sample_data(ps.pa)$Sample_or_control == "sample", ps.pa)

# Make data.frame of prevalence in positive and negative samples
df.pa <- data.frame(pa.pos=taxa_sums(ps.pa.pos), pa.neg=taxa_sums(ps.pa.neg),
                    contaminant=contamdf.prev05$contaminant)
ggplot(data=df.pa, aes(x=pa.neg, y=pa.pos, color=contaminant)) + geom_point() +
  xlab("Prevalence (Negative Controls)") + ylab("Prevalence (True Samples)")

#Remove contaminants
ps.decontam <- prune_taxa(!contamdf.prev05$contaminant, ps)

#Save for use with other analyses
save(ps.decontam, file = "Data/ITS/ITS_phyloseq.decontam.RData")
```

Take a first glance at the data. Helpful for deciding how to filter data. Prevelance is how many samples have the ASV, and total abundance is how many times ASV is present across all samples. I don't do any filtering until I split up samples into relevant experiment.
```{r A look at prevalence}
#Prevalence-number of samples in which a taxon appears at least once
#Total abundance
prevdf = apply(X = phyloseq::otu_table(ps.decontam),
               MARGIN = ifelse(taxa_are_rows(ps.decontam), yes = 1, no = 2),
               FUN = function(x){sum(x > 0)})

# Add taxonomy and total read counts to this data.frame
prevdf = data.frame(Prevalence = prevdf,
                    TotalAbundance = taxa_sums(ps.decontam),
                    tax_table(ps.decontam))
```

Now I will focus on 100 plot experiment and prepare dataset for further analysis by subsetting these samples from the rest of the sample data. I will also remove allow abundance ASVs not present in at least one sample and not present at least five times
```{r Subset dataset, include=TRUE}
#Experiment 1-100 plots base of tree
plot100exp.ps<-phyloseq::subset_samples(ps.decontam, Hundredplot_tree_EP=="1")
#Prune ASVs not present in at least one sample
plot100exp.ps<- phyloseq::prune_taxa(taxa_sums(plot100exp.ps) > 0, plot100exp.ps)
save(plot100exp.ps, file="Data/ITS/plot100exp.RData")
#Prune ASVs not present in at least 5 times
exp1_5_filter.ps<- phyloseq::prune_taxa(taxa_sums(plot100exp.ps) > 5, plot100exp.ps)
exp1_5_filter.ps<- phyloseq::prune_taxa(taxa_sums(exp1_5_filter.ps) > 0, exp1_5_filter.ps)
exp1_5_filter.ps@sam_data$Treatuntreat_EP<-factor(exp1_5_filter.ps@sam_data$Treatuntreat_EP, levels = c(0, 1), labels = c("untreated", "treated"))

save(exp1_5_filter.ps, file="Data/ITS/exp1_5_filter.RData")

readsums<-sort(sample_sums(plot100exp.ps))
median(readsums)
summary(readsums)
```


```{r check prevalence dataframes after filterig}
#Double check that the filtering makes sense by making new prevalence dataframe
prevdf_check = apply(X = phyloseq::otu_table(exp1_5_filter.ps),
               MARGIN = ifelse(taxa_are_rows(exp1_5_filter.ps), yes = 1, no = 2),
               FUN = function(x){sum(x > 0)})

# Add taxonomy and total read counts to this data.frame
prevdf_check = data.frame(Prevalence = prevdf_check,
                    TotalAbundance = taxa_sums(exp1_5_filter.ps),
                    tax_table(exp1_5_filter.ps))

```

```{r subset tracked reads}
samples_to_subset <- samdf %>%
  filter(Hundredplot_tree_EP == 1) %>% 
  pull(Sample_nameITS)
# Filter the track dataframe

track$sample <- paste0(rownames(track), "_L001")
subset_track_reads <- track[track$sample %in% samples_to_subset, ]
write.csv(subset_track_reads, file="Results/ITS/track_readsITS_100plot.csv",row.names = TRUE)
# Print summary of the filtered track dataframe
summary(subset_track_reads)
```

```{r prevalence plots organized by phylum}
prevdf1 = subset(prevdf_check, Phylum %in% get_taxa_unique(exp1_5_filter.ps, "Phylum"))
ggplot(prevdf1, aes(TotalAbundance, Prevalence / nsamples(exp1_5_filter.ps),color=Phylum)) +
  # Include a guess for parameter
  geom_hline(yintercept = 0.005, alpha = 0.5, linetype = 2) +  geom_point(size = 2, alpha = 0.7) +
  scale_x_log10() +  xlab("Total Abundance") + ylab("Prevalence [Frac. Samples]") +
  facet_wrap(~Phylum) + theme(legend.position="none")
```

```{r Reads sequenced by experiment}
#Before filtering reads
sdt = data.table::data.table(as(sample_data(plot100exp.ps), "data.frame"),
                 TotalReads = sample_sums(plot100exp.ps), keep.rownames = TRUE)
ps2eqDepth = ggplot(sdt, aes(TotalReads)) + geom_histogram(color="black", fill="lightblue") + ggtitle("Histogram of sequencing depth-before filtering samples with low read depths") + labs(x="Depth", y="Frequency")
ps2eqDepth

#After filtering reads
sdt_filt = data.table::data.table(as(sample_data(exp1_5_filter.ps), "data.frame"),
                 TotalReads = sample_sums(exp1_5_filter.ps), keep.rownames = TRUE)
ps2eqDepth_filt = ggplot(sdt_filt, aes(TotalReads)) + geom_histogram(color="black", fill="lightblue") + ggtitle("Histogram of sequencing depth after filtering samples with low read depths") + labs(x="Depth", y="Frequency")
ps2eqDepth_filt
```

Bar plot conglomerated by Genus-using filtered dataset and facet wrapped by treatment
```{r Barplot Phyloseq barplot}
#The function "transform_sample_counts()" will perform a given function across all taxa counts
#We will use it to convert the counts to percentages
ps_transform <- transform_sample_counts(exp1_5_filter.ps,  function(x) x / sum(x))
pdf("Figures/ITS/RelAbundance_Plot_Treatment.pdf", width = 40, height = 30)
plot_bar(ps_transform, fill="Phylum", title = "Proportional abundance plot-ASVs color-coded at Phylum level and faceted by treatment") + facet_wrap(~Treatuntreat_EP, scales="free_x", nrow=1)
dev.off()
```


```{r barplot-Relative abundance of taxa vs. sample}
#load("Data/ITS/ee2_boot60/exp1_5_filter.RData")
dataITS <- exp1_5_filter.ps %>%
  tax_glom(taxrank = "Genus") %>%                     # agglomerate at Genus level
  transform_sample_counts(function(x) {x/sum(x)} ) %>% # Transform to rel. abundance
  psmelt() %>%                                         # Melt to long format
  filter(Abundance > 0.05) %>%                       # Filter out low abundance taxa
  arrange(Genus)

#dataITS<- data.frame(lapply(dataITS, function(x) ifelse(is.na(x), "Unclassified", x)))

# Plot 
abund_plot <- ggplot(dataITS, aes(x = Sample, y = Abundance, fill = Genus)) + 
  geom_bar(stat = "identity", position = "stack", color = "black", size = 0.2) +
  scale_fill_viridis_d() +
  scale_color_viridis_d() +
  theme_minimal() +
  labs(
    y = "Relative Abundance",
    title = "Relative abundance of taxa by sample",
    fill = "Genus"
  ) +
  facet_grid(~factor(Treatuntreat_EP, labels = c("Untreated", "Treated")), scales = "free_x") +
  theme(
    axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5, size = 14),
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    legend.position = "bottom",
    legend.text = element_text(size = 20),
    legend.title = element_text(size = 20),  # Adjust legend title size
    legend.key.size = unit(2, "lines"),  # Adjust the size of the legend color key
    strip.text = element_text(size = 20),
    strip.background = element_blank()
  ) +
  guides(
    fill = guide_legend(
      keywidth = 1,
      keyheight = 1,
      title.position = "top",
      title.hjust = 0.5,  # Center the legend title
      label.theme = element_text(size = 18)  # Adjust the size of the legend labels
    )
  )

ggsave("Figures/ITS/RelAbundance_Plot_Treatment_abund_Genus_agglom0.05.pdf", abund_plot, width = 45, height = 45)
```

We are also curious about top taxa 
```{r Top taxa}
new_names <- sapply(strsplit(plot100exp.ps@sam_data$Sample_nameITS, "-"), `[`, 1)

# Rename the Sample_nameITS column
plot100exp.ps@sam_data$Sample_nameITS <- new_names

plot100exp.ps@sam_data$Treatuntreat_EP <- factor(plot100exp.ps@sam_data$Treatuntreat_EP, labels = c("untreated", "treated"))

top20 <- names(sort(taxa_sums(plot100exp.ps), decreasing=TRUE))[1:20]
ps.top20 <- transform_sample_counts(plot100exp.ps, function(OTU) OTU/sum(OTU))
ps.top20<- tax_glom(ps.top20, taxrank = "Species") 
ps.top20 <- prune_taxa(top20, ps.top20)

pdf("Figures/ITS/ee5_top20_species_boot60.pdf", width = 34, height = 20)
plot_bar(ps.top20, x="Sample_nameITS", fill="Species") + facet_wrap(~Treatuntreat_EP, scales="free_x")
dev.off()

plot_bar(ps.top20, x="Sample_nameITS", fill="Species") + facet_wrap(~Treatuntreat_EP, scales="free_x")

data <-
  plot100exp.ps %>%
  tax_glom("Species") %>%
  psmelt() %>%
  as_tibble()

data1 <-
  plot100exp.ps %>%
  tax_glom("Species") %>%
  transform_sample_counts(function(x) {x/sum(x)} ) %>%
  psmelt() %>%
  as_tibble()

data_summed1 <- data1 %>%
  filter(Abundance > 0) %>%
  group_by(Treatuntreat_EP, Kingdom, Phylum, Class, Order, Family, Genus, Species) %>%
  summarise(Abundance = sum(Abundance),
            Prevalence = n_distinct(Sample)) 

write.csv(data_summed1, file="Results/ITS/mostabundanttaxa_relativeabundance_untreatedtreated_boot60.csv",row.names = TRUE)

data2 <-
  plot100exp.ps %>%
  tax_glom("Species") %>%
  psmelt() %>%
  as_tibble()

# Calculate percentages
data_summed2 <- data2 %>%
  filter(Abundance > 0) %>%
  group_by(Treatuntreat_EP, Kingdom, Phylum, Class, Order, Family, Genus, Species) %>%
  summarise(
    Abundance = sum(Abundance),
    Prevalence = n_distinct(Sample)
  ) %>%
  group_by(Treatuntreat_EP) %>%
  mutate(
    Total_Abundance = sum(Abundance),
    Percent_Abundance = (Abundance / Total_Abundance) * 100
  )

write.csv(data_summed2, file = "Results/ITS/mostabundanttaxa_untreatedtreated_boot60.csv", row.names = TRUE)

data3 <-
  plot100exp.ps %>%
  tax_glom("Species") %>%
  psmelt() %>%
  as_tibble()

# Calculate percentages
data_summed3 <- data3 %>%
  filter(Abundance > 0) %>%
  group_by(Kingdom, Phylum, Class, Order, Family, Genus, Species) %>%
  summarise(
    Abundance = sum(Abundance),
    Total_Abundance = sum(data3$Abundance),
    Prevalence = n_distinct(Sample),
    Percent_Abundance = (sum(Abundance) / sum(data3$Abundance)) * 100
  )
write.csv(data_summed3, file="Results/ITS/mostabundanttaxa_boot60.csv",row.names = TRUE)
```

```{r Map exploration}
new_names <- sapply(strsplit(plot100exp.ps@sam_data$Sample_nameITS, "-"), `[`, 1)

# Rename the Sample_nameITS column so sample names are shorter
plot100exp.ps@sam_data$Sample_nameITS <- new_names

plot100exp.ps@sam_data$Treatuntreat_EP <- factor(plot100exp.ps@sam_data$Treatuntreat_EP, labels = c("untreated", "treated"))

pdf("Figures/ITS/location_of_plots.pdf", width = 16, height = 16)

map_plot <- ggplot(plot100exp.ps@sam_data, aes(x = Longitude, y = Latitude, label = Sample_nameITS, color = Treatuntreat_EP)) +
  geom_point(size = 5, alpha = 0.5) +  # Add transparency to points to visualize overlapping points
  geom_text(size = 4, position = position_jitter(width = 0.025, height = 0.025), check_overlap = TRUE) +  # Jitter labels
  labs(x = "Longitude", y = "Latitude", title = "Sampling Locations") +
  scale_color_manual(values = c("untreated" = "blue", "treated" = "red")) +  # Set colors for treated and untreated
  theme_minimal() +  # Remove gray background and gridlines
  theme(  # Adjust title and axis label size, and change axis marks color
    plot.title = element_text(size = 20), 
    axis.title = element_text(size = 16),
    axis.text = element_text(color = "black")
  )

print(map_plot)
dev.off()

print(map_plot)
```

Let's now visualize similar relative abundance plots as heat trees for alternative figures
First convert to taxmap object
```{r Convert Phyloseq object to taxmap object, include=FALSE}
obj<-parse_phyloseq(exp1_5_filter.ps)

obj$data$tax_abund <- calc_taxon_abund(obj, data = "otu_table", cols = obj$data$sample_data$Sample_nameITS)

obj$data$abund_prop <- calc_obs_props(obj, data = "otu_table", cols = obj$data$sample_data$Sample_nameITS, groups = rep("tax_prop", nrow(obj$data$sample_data)))

obj$data$tax_prop <- calc_taxon_abund(obj, data = "abund_prop", cols = "tax_prop")
obj$data$abund_prop <- NULL

# Calculate total abundance per sample
asv_total_abundance <- rowSums(obj$data$tax_abund[, -1])

obj$data$tax_data <- data.frame(lapply(obj$data$tax_data, function(x) ifelse(is.na(x), "Unclassified", x)))
save(obj, file="Data/ITS/obj.RData")
```

Software used
```{r, software used}
sessionInfo()
```
