---
title: "Differential abundance analysis ITS"
output: html_document
date: "2024-02-20"
---
```{r setup, include=FALSE}
library(gt)
library(magrittr)
library(data.table)
library(phyloseq)
library(DESeq2)
library(cplm)
library(ggplot2)
library(cowplot)
theme_set(theme_cowplot())
library(ggbeeswarm)
library(doParallel)
library(foreach)
library(metacoder)

knitr::opts_knit$set(root.dir = "~/Documents/Postdoc/SOD_microbiome")
inDir <- "." # may need to change
```

```{r ad-hoc-functions}
load("Data/ITS/exp1_5_filter.RData")

sample.data.table <- function(ps) {
  as(sample_data(ps), "data.frame") %>%
    as.data.table(keep.rownames = "Sample") %>% 
    setkey(Sample) %>%
    return()
}
asv.data.table <- function(ps) {
  if (taxa_are_rows(ps)) {
    otu.mat <- t(as(otu_table(ps), "matrix"))
  } else {
    otu.mat <- as(otu_table(ps), "matrix")
  }
  return(as.data.table(otu.mat, keep.rownames = "Sample") %>% setkey(Sample))
}
cat.n <- function(x) cat(x, sep = "\n")
```

Below, we import the data we'll be using for the rest of the practical, and put it into a format that makes working with it easier. This section shouldn't need to be modified.
```{r data-import-and-prep}
exp1_5.data <- sample.data.table(exp1_5_filter.ps)
exp1_5.asvs <- asv.data.table(exp1_5_filter.ps)
exp1_5.asv.ids <- taxa_names(exp1_5_filter.ps)
exp1_5.dt <- exp1_5.data[exp1_5.asvs]
# Assuming you have the factor as "Untreated" and "Treated" currently

#exp1_5_filter.ps@sam_data$Treatuntreat_EP <- factor(exp1_5_filter.ps@sam_data$Treatuntreat_EP, levels = c(0, 1), labels = c("untreated", "treated"))

exp1_5_filter.ps@sam_data$Fire_EP <- factor(exp1_5_filter.ps@sam_data$Fire_EP, levels = c("Unburned", "SODTrt", "Wildfire"))
```


For differential abundance analyses of treated vs. untreated plots: we will serially conduct Wilcoxon tests on the exp1_5 data set. This test is the non-parametric (distribution-free) version of the Student's *t*-test and will tell us if the abundances of each taxon different significantly between our two groups: Unexposed vs Exposed. After running the tests, we correct for multiple tests by adjusting the p-values. There are two examples of methods for adjusting the p-values provided.
```{r wilcoxon-tests}
wilcox.tests <- lapply(exp1_5.asv.ids, function(asv) {
  frm <- paste(asv, "~ Treatuntreat_EP") # this combined with the as.formula() function below makes it possible to generate formulas in R programmatically
  return(
    wilcox.test(as.formula(frm), data = exp1_5.dt)
  )
})
exp1_5.raw.pvals <- sapply(wilcox.tests, function(mod) {return(mod$p.value)})

exp1_5.bon.pvals <- p.adjust(exp1_5.raw.pvals, method = "bonferroni")
# Printing the number of associations with adjusted p-values below our 0.05 cutoff. This value can be adjusted depending on your data and expectations
paste("Significant associations:", length(exp1_5.bon.pvals[exp1_5.bon.pvals <= 0.05])) %>%
  cat.n()

exp1_5.fdr.pvals <- p.adjust(exp1_5.raw.pvals, method = "fdr")
paste("Significant associations:", length(exp1_5.fdr.pvals[exp1_5.fdr.pvals <= 0.05])) %>%
  cat.n()
```

We will now plot the results
```{r plot }
exp1_5.melt.dt <- melt(
  exp1_5.dt, 
  measure.vars = exp1_5.asv.ids, 
  variable.name = "ASV", 
  value.name = "Abundance"
)
sig.asv <- exp1_5.asv.ids[exp1_5.bon.pvals <= 0.05]
exp1_5.plot.dt <- exp1_5.melt.dt[ASV %in% sig.asv]
exp1_5.plot.dt[, Treatuntreat_EP := factor(Treatuntreat_EP, levels = c("untreated", "treated"))] # the default is for levels to be in alphanumeric order, this way the control group, Unexposed, will be first along the x-axes

exp1_5.plot <- ggplot(exp1_5.plot.dt, aes(x = Treatuntreat_EP, y = Abundance)) + 
  geom_beeswarm(alpha = 0.6) + 
  stat_summary(
    fun.data = "mean_cl_boot", # This summary statistical calculates the 95% C.I. of the mean via non-parametric methods
    color = "red", 
    geom = "errorbar", 
    width = 0.8

  ) + 
  facet_wrap(~ ASV, scales = "free_y") +
  theme(axis.text.x = element_text(angle = 30, hjust = 1))

#Cannot plot as now sig associations
#exp1_5.plot 
```

Now we will look at an example where we have more than two factors and use kruskal wallis test
```{r kw test}
kw.tests <- lapply(exp1_5.asv.ids, function(asv) {
  frm <- paste(asv, "~ Fire_EP")
  return(
    kruskal.test(as.formula(frm), data = exp1_5.dt)
  )
})

exp1_5.raw.pvals <- sapply(kw.tests, function(mod) {return(mod$p.value)})

exp1_5.bon.pvals <- p.adjust(exp1_5.raw.pvals, method = "bonferroni")
paste(
  "Significant associations:", 
  length(exp1_5.bon.pvals[exp1_5.bon.pvals <= 0.05])
) %>% 
  cat.n()


exp1_5.fdr.pvals <- p.adjust(exp1_5.raw.pvals, method = "fdr")
paste(
  "Significant associations:", 
  length(exp1_5.fdr.pvals[exp1_5.fdr.pvals <= 0.05])
) %>% 
  cat.n()
```

```{r plotting-kruskal-wallis-tests}
exp1_5.melt.dt <- melt(
  exp1_5.dt, 
  measure.vars = exp1_5.asv.ids, 
  variable.name = "ASV", 
  value.name = "Abundance"
)
sig.asvs <- exp1_5.asv.ids[exp1_5.bon.pvals <= 0.05] # or exp1_5.fdr.pvals, Their choice
exp1_5.plot.dt <- exp1_5.melt.dt[ASV %in% sig.asvs]

wrs<-ggplot(exp1_5.plot.dt, aes(x = Fire_EP, y = Abundance)) + 
  geom_beeswarm(alpha = 0.6) + 
  stat_summary(
    fun.data = "mean_cl_boot", 
    color = "red", 
    geom = "errorbar", 
    width = 0.8
  ) + 
  facet_wrap(~ ASV, scales = "free_y")
ggsave("Figures/ITS/wrs_plots_fire.pdf", wrs, width = 15, height = 17)
```


```{r deseq2-treatuntreat}
exp1_5.deseq <- phyloseq_to_deseq2(exp1_5_filter.ps, ~ Treatuntreat_EP) # this function does all the heavy lifting for us to turn a phyloseq object into a DESeq2 object

#This was required so DESEQ doesn't error out (there are several approaches, but the deseq developer recommended this most recently, if going from phyloseq to deseq option)
exp1_5.dds <- estimateSizeFactors(exp1_5.deseq, type = 'poscounts')
exp1_5.dds = DESeq(exp1_5.dds)
exp1_5.res.dt <- results(exp1_5.dds) %>% as.data.table(keep.rownames = "ASV")

exp1_5.sig.dt <- exp1_5.res.dt[padj <= 0.05]

exp1_5.sig.dt = cbind(as(exp1_5.sig.dt, "data.frame"), as(tax_table(exp1_5_filter.ps)[exp1_5.sig.dt$ASV, ], "matrix"))
write.csv(exp1_5.sig.dt, file="Results/ITS/deseq_treatuntreat.csv",row.names = TRUE)

deseqplot2 <- ggplot(exp1_5.sig.dt,  aes(y = reorder(Species, log2FoldChange), x = log2FoldChange, fill=Family)) +
  geom_vline(xintercept = 0.0, color = "gray", size = 0.5) +
  geom_bar(stat = "identity", position = "dodge", width = 0.8) +  
  labs(x = "ASV", y = "Log2 Fold Change", title = "Treated vs Untreated Plots") + 
  theme(axis.text.x = element_text(angle = -90, hjust = 0, vjust = 0.5))
ggsave("Figures/ITS/deseq_trtuntrt_barplot.pdf", deseqplot2, width = 10, height = 12)

deseqplot2
```

Heat tree results have some slight differences from above, due in part to how the ASV abundance data is input into DESEQ2. I tried to make a heat tree using same DESEQ results as above but need to return to this as it was becoming time consuming. For now, please note the slight discrepancies in the results.
```{r deseq2-fire}
exp1_5_filter.ps@sam_data$Fire_EP <- factor(exp1_5_filter.ps@sam_data$Fire_EP, levels = c("Unburned", "SODTrt", "Wildfire"))

exp1_5.deseq <- phyloseq_to_deseq2(exp1_5_filter.ps, ~ Fire_EP)

exp1_5.dds <- estimateSizeFactors(exp1_5.deseq, type = 'poscounts')
exp1_5.dds = DESeq(exp1_5.dds, test="Wald", fitType="parametric")

resultsNames(exp1_5.dds) # the second and thirds names here indicate the comparisons we want to make below

exp1_5.SvU.res.dt <- as.data.table(
  results(exp1_5.dds, name = "Fire_EP_SODTrt_vs_Unburned"), 
  keep.rownames = "ASV"
)

# instead of typing out the full name yourself, you could point to the correct name programmatically by saving the output from resultsNames(exp1_5.dds) and using that in the results() function. If you're up for it, go ahead and do so.

exp1_5.SvU.sig.dt <- exp1_5.SvU.res.dt[padj <= 0.05]
#gt(exp1_5.SvU.sig.dt[order(log2FoldChange)]) %>% 
  #tab_header(title = "Fire- SODTrt vs Unburned") %>%
  #fmt_number(columns = 2:ncol(exp1_5.SvU.sig.dt), decimals = 3)

exp1_5.SvU.sig.dt = cbind(as(exp1_5.SvU.sig.dt, "data.frame"), as(tax_table(exp1_5_filter.ps)[exp1_5.SvU.sig.dt$ASV, ], "matrix"))
write.csv(exp1_5.SvU.sig.dt, file="Results/ITS/deseq_fireSvU.csv",row.names = TRUE)

deseqplot3 <- ggplot(exp1_5.SvU.sig.dt, aes(y = reorder(Species, log2FoldChange), x = log2FoldChange, fill=Family)) +
  geom_vline(xintercept = 0.0, color = "gray", size = 0.5) +
  geom_bar(stat = "identity", position = "dodge", width = 0.8) +  # Use geom_bar for bar plot
  labs(x = "ASV", y = "Log2 Fold Change", title = "Fire-SODtrt vs Untreated") +
  theme(axis.text.x = element_text(angle = -90, hjust = 0, vjust = 0.5))
ggsave("Figures/ITS/deseq_fire1_barplot.pdf", deseqplot3, width = 10, height = 12)
deseqplot3 

exp1_5.WvU.res.dt <- as.data.table(
  results(exp1_5.dds, name = "Fire_EP_Wildfire_vs_Unburned"), 
  keep.rownames = "ASV"
)
exp1_5.WvU.sig.dt <- exp1_5.WvU.res.dt[padj <= 0.05]

#gt(exp1_5.WvU.sig.dt[order(log2FoldChange)]) %>% 
  #tab_header(title = "Wildfire vs Unburned") %>%
  #fmt_number(columns = 2:ncol(exp1_5.WvU.sig.dt), decimals = 3)

exp1_5.WvU.sig.dt = cbind(as(exp1_5.WvU.sig.dt, "data.frame"), as(tax_table(exp1_5_filter.ps)[exp1_5.WvU.sig.dt$ASV, ], "matrix"))
write.csv(exp1_5.WvU.sig.dt, file="Results/ITS/deseq_fireWvU.csv",row.names = TRUE)

deseqplot4 <- ggplot(exp1_5.WvU.sig.dt,aes(y = reorder(Species, log2FoldChange), x = log2FoldChange, fill=Family)) +
  geom_vline(xintercept = 0.0, color = "gray", size = 0.5) +
  geom_bar(stat = "identity", position = "dodge", width = 0.8) +  # Use geom_bar for bar plot
  labs(x = "ASV", y = "Log2 Fold Change", title = "Fire-Wildfire vs Untreated") +
  theme(axis.text.x = element_text(angle = -90, hjust = 0, vjust = 0.5))
ggsave("Figures/ITS/deseq_fire2_barplot.pdf", deseqplot4, width = 15, height = 15)
deseqplot4

plot_grid(deseqplot3, deseqplot4, ncol = 1)
```