---
title: "ITS SOD Preparation Steps"
output: html_document
date: "2024-06-11"
---

```{r setup}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(root.dir = "~/SOD_DADS_microbiome")

library(phyloseq); packageVersion("phyloseq")
library(ggplot2); packageVersion("ggplot2")
library(readr); packageVersion("readr")
library(purrr); packageVersion("purrr")
library(furrr); packageVersion("furrr")
library(dplyr); packageVersion("dplyr")
library(stringr); packageVersion("stringr")
library(metacoder); packageVersion("metacoder")
library(data.table); packageVersion("data.table")
library(decontam); packageVersion("decontam")
```

### Load data in order to make phyloseq and taxmap objects
```{r load files, include=TRUE}
seqtab.nochim <- read.table("Data/ITS/dada2_outputs/its_seqtab_nochim.out",sep="\t",stringsAsFactors=F,header=T)
rownames(seqtab.nochim) <- gsub("__", "_", rownames(seqtab.nochim))

taxa<- read.table("Data/ITS/dada2_outputs/its_taxa_07_2023.out", sep="\t", stringsAsFactors=F,header=T)

taxa_filt <- taxa %>%
  mutate(across(starts_with("tax."), ~ ifelse(get(str_replace(cur_column(), "tax.", "boot.")) < 60, NA, .))) %>%
  set_names(str_replace(names(.), "tax.", ""))


#I want to rename tax input so more in line with other two datasets for final figures
remove_prefix <- function(data) {
  data <- gsub("[a-z]__", "", data)  
  return(data)
}

# Apply the function to each column in the table
taxa_filt[] <- lapply(taxa_filt, remove_prefix)

revise_species <- function(genus, species) {
  # Combine genus and species with an underscore if species is not NA
  revised_species <- ifelse(is.na(species), NA, paste(genus, species, sep = "_"))
  return(revised_species)
}

# Apply the function to revise Species column
taxa_filt$Species <- revise_species(taxa_filt$Genus, taxa_filt$Species)

samdf <- read.csv("Data/ITS/metadata_dbs/Metadata_ITS_short_mod.csv")
```


### Track number of reads throughout the pipeline
```{r track reads through DADA2 analysis, include=TRUE}
track <- read.table("Data/ITS/dada2_outputs/its_trackreads.out", sep="\t", stringsAsFactors = F, header=T)
rownames(track) <- gsub("__", "_", rownames(track))
```

### For ease of analysis, convert the DADA2 outputs into a Phyloseq object, and filter out any necessary samples
```{r Phyloseq object, include=TRUE}
all(rownames(seqtab.nochim) %in% samdf$Sample_nameITS)

rownames(samdf) <- samdf$Sample_nameITS

samdata.ps <- sample_data(samdf)

taxa.ps <- tax_table(as.matrix(taxa_filt))

seqtab.nochim.ps <- otu_table(seqtab.nochim, taxa_are_rows=FALSE)
ps <- phyloseq(seqtab.nochim.ps, samdata.ps, taxa.ps)

#Add refseq slot
dna <- Biostrings::DNAStringSet(taxa_names(ps))
names(dna) <- taxa_names(ps)
ps <- merge_phyloseq(ps, dna)
taxa_names(ps) <- paste0("ASV", seq(ntaxa(ps)))
```

### Decontamination step using decontam package-this was most necessary for 16S analysis. ITS control samples don't have reads, so I skipped this.
```{r Decontam, include=TRUE}
#Inspect library sizes
df <- as.data.frame(sample_data(ps)) # Put sample_data into a ggplot-friendly data.frame
df$LibrarySize <- sample_sums(ps)
df <- df[order(df$LibrarySize),]
df$Index <- seq(nrow(df))
ggplot(data=df, aes(x=Index, y=LibrarySize, color=Sample_or_control)) + geom_point()

sample_data(ps)$is.neg <- sample_data(ps)$Sample_or_control == "control"
contamdf.prev <- isContaminant(ps, method="prevalence", neg="is.neg")
table(contamdf.prev$contaminant)
# head(which(contamdf.prev$contaminant))

contamdf.prev05 <- isContaminant(ps, method="prevalence", neg="is.neg", threshold=0.5)
table(contamdf.prev05$contaminant)
which(contamdf.prev05$contaminant)

ps.pa <- transform_sample_counts(ps, function(abund) 1*(abund>0))
ps.pa.neg <- prune_samples(sample_data(ps.pa)$Sample_or_control == "control", ps.pa)
ps.pa.pos <- prune_samples(sample_data(ps.pa)$Sample_or_control == "sample", ps.pa)

# Make data.frame of prevalence in positive and negative samples
df.pa <- data.frame(pa.pos=taxa_sums(ps.pa.pos), pa.neg=taxa_sums(ps.pa.neg),
                    contaminant=contamdf.prev05$contaminant)
ggplot(data=df.pa, aes(x=pa.neg, y=pa.pos, color=contaminant)) + geom_point() +
  xlab("Prevalence (Negative Controls)") + ylab("Prevalence (True Samples)")

#Remove contaminants
ps.decontam <- prune_taxa(!contamdf.prev05$contaminant, ps)

ps.decontam <- ps.decontam %>%
  subset_taxa(
    Kingdom == "Fungi" 
  )

#Save for use with other analyses
save(ps.decontam, file = "Data/ITS/filtered_data/ITS_phyloseq.decontam.RData")
```

###Take a first glance at the data

Helpful for deciding how to filter data. Prevelance is how many samples have the ASV, and total abundance is how many times ASV is present across all samples. I don't do any filtering until I split up samples into relevant experiment.
```{r A look at prevalence}
#Prevalence-number of samples in which a taxon appears at least once
#Total abundance
prevdf = apply(X = phyloseq::otu_table(ps.decontam),
               MARGIN = ifelse(taxa_are_rows(ps.decontam), yes = 1, no = 2),
               FUN = function(x){sum(x > 0)})

# Add taxonomy and total read counts to this data.frame
prevdf = data.frame(Prevalence = prevdf,
                    TotalAbundance = taxa_sums(ps.decontam),
                    tax_table(ps.decontam))
```

Now I will focus on sod experiment and prepare dataset for further analysis by subsetting these samples from the rest of the sample data. I will also remove allow abundance ASVs not present in at least one sample and not present at least five times
```{r Subset dataset, include=TRUE}
sod.ps<-subset_samples(ps.decontam, Experiment=="basetree_100plots")
sod.ps<- prune_taxa(taxa_sums(sod.ps) > 0, sod.ps)

sod.ps@sam_data$Treatuntreat_EP<-factor(sod.ps@sam_data$Treatuntreat_EP, levels = c(0, 1), labels = c("untreated", "treated"))
sod.ps@sam_data$YearSam_EP <- as.factor(sod.ps@sam_data$YearSam_EP)
sod.ps@sam_data$YrSinceTrt_BR <- as.factor(sod.ps@sam_data$YrSinceTrt_BR)
sod.ps@sam_data$YrSinceTrt_BR <- factor(sod.ps@sam_data$YrSinceTrt_BR, levels = c("untreated", "1to2", "3to4", "5to8", "10to12"))
sod.ps@sam_data$Fire_Mod <- factor(sod.ps@sam_data$Fire_Mod, levels = c("untreated", "wildfire", "SODtrt"))
sod.ps@sam_data$VegBin <- as.factor(sod.ps@sam_data$VegBin)
sod.ps@sam_data$SoilBin <- as.factor(sod.ps@sam_data$SoilBin)

readsums<-sort(sample_sums(sod.ps))
median(readsums)

#Let's filter an ASVs not present at least 5 times
sod_5asv.ps <- prune_taxa(taxa_sums(sod.ps) > 5, sod.ps)
sod_5asv.ps<- prune_taxa(taxa_sums(sod_5asv.ps) > 0, sod_5asv.ps)

save(sod.ps, file="Data/ITS/filtered_data/sodexp.RData")
save(sod_5asv.ps, file="Data/ITS/filtered_data/sod_5asv.RData")

readsums<-sort(sample_sums(sod.ps))
median(readsums)
summary(readsums)
```


### Check prevalence after filtering
```{r check prevalence dataframes after filterig}
#Double check that the filtering makes sense by making new prevalence dataframe
prevdf_check = apply(X = phyloseq::otu_table(sod_5asv.ps),
               MARGIN = ifelse(taxa_are_rows(sod_5asv.ps), yes = 1, no = 2),
               FUN = function(x){sum(x > 0)})

# Add taxonomy and total read counts to this data.frame
prevdf_check = data.frame(Prevalence = prevdf_check,
                    TotalAbundance = taxa_sums(sod_5asv.ps),
                    tax_table(sod_5asv.ps))

```

### Subset read counts so we are just looking at sod samples
```{r subset tracked reads}
samples_to_subset <- samdf %>%
  filter(Experiment=="basetree_100plots") %>% 
  pull(Sample_nameITS)
# Filter the track dataframe

#track$sample <- rownames(track)
subset_track_reads <- track[rownames(track) %in% samples_to_subset, ]
write.csv(subset_track_reads, file="Data/ITS/filtered_data/ITS_subset_track_reads_sod.csv",row.names = TRUE)
# Print summary of the filtered track dataframe
summary(subset_track_reads)
```

### Create prevalance plots (Rel abundance vs. total abundance) facetted by genus
```{r prevalence plots organized by phylum}
prevdf1 = subset(prevdf_check, Genus %in% get_taxa_unique(sod_5asv.ps, "Phylum"))
ggplot(prevdf1, aes(TotalAbundance, Prevalence / nsamples(sod_5asv.ps),color=Phylum)) +
  # Include a guess for parameter
  geom_hline(yintercept = 0.005, alpha = 0.5, linetype = 2) +  geom_point(size = 2, alpha = 0.7) +
  scale_x_log10() +  xlab("Total Abundance") + ylab("Prevalence [Frac. Samples]") +
  facet_wrap(~Phylum) + theme(legend.position="none")
```

### Let's make histograms of read counts before and after filtering
```{r Reads sequenced by experiment}
#Before filtering reads
sdt = data.table::data.table(as(sample_data(sod.ps), "data.frame"),
                 TotalReads = sample_sums(sod.ps), keep.rownames = TRUE)
pseqDepth = ggplot(sdt, aes(TotalReads)) + geom_histogram(color="black", fill="lightblue") + ggtitle("Histogram of sequencing depth-before filtering samples with low read depths") + labs(x="Depth", y="Frequency")
pseqDepth

#After filtering reads
sdt_filt = data.table::data.table(as(sample_data(sod_5asv.ps), "data.frame"),
                 TotalReads = sample_sums(sod_5asv.ps), keep.rownames = TRUE)
pseqDepth_filt = ggplot(sdt_filt, aes(TotalReads)) + geom_histogram(color="black", fill="lightblue") + ggtitle("Histogram of sequencing depth after filtering samples with low read depths") + labs(x="Depth", y="Frequency")
pseqDepth_filt
```

### Software used
```{r, software used}
sessionInfo()
```
