---
title: "ITS SOD Experiments"
output: html_document
date: "2024-05-21"
---

```{r setup, include=TRUE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(root.dir = "~/SOD_DADS_microbiome")
library(phyloseq); packageVersion("phyloseq")
library(ggplot2); packageVersion("ggplot2")
library(readr); packageVersion("readr")
library(purrr); packageVersion("purrr")
library(furrr); packageVersion("furrr")
library(dplyr); packageVersion("dplyr")
library(stringr); packageVersion("stringr")
library(metacoder); packageVersion("metacoder")
library(data.table); packageVersion("data.table")
library(Biostrings); packageVersion("Biostrings")
library(vegan);packageVersion("vegan")
library(agricolae);packageVersion("agricolae")
library(magrittr);packageVersion("magrittr")
library(data.table);packageVersion("data.table")
library(ggbeeswarm);packageVersion("ggbeeswarm")
library(DESeq2);packageVersion("DESeq2")
library(cowplot);packageVersion("cowplot")
theme_set(theme_cowplot())
library(doParallel);packageVersion("doParallel")
library(viridis);packageVersion("viridis")
library(geosphere)
```

### Load data
```{r subset data}
load("Data/ITS/filtered_data/sodexp.RData")
load("Data/ITS/filtered_data/sod_5asv.RData")

filtered_read_counts<-read.csv("Data/ITS/filtered_data/ITS_subset_track_reads_sod.csv")
summary(filtered_read_counts)

new_names <- sapply(strsplit(sod_5asv.ps@sam_data$Sample_nameITS, "-"), function(x) {
  substr(x[1], 3, nchar(x[1]))
})

sod_5asv.ps@sam_data$Sample_nameITS <- new_names
```

### Bar plot agglomerated by species-using filtered dataset and facet wrapped by treatment
```{r barplot-Relative abundance-treatment-order}
dataITS <- sod_5asv.ps %>%
  transform_sample_counts(function(x) {x / sum(x)}) %>% 
  tax_glom(taxrank = "Order") %>%
  psmelt() %>% 
  #mutate(across(Kingdom:Species, ~ ifelse(is.na(.), "Unclassified", .))) %>%  
  arrange(desc(Abundance))

abund_plot <- ggplot(dataITS, aes(x = Sample_nameITS, y = Abundance, fill = Order)) + 
  geom_bar(stat = "identity", position = "stack", color = "black", size = 0.2) +
  scale_fill_viridis_d() +
  scale_color_viridis_d() +
  theme_minimal() +
  labs(
    y = "Relative Abundance",
    fill = "Order"
  ) +
  facet_grid(~factor(Treatuntreat_EP), scales = "free_x") +
  theme(
    axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5, size = 16),
    axis.text.y = element_text(size = 20),  # Increase y-axis label size
    axis.title.x = element_blank(),  # Remove the x-axis title if needed
    axis.title.y = element_text(size = 20), 
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    legend.position = "bottom",
    legend.text = element_text(size = 20),
    legend.title = element_text(size = 20, face = "bold"),
    legend.key.size = unit(1, "lines"),
    strip.text = element_text(size = 24, face= "bold"),
    strip.background = element_blank()
  ) +
  guides(
    fill = guide_legend(
      reverse = FALSE,
      keywidth = 1,
      keyheight = 1,
      title.position = "top",
      title.hjust = 0.5,
      label.theme = element_text(size = 20)
    )
  )

abund_plot <- abund_plot + 
  scale_x_discrete(labels = function(x) gsub("Sample_nameITS", "", x))  # Remove unwanted text

abund_plot

# Save the plot
ggsave("Figures/SOD/ITS/relabundance_plot_treatment_relabund_order_sod.pdf", abund_plot, width = 44, height = 28)
```

### Make a similar plot, but color by Phylum instead
```{r barplot-Relative abundance-treatment-Phylum}
dataITS <- sod_5asv.ps %>%
  transform_sample_counts(function(x) {x / sum(x)}) %>% 
  tax_glom(taxrank = "Phylum") %>%
  psmelt() %>% 
  #mutate(across(Kingdom:Species, ~ ifelse(is.na(.), "Unclassified", .))) %>%  
  arrange(desc(Abundance))

abund_plot <- ggplot(dataITS, aes(x = Sample_nameITS, y = Abundance, fill = Phylum)) + 
  geom_bar(stat = "identity", position = "stack", color = "black", size = 0.2) +
  scale_fill_viridis_d() +
  scale_color_viridis_d() +
  theme_minimal() +
  labs(
    y = "Relative Abundance",
    fill = "Phylum"
  ) +
  facet_grid(~factor(Treatuntreat_EP), scales = "free_x") +
  theme(
    axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5, size = 16),
    axis.text.y = element_text(size = 20),  # Increase y-axis label size
    axis.title.x = element_blank(),  # Remove the x-axis title if needed
    axis.title.y = element_text(size = 20), 
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    legend.position = "bottom",
    legend.text = element_text(size = 20),
    legend.title = element_text(size = 20, face = "bold"),
    legend.key.size = unit(1, "lines"),
    strip.text = element_text(size = 24, face= "bold"),
    strip.background = element_blank()
  ) +
  guides(
    fill = guide_legend(
      reverse = FALSE,
      keywidth = 1,
      keyheight = 1,
      title.position = "top",
      title.hjust = 0.5,
      label.theme = element_text(size = 20)
    )
  )

abund_plot <- abund_plot + 
  scale_x_discrete(labels = function(x) gsub("Sample_nameITS", "", x))  # Remove unwanted text

abund_plot

# Save the plot
ggsave("Figures/SOD/ITS/relabundance_plot_treatment_relabund_Phylum_sod.pdf", abund_plot, width = 24, height = 10)
```

### Bar plot agglomerated by order-facetted by fire factor
```{r barplot-Relative abundance-fire}
dataITS <- sod_5asv.ps %>%
  transform_sample_counts(function(x) {x / sum(x)}) %>% 
  tax_glom(taxrank = "Order") %>%
  psmelt() %>% 
  #mutate(across(Kingdom:Species, ~ ifelse(is.na(.), "Unclassified", .))) %>%  
  arrange(desc(Abundance))

abund_plot <- ggplot(dataITS, aes(x = Sample_nameITS, y = Abundance, fill = Order)) + 
  geom_bar(stat = "identity", position = "stack", color = "black", size = 0.2) +
  scale_fill_viridis_d() +
  scale_color_viridis_d() +
  theme_minimal() +
  labs(
    y = "Relative Abundance",
    fill = "Order"
  ) +
  facet_grid(~factor(Fire_Mod), scales = "free_x") +
  theme(
    axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5, size = 16),
    axis.text.y = element_text(size = 20),  # Increase y-axis label size
    axis.title.x = element_blank(),  # Remove the x-axis title if needed
    axis.title.y = element_text(size = 20), 
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    legend.position = "bottom",
    legend.text = element_text(size = 20),
    legend.title = element_text(size = 20, face = "bold"),
    legend.key.size = unit(1, "lines"),
    strip.text = element_text(size = 24, face= "bold"),
    strip.background = element_blank()
  ) +
  guides(
    fill = guide_legend(
      reverse = FALSE,
      keywidth = 1,
      keyheight = 1,
      title.position = "top",
      title.hjust = 0.5,
      label.theme = element_text(size = 20)
    )
  )

abund_plot <- abund_plot + 
  scale_x_discrete(labels = function(x) gsub("Sample_nameITS", "", x))  # Remove unwanted text

abund_plot

# Save the plot
ggsave("Figures/SOD/ITS/relabundance_plot_fire_relabund_order_sod.pdf", abund_plot, width = 44, height = 28)
```

### Bar plot agglomerated by order-using filtered dataset and faceted by yr since trt
```{r barplot-Relative abundance-yrtrt}
dataITS <- sod_5asv.ps %>%
  transform_sample_counts(function(x) {x / sum(x)}) %>% 
  tax_glom(taxrank = "Order") %>%
  psmelt() %>% 
  #mutate(across(Kingdom:Species, ~ ifelse(is.na(.), "Unclassified", .))) %>%  
  arrange(desc(Abundance))

abund_plot <- ggplot(dataITS, aes(x = Sample_nameITS, y = Abundance, fill = Order)) + 
  geom_bar(stat = "identity", position = "stack", color = "black", size = 0.2) +
  scale_fill_viridis_d() +
  scale_color_viridis_d() +
  theme_minimal() +
  labs(
    y = "Relative Abundance",
    fill = "Family"
  ) +
  facet_grid(~factor(YrSinceTrt_BR), scales = "free_x") +
  theme(
    axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5, size = 16),
    axis.text.y = element_text(size = 20),  # Increase y-axis label size
    axis.title.x = element_blank(),  # Remove the x-axis title if needed
    axis.title.y = element_text(size = 20), 
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    legend.position = "bottom",
    legend.text = element_text(size = 20),
    legend.title = element_text(size = 20, face = "bold"),
    legend.key.size = unit(1, "lines"),
    strip.text = element_text(size = 24, face= "bold"),
    strip.background = element_blank()
  ) +
  guides(
    fill = guide_legend(
      reverse = FALSE,
      keywidth = 1,
      keyheight = 1,
      title.position = "top",
      title.hjust = 0.5,
      label.theme = element_text(size = 20)
    )
  )

abund_plot <- abund_plot + 
  scale_x_discrete(labels = function(x) gsub("Sample_nameITS", "", x))  # Remove unwanted text

abund_plot

# Save the plot
ggsave("Figures/SOD/ITS/relabundance_plot_yrsincetrt_relabund_order_sod.pdf", abund_plot, width = 44, height = 28)
```

### Let's look top taxa at order level
```{r Top taxa order}
top40 <- names(sort(taxa_sums(sod_5asv.ps), decreasing=TRUE))[1:40]
ps.top40 <- transform_sample_counts(sod_5asv.ps, function(OTU) OTU/sum(OTU))
#ps.top40@tax_table[is.na(ps.top40@tax_table)] <- "Unclassified"
ps.top40 <- tax_glom(ps.top40, taxrank = "Order")
ps.top40 <- prune_taxa(top40, ps.top40)

pdf("Figures/SOD/ITS/top40_order_sod.pdf", width = 34, height = 20)
plot_bar(ps.top40, x="Sample_nameITS", fill="Order") + facet_wrap(~Treatuntreat_EP, scales="free_x")
dev.off()
```

### Let's look top taxa at family level
```{r Top taxa family}
top40 <- names(sort(taxa_sums(sod_5asv.ps), decreasing=TRUE))[1:40]
ps.top40 <- transform_sample_counts(sod_5asv.ps, function(OTU) OTU/sum(OTU))
#ps.top40@tax_table[is.na(ps.top40@tax_table)] <- "Unclassified"
ps.top40 <- tax_glom(ps.top40, taxrank = "Family")
ps.top40 <- prune_taxa(top40, ps.top40)

pdf("Figures/SOD/ITS/top40_family_sod.pdf", width = 34, height = 20)
plot_bar(ps.top40, x="Sample_nameITS", fill="Family") + facet_wrap(~Treatuntreat_EP, scales="free_x")
dev.off()
```

### Let's look top taxa at genus level
```{r Top taxa genus}
top40 <- names(sort(taxa_sums(sod_5asv.ps), decreasing=TRUE))[1:40]
ps.top40 <- transform_sample_counts(sod_5asv.ps, function(OTU) OTU/sum(OTU))
#ps.top40@tax_table[is.na(ps.top40@tax_table)] <- "Unclassified"
ps.top40 <- tax_glom(ps.top40, taxrank = "Genus")
ps.top40 <- prune_taxa(top40, ps.top40)

pdf("Figures/SOD/ITS/top40_genus_sod.pdf", width = 34, height = 20)
plot_bar(ps.top40, x="Sample_nameITS", fill="Genus") + facet_wrap(~Treatuntreat_EP, scales="free_x")
dev.off()
```

```{r make data tables-top taxa}
data <-
  sod_5asv.ps %>%
  tax_glom("Species") %>%
  psmelt() %>%
  as_tibble()

data_summed <- data %>%
  filter(Abundance > 0) %>%
  group_by(Kingdom, Phylum, Class, Order, Family, Genus, Species) %>%
  summarise(Abundance = sum(Abundance),
            Prevalence = n_distinct(Sample)) %>%
  ungroup() %>%
  mutate(Total_Abundance = sum(Abundance)) %>%
  mutate(Percentage = 100*(Abundance / Total_Abundance))
  
write.csv(data_summed, file="Results/SOD/ITS/mostabundanttaxa_abundance_sod_species.csv",row.names = TRUE)

data <-
  sod_5asv.ps %>%
  tax_glom("Genus") %>%
  psmelt() %>%
  as_tibble()

data_summed <- data %>%
  filter(Abundance > 0) %>%
  group_by(Kingdom, Phylum, Class, Order, Family, Genus) %>%
  summarise(Abundance = sum(Abundance),
            Prevalence = n_distinct(Sample)) %>%
  ungroup() %>%
  mutate(Total_Abundance = sum(Abundance)) %>%
  mutate(Percentage = 100*(Abundance / Total_Abundance))
  
write.csv(data_summed, file="Results/SOD/ITS/mostabundanttaxa_abundance_sod_genus.csv",row.names = TRUE)

data <-
  sod_5asv.ps %>%
  tax_glom("Family") %>%
  psmelt() %>%
  as_tibble()

data_summed <- data %>%
  filter(Abundance > 0) %>%
  group_by(Kingdom, Phylum, Class, Order, Family) %>%
  summarise(Abundance = sum(Abundance),
            Prevalence = n_distinct(Sample)) %>%
  ungroup() %>%
  mutate(Total_Abundance = sum(Abundance)) %>%
  mutate(Percentage = 100*(Abundance / Total_Abundance))
  
write.csv(data_summed, file="Results/SOD/ITS/mostabundanttaxa_abundance_sod_family.csv",row.names = TRUE)

data1 <-
  sod_5asv.ps %>%
  tax_glom("Species") %>%
  transform_sample_counts(function(x) {x/sum(x)} ) %>%
  psmelt() %>%
  as_tibble()

data_summed1 <- data1 %>%
  filter(Abundance > 0) %>%
  group_by(Treatuntreat_EP, Kingdom, Phylum, Class, Order, Family, Genus, Species) %>%
  summarise(Abundance = sum(Abundance),
            Prevalence = n_distinct(Sample)) 

write.csv(data_summed1, file="Results/SOD/ITS/mostabundanttaxa_relativeabundance_untreatedtreated_sod.csv",row.names = TRUE)

data2 <-
  sod_5asv.ps %>%
  tax_glom("Species") %>%
  psmelt() %>%
  as_tibble()

# Calculate percentages
data_summed2 <- data2 %>%
  filter(Abundance > 0) %>%
  group_by(Treatuntreat_EP, Kingdom, Phylum, Class, Order, Family, Genus, Species) %>%
  summarise(Abundance = sum(Abundance),
            Prevalence = n_distinct(Sample)) %>%
  ungroup() %>%
  mutate(Total_Abundance = sum(Abundance)) %>%
  mutate(Percentage = 100*(Abundance / Total_Abundance))

write.csv(data_summed2, file="Results/SOD/ITS/mostabundanttaxa_untreatedtreated_sod_species.csv",row.names = TRUE)

data3 <-
  sod_5asv.ps %>%
  tax_glom("Genus") %>%
  psmelt() %>%
  as_tibble()

# Calculate percentages
data_summed3 <- data3 %>%
  filter(Abundance > 0) %>%
  group_by(Treatuntreat_EP, Kingdom, Phylum, Class, Order, Family, Genus) %>%
  summarise(Abundance = sum(Abundance),
            Prevalence = n_distinct(Sample)) %>%
  ungroup() %>%
  mutate(Total_Abundance = sum(Abundance)) %>%
  mutate(Percentage = 100*(Abundance / Total_Abundance))

write.csv(data_summed3, file="Results/SOD/ITS/mostabundanttaxa_untreatedtreated_sod_genus.csv",row.names = TRUE)
```

### Let's look at spacial placement of different plots
```{r Map exploration}
pdf("Figures/SOD/ITS/location_of_plots.pdf", width = 20, height = 12)
map_plot <- ggplot(sod_5asv.ps@sam_data, aes(x = Longitude, y = Latitude)) +
  geom_text(aes(color = Fire_Mod, label = Sample_nameITS), size = 2.5, 
            check_overlap = TRUE) +
  labs(x = "Longitude", y = "Latitude", title = "Sampling Locations") +
  theme_minimal() +
  theme(
    plot.title = element_text(size = 20), 
    axis.title = element_text(size = 16),
    axis.text = element_text(color = "black"),
    legend.title = element_text(size = 14),
    legend.text = element_text(size = 12)
  )
print(map_plot)
dev.off()
```

### Let's normalize our data in preparation for beta diversity analyses
```{r Normalization}
#Prune ASVs not present in least 5 times (across one more more samples)
#sod_5asv.ps<- phyloseq::prune_taxa(taxa_sums(sod.ps) > 5, sod.ps)
sdt = data.table::data.table(as(sample_data(sod_5asv.ps), "data.frame"),
                 TotalReads = sample_sums(sod_5asv.ps), keep.rownames = TRUE)
min_reads<-min(sdt$TotalReads)

rarefied.ps <- rarefy_even_depth(sod_5asv.ps, sample.size = min_reads, rngseed = 100)

#Relative abundance
strat_rel.ps<-transform_sample_counts(sod_5asv.ps, function(x) x / sum(x))
```

### Alpha diversity-use non-normalized or pruned data
Let's start with t-test as point of comparison with anova analysis
```{r alpha diversity analyses-t-test}
set.seed(1)
AD <- estimate_richness(sod.ps, measures = c("Observed", "InvSimpson", "Shannon"))
rownames(AD) <- gsub("\\.", "-", rownames(AD))

# Combine richness estimates with sample metadata
AD <- cbind(AD, sample_data(sod.ps)[row.names(AD), c("Treatuntreat_EP", "Lineage_EP", "SoilBin", "VegBin")])

# Convert to data.table and keep row names as a column
AD <- data.table(AD, keep.rownames = TRUE)

# Reshape the data frame
AD <- melt(AD, id.vars = c('rn', c("Treatuntreat_EP", "Lineage_EP", "SoilBin", "VegBin")))

# Now conduct t-tests (assuming you have appropriate factors)
factors <- c("Treatuntreat_EP", "Lineage_EP", "SoilBin", "VegBin")
# Iterate through each factor
ttest_list <- list()

# Iterate through each factor
for (factor_name in factors) {
  # Initialize an empty data.table to store t-test results for each alpha diversity measure
  ttest <- data.table(alpha_index = character(), pvalue = numeric())
  
  # Iterate through each alpha diversity measure
  for (alphad in c("InvSimpson", "Shannon", "Observed")) {
    # Perform t-test
    ttest_res <- t.test(value ~ get(factor_name), data = AD[variable == alphad], var.equal = TRUE)
    
    # Extract p-value from t-test results and store it in the data.table
    p_value <- ttest_res$p.value
    ttest <- rbind(ttest, data.table(alpha_index = alphad, pvalue = p_value))
  }
  
  # Add t-test results for the current factor to the list
  ttest_list[[factor_name]] <- ttest
}

# Print t-test results
for (factor_name in factors) {
  cat(paste("T-Test Results for Factor:", factor_name), "\n")
  print(knitr::kable(ttest_list[[factor_name]], format = "markdown"))
  cat("\n")
}
``` 

### Let's now proceed with ANOVA analyses and then compile our alpha diversity stats
```{r compiled alpha diversity table}
# Initialize an empty data.table to store combined ANOVA results for all factors
set.seed(1)
anova_list <- list()

AD <- estimate_richness(sod.ps, measures = c("Observed", "InvSimpson", "Shannon"))
rownames(AD) <- gsub("\\.", "-", rownames(AD))

# Combine richness estimates with sample metadata
AD <- cbind(AD, sample_data(sod.ps)[row.names(AD), c("Treatuntreat_EP", "Fire_Mod", "YrSinceTrt_BR")])

# Convert to data.table and keep row names as a column
AD <- data.table(AD, keep.rownames = TRUE)

# Reshape the data frame
AD <- melt(AD, id.vars = c('rn', c("Treatuntreat_EP", "Fire_Mod", "YrSinceTrt_BR")))

factors2 <- c("Treatuntreat_EP", "Fire_Mod", "YrSinceTrt_BR")

# Initialize a vector to store factor names for which a plot has been created
plotted_factors <- character()
 

final_anova_table <- data.table(factor_name = character(), alpha_index = character(), pvalue = numeric(), df_between = numeric(), df_within = numeric(), ss_between = numeric(), ss_within = numeric(), ms_between = numeric(), ms_within = numeric(), F_value = numeric())

for (factor_name in factors2) {
  # Initialize an empty data.table to store ANOVA results for each alpha diversity measure
  anova <- data.table(factor_name = character(), alpha_index = character(), pvalue = numeric(), df_between = numeric(), df_within = numeric(), ss_between = numeric(), ss_within = numeric(), ms_between = numeric(), ms_within = numeric(), F_value = numeric())
  
  # Iterate through each alpha diversity measure
  for (alphad in c("Observed","InvSimpson", "Shannon")) {
    # Perform ANOVA
    anova_res <- aov(value ~ get(factor_name), data = AD[variable == alphad])
    
    # Extract relevant statistics from ANOVA results
    anova_summary <- summary(anova_res)
    
    # Extract p-value from ANOVA results
    p_value <- anova_summary[[1]]$`Pr(>F)`[1]
    
    # Extract other statistics
    df_between <- anova_summary[[1]]$"Df"[1]
    df_within <- anova_summary[[1]]$"Df"[2]
    ss_between <- anova_summary[[1]]$"Sum Sq"[1]
    ss_within <- anova_summary[[1]]$"Sum Sq"[2]
    ms_between <- anova_summary[[1]]$"Mean Sq"[1]
    ms_within <- anova_summary[[1]]$"Mean Sq"[2]
    F_value <- anova_summary[[1]]$"F value"[1]
    
    # Store statistics in the data.table
    anova <- rbind(anova, data.table(factor_name = factor_name, alpha_index = alphad, pvalue = p_value, df_between = df_between, df_within = df_within, ss_between = ss_between, ss_within = ss_within, ms_between = ms_between, ms_within = ms_within, F_value = F_value))
  }
  
  # Append ANOVA results for the current factor to the final table
  final_anova_table <- rbind(final_anova_table, anova)
}

# Print final ANOVA table
print(knitr::kable(final_anova_table, format = "markdown"))
write.csv(final_anova_table, "Results/SOD/ITS/final_anova_table_sod.csv", row.names = FALSE)
```

### Let's make faceted boxplots for our alpha diversity analyses
```{r alpha diversity faceted plots}
# Estimate alpha diversity
set.seed(1)
AD <- estimate_richness(sod.ps, measures = c("Observed", "InvSimpson", "Shannon"))
rownames(AD) <- gsub("\\.", "-", rownames(AD))

# Combine richness estimates with sample metadata
AD <- cbind(AD, sample_data(sod_5asv.ps)[rownames(AD), c("Treatuntreat_EP", "Fire_Mod", "YrSinceTrt_BR")])


variables <- c("Treatuntreat_EP", "Fire_Mod", "YrSinceTrt_BR")


# Create an empty list to store plots
plots <- list()

# Loop through each variable and create the plot
for (variable in variables) {
  p <- plot_richness(sod.ps, x = variable, measures = c("Shannon", "InvSimpson", "Observed")) + 
    geom_boxplot()
  plots[[variable]] <- p
}

# Arrange plots in a grid
plots_arrange <- gridExtra::grid.arrange(
  plots[["Treatuntreat_EP"]], plots[["Fire_Mod"]], plots[["YrSinceTrt_BR"]], plots[["YearSam_EP"]], plots[["Lineage_EP"]], plots[["VegBin"]], plots[["SoilBin"]],ncol=1)

# Add annotations to the top of each plot
plots_arrange_with_text <- cowplot::ggdraw() +
  cowplot::draw_plot(plots_arrange) +
  cowplot::draw_plot("Treatuntreat_EP", x = 0.28, y = 0.98) +
  cowplot::draw_plot("Fire_EP", x = 0.28, y = 0.98) +
  cowplot::draw_plot("YrSinceTrt_BR", x = 0.28, y = 0.98) 

# Save the arranged plots
ggsave(plots_arrange_with_text, path = "Figures/SOD/ITS", filename = "alpha_diversity_each_factor_sod.pdf", 
       width = 16, height = 24)
```

### Let's perform TUKEY HSD tests
```{r alpha diversity compilation with tukey summary}
# Define your variables of interest
set.seed(1)
variables <- c("Treatuntreat_EP", "Fire_Mod", "YrSinceTrt_BR")
measures <- c("Observed", "InvSimpson", "Shannon")

# Create an empty list to store ANOVA and Tukey's results
tukey_results <- list()

# Loop through each variable and measure
for (variable in variables) {
  for (measure in measures) {
    # Subset the data for the current variable and measure
    data_subset <- AD[, c(variable, measure)]
    
    # Perform ANOVA
    formula <- as.formula(paste(measure, "~ ."))
    anova_res <- aov(formula, data = data_subset)
    
    # Perform Tukey's post-hoc test
    tukey_result <- HSD.test(anova_res, variable)
    
    # Store Tukey's results
    print(tukey_result)
  }
}
```

### Beta diversity-PERMANOVA analyses-ASV counts transformed (rel. abundance)
```{r PERMANOVA ANALYSES-relative abundance, include=TRUE}
#I used the filtered dataset (removed low abundance counts less than 2)
set.seed(1)
mapping <- as.data.frame(sample_data(strat_rel.ps))

mapping2=mapping[, c("Treatuntreat_EP", "Fire_Mod", "YrSinceTrt_BR")]

ASV_tables_bray <- phyloseq::distance(strat_rel.ps, method = "bray")

df_metadata <- data.frame(sample_data(strat_rel.ps))

df_metadata<-df_metadata[, c("Treatuntreat_EP", "Fire_Mod", "YrSinceTrt_BR")]

pvals <- sapply(colnames(mapping2), function(variable){
  
  #Create formula for adonis() 
  form1<-as.formula(paste("ASV_tables_bray", variable, sep="~"))
  
  res <- adonis2(form1, data = df_metadata, permutations = 9999)
  
  # extract pvalue from adonis output
  pval <- res$`Pr(>F)`[[1]] # we take the first p value which corresponds to the variable of interest
  return(pval)
})

#Adjust p_values with fdr correction
pval_adj <- p.adjust(pvals, method = "fdr")

#Find significant variables with lower than 0.05 for their p values
save.pval.adonis0.05_braycurtis <- pval_adj[pval_adj < 0.05]
save.pval.adonis0.05_braycurtis

print(pval_adj)
```

### Compile PERMANOVA stats
```{r compile PERMANOVA stats}
# Perform PERMANOVA -main effects
set.seed(1)
res <- adonis2(ASV_tables_bray ~Treatuntreat_EP+Fire_Mod+YrSinceTrt_BR, data = df_metadata, permutations = 9999)

# Extract data from res
res_data <- data.frame(res)
colnames(res_data)[colnames(res_data) == "Pr..F."] <- "pr_f"

# Remove the last two rows (Residual and Total)
res_data <- res_data[-c(nrow(res_data)-1, nrow(res_data)),]

# Add FDR and Bonferroni adjusted p-values
res_data$FDR_adj_p <- p.adjust(res_data$pr_f, method = "fdr")
res_data$Bonferroni_adj_p <- p.adjust(res_data$pr_f, method = "bonferroni")
res_data$Factor <- rownames(res_data)

# Reorder columns
res_data <- res_data[, c("Factor", "Df", "SumOfSqs", "R2", "F", "pr_f", "FDR_adj_p", "Bonferroni_adj_p")]
rownames(res_data) <- NULL

# Perform beta dispersion analysis for main effects
beta_disp_results <- lapply(c("Treatuntreat_EP", "Fire_Mod", "YrSinceTrt_BR"), function(variable) {
  res_betadisper <- betadisper(d = ASV_tables_bray, group = df_metadata[[variable]])
  perm_test_res <- permutest(res_betadisper)
  pval_betadisper <- perm_test_res$tab$`Pr(>F)`[[1]]
  
  data.frame(Factor = variable, P_Value_BetaDispersion = pval_betadisper)
})

beta_disp_results_df <- do.call(rbind, beta_disp_results)

# Merge PERMANOVA results with beta dispersion results
final_results <- merge(res_data, beta_disp_results_df, by = "Factor", all.x = TRUE)

# Reorder columns
final_results <- final_results[, c("Factor", "Df", "SumOfSqs", "R2", "F", "pr_f", "FDR_adj_p", "Bonferroni_adj_p", "P_Value_BetaDispersion")]

# Print the table
print(final_results, row.names = FALSE)

# Save results
write.csv(final_results, file = "Results/SOD/ITS/PERMANOVA_interactions_with_betadisper_maineffects.csv", row.names = FALSE)
```

### Compile PERMANOVA stats-interaction
```{r compile interaction stats}
# Perform PERMANOVA -main effects
set.seed(1)
res <- adonis2(ASV_tables_bray ~ Treatuntreat_EP*YrSinceTrt_BR*Fire_Mod, data = df_metadata, permutations = 9999)
# Extract data from res
res_data <- data.frame(res)
colnames(res_data)[colnames(res_data) == "Pr..F."] <- "pr_f"

# Remove the last two rows (Residual and Total)
res_data <- res_data[-c(nrow(res_data)-1, nrow(res_data)),]

# Add FDR and Bonferroni adjusted p-values
res_data$FDR_adj_p <- p.adjust(res_data$pr_f, method = "fdr")
res_data$Bonferroni_adj_p <- p.adjust(res_data$pr_f, method = "bonferroni")
res_data$Factor <- rownames(res_data)

# Reorder columns
res_data <- res_data[, c("Factor", "Df", "SumOfSqs", "R2", "F", "pr_f", "FDR_adj_p", "Bonferroni_adj_p")]
rownames(res_data) <- NULL

# Perform beta dispersion analysis for main effects
beta_disp_results <- lapply(c("Treatuntreat_EP", "Fire_Mod"), function(variable) {
  res_betadisper <- betadisper(d = ASV_tables_bray, group = df_metadata[[variable]])
  perm_test_res <- permutest(res_betadisper)
  pval_betadisper <- perm_test_res$tab$`Pr(>F)`[[1]]
  
  data.frame(Factor = variable, P_Value_BetaDispersion = pval_betadisper)
})

beta_disp_results_df <- do.call(rbind, beta_disp_results)

# Merge PERMANOVA results with beta dispersion results
final_results <- merge(res_data, beta_disp_results_df, by = "Factor", all.x = TRUE)

# Reorder columns
final_results <- final_results[, c("Factor", "Df", "SumOfSqs", "R2", "F", "pr_f", "FDR_adj_p", "Bonferroni_adj_p", "P_Value_BetaDispersion")]

# Print the table
print(final_results, row.names = FALSE)

# Save results
write.csv(final_results, file = "Results/SOD/ITS/PERMANOVA_interactions_with_betadisper_firemod.csv", row.names = FALSE)
```

### Let's now examine beta diversity through ordination plots
We'll will examine outputs for rarefied data as well as data adjusted by relative abundance
We'll now make ordination plots on data for which we've calculated relative abundance
```{r Beta diversity-Rel. abundance-treated and untreated samples}
#Relative abundance
set.seed(1)
ps_NMDS = ordinate(strat_rel.ps, 
                   method="NMDS", 
                   distance="bray")

plot<- plot_ordination(strat_rel.ps, 
                         ps_NMDS,
                         color= "Treatuntreat_EP") +
  ggtitle("NMDS ordination plot - relative abundance") +
  labs(color = "Factor") + 
  theme(axis.text=element_text(size=14), axis.title=element_text(size=14), title =  element_text(size=13), legend.text = element_text(size = 18))

plot_ellipses<-plot +
   stat_ellipse(type = "norm", linetype = 2) +
  theme_bw()

plot_ellipses

#Export NMDS coordinates per sample
write.csv(ps_NMDS[["points"]], file="Results/SOD/ITS/ITS_NMDS_coordinates_persample.csv",row.names = TRUE)
```

### Now, we'll look at ordination plot (rel abundance dataset), with Fire_Mod factor
```{r Beta diversity-Rel. abundance-Fire_Mod}
#Relative abundance
set.seed(1)
ps_NMDS = ordinate(strat_rel.ps, 
                   method="NMDS",
                   distance="bray")

plot<- plot_ordination(strat_rel.ps, 
                         ps_NMDS,
                         labs(color = "Factor"),
                         color= "Fire_Mod") +
  ggtitle("NMDS ordination plot - relative abundance") +
  theme(axis.text=element_text(size=14), axis.title=element_text(size=14), title =  element_text(size=13), legend.text = element_text(size = 18))


plot_ellipses<-plot +
   stat_ellipse(type = "norm", linetype = 2) +
  theme_bw()

plot_ellipses
```

### Now, we'll look at ordination plot rel abundance-with YrSinceTrt_BR factor
```{r Beta diversity-Rel. abundance-YrSinceTrt_BR}
#Relative abundance
set.seed(1)
ps_NMDS = ordinate(strat_rel.ps, 
                   method="NMDS",
                   distance="bray")

plot<- plot_ordination(strat_rel.ps, 
                         ps_NMDS,
                         labs(color = "Factor"),
                         color= "YrSinceTrt_BR") +
  ggtitle("NMDS ordination plot - relative abundance") +
  theme(axis.text=element_text(size=14), axis.title=element_text(size=14), title =  element_text(size=13), legend.text = element_text(size = 18))


plot_ellipses<-plot +
   stat_ellipse(type = "norm", linetype = 2) +
  theme_bw()

plot_ellipses
```


### Are communities more similar if geographically closeby? Some plots, like wildfire plots are close together, for example
Very preliminary experiment
```{r Beta diversity-Rel. abundance-YearSam_EP}
# Extract OTU table and sample data from phyloseq object
otu_table <- otu_table(strat_rel.ps)
sample_data <- sample_data(strat_rel.ps)

# Transpose OTU table for correct format
otu_table_matrix <- t(as(otu_table, "matrix"))

# Extract longitude and latitude
lon_lat <- cbind(sample_data$Longitude, sample_data$Latitude)

# Calculate distance matrices
geo_dist <- distm(lon_lat, fun = distHaversine)
comm_dist <- vegdist(otu_table_matrix, method = "bray", na.rm = TRUE)

# Convert Bray-Curtis distance to matrix if needed
comm_dist_matrix <- as.matrix(comm_dist)

# Extract sample IDs
sample_ids <- rownames(sample_data)
sample_ids_in_otu <- rownames(otu_table)

# Match sample IDs
common_samples <- intersect(sample_ids_in_otu, sample_ids)
common_indices <- which(sample_ids %in% common_samples)

# Subset distance matrices
geo_dist_aligned <- geo_dist[common_indices, common_indices]
comm_dist_aligned <- comm_dist_matrix[common_indices, common_indices]

# Check dimensions
print(dim(geo_dist_aligned))
print(dim(comm_dist_aligned))

# Perform Mantel test
mantel_result <- mantel(geo_dist_aligned, comm_dist_aligned, method = "spearman", permutations = 999)
print(mantel_result)

# Extract OTU table and sample data from phyloseq object
otu_table <- otu_table(strat_rel.ps)
sample_data <- sample_data(strat_rel.ps)

# Transpose OTU table for correct format
otu_table_matrix <- as(otu_table, "matrix")

# Calculate Bray-Curtis dissimilarity matrix
comm_dist <- vegdist(otu_table_matrix, method = "bray")

# Run NMDS
nmds_result <- metaMDS(comm_dist, k = 2, trymax = 100)

# Extract NMDS scores
nmds_scores <- as.data.frame(scores(nmds_result))
nmds_scores$SampleID <- rownames(nmds_scores)

# Extract sample IDs and coordinates correctly
sample_ids <- rownames(sample_data(strat_rel.ps))  # Correct sample IDs
sample_coords <- data.frame(SampleID = sample_ids,
                             Longitude = sample_data(strat_rel.ps)$Longitude,
                             Latitude = sample_data(strat_rel.ps)$Latitude, 
                             Treatuntreat=sample_data(strat_rel.ps)$Treatuntreat_EP,
                             Fire_Mod=sample_data(strat_rel.ps)$Fire_Mod
                            )

# Merge NMDS scores with the correct sample coordinates
nmds_scores <- merge(nmds_scores, sample_coords, by = "SampleID")


# Create NMDS plot with color for Latitude and size for Longitude
nmds_plot <- ggplot(nmds_scores, aes(x = NMDS1, y = NMDS2)) +
  geom_point(aes(color = Latitude, size = Longitude), alpha = 0.7) +  # Use size for Longitude
  labs(title = "NMDS Plot with Latitude and Longitude",
       x = "NMDS Dimension 1",
       y = "NMDS Dimension 2") +
  theme_minimal() +
  scale_color_gradient(low = "blue", high = "red") +  # Color gradient for Latitude
  scale_size_continuous(range = c(2, 6))  # Adjust size range as needed

print(nmds_plot)

nmds_plot <- ggplot(nmds_scores, aes(x = NMDS1, y = NMDS2)) +
  geom_point(aes(color = Latitude, size = Longitude), alpha = 0.7) +
  facet_wrap(~ Fire_Mod) +
  labs(title = "NMDS Plot with Latitude, Longitude, and Treatment",
       x = "NMDS Dimension 1",
       y = "NMDS Dimension 2") +
  theme_minimal() +
  scale_color_gradient(low = "blue", high = "red") +
  scale_size_continuous(range = c(2, 6))

print(nmds_plot)

```

### Functions to prepare to do differential abundance analyses
```{r organizing data for downstream steps}
sample.data.table <- function(ps) {
  as(sample_data(ps), "data.frame") %>%
    as.data.table(keep.rownames = "Sample") %>% 
    setkey(Sample) %>%
    return()
}
asv.data.table <- function(ps) {
  if (taxa_are_rows(ps)) {
    otu.mat <- t(as(otu_table(ps), "matrix"))
  } else {
    otu.mat <- as(otu_table(ps), "matrix")
  }
  return(as.data.table(otu.mat, keep.rownames = "Sample") %>% setkey(Sample))
}
cat.n <- function(x) cat(x, sep = "\n")
```

### Now we import data
```{r data-import-and-prep}
sod.data <- sample.data.table(sod_5asv.ps)
sod.asvs <- asv.data.table(sod_5asv.ps)
sod.asv.ids <- taxa_names(sod_5asv.ps)
sod.dt <- sod.data[sod.asvs]
```

### Let's now do a DESEQ2 analysis-firemod
### DESEQ analysis-Fire_mod
```{r, diff abundance-deseq2 fire_mod}
set.seed(1)

sod.deseq <- phyloseq_to_deseq2(sod_5asv.ps, ~ Fire_Mod)

sod.dds <- estimateSizeFactors(sod.deseq, type = 'poscounts')

sod.dds <- DESeq(sod.dds)

all_results <- resultsNames(sod.dds)

all_results <- all_results[all_results != "Intercept"]

process_result <- function(result_name) {
  res.dt <- as.data.table(results(sod.dds, name = result_name), keep.rownames = "ASV")
  
  sig.dt <- res.dt[padj <= 0.05]
  
  sig.dt <- cbind(as(sig.dt, "data.frame"), as(tax_table(sod_5asv.ps)[sig.dt$ASV, ], "matrix"))
  
  sig.dt <- sig.dt %>% mutate(across(everything(), ~ifelse(is.na(.), "Unclassified", .)))
  
  write.csv(sig.dt, file = paste0("Results/SOD/ITS/deseq_", result_name, ".csv"), row.names = TRUE)
  
  plot <- ggplot(sig.dt, aes(y = reorder(ASV, log2FoldChange), x = log2FoldChange, fill = Genus)) +
    geom_vline(xintercept = 0.0, color = "gray", size = 0.5, linetype = "dashed") +
    geom_bar(stat = "identity", position = "dodge", width = 0.8) +
    scale_fill_viridis(discrete = TRUE, option = "D") +
    coord_flip() +
    labs(x = "Log2 Fold Change",
         y = "ASV",
         title = paste("Differential Abundance Analysis:", result_name)) +
    theme_minimal() +
    theme(
      axis.text.y = element_text(size = 12),
      axis.text.x = element_text(angle = 45, size=12),
      axis.title = element_text(face = "bold"),
      legend.position = "right",
      legend.title = element_text(face = "bold"),
      plot.title = element_text(size = 16, face = "bold"),
      plot.subtitle = element_text(size = 12),
      panel.grid.major.y = element_blank(),
      panel.grid.minor = element_blank()
    )
  
  ggsave(paste0("Figures/SOD/ITS/deseq_", result_name, "_barplot.pdf"), plot, width = 34, height = 12)
  
  print(plot)
}
lapply(all_results, process_result)
```

### Let's now do a DESEQ2 analysis-firemod2
### DESEQ analysis-Fire_mod
```{r, diff abundance-deseq2 fire_mod2}
set.seed(1)

sod_5asv.ps@sam_data$Fire_Mod <- factor(sod_5asv.ps@sam_data$Fire_Mod, levels = c( "wildfire", "untreated", "SODtrt"))

sod.deseq <- phyloseq_to_deseq2(sod_5asv.ps, ~ Fire_Mod)

sod.dds <- estimateSizeFactors(sod.deseq, type = 'poscounts')

sod.dds <- DESeq(sod.dds)

all_results <- resultsNames(sod.dds)

all_results <- all_results[all_results != "Intercept"]

process_result <- function(result_name) {
  res.dt <- as.data.table(results(sod.dds, name = result_name), keep.rownames = "ASV")
  
  sig.dt <- res.dt[padj <= 0.05]
  
  sig.dt <- cbind(as(sig.dt, "data.frame"), as(tax_table(sod_5asv.ps)[sig.dt$ASV, ], "matrix"))
  
  sig.dt <- sig.dt %>% mutate(across(everything(), ~ifelse(is.na(.), "Unclassified", .)))
  
  write.csv(sig.dt, file = paste0("Results/SOD/ITS/deseq_", result_name, ".csv"), row.names = TRUE)
  
  plot <- ggplot(sig.dt, aes(y = reorder(ASV, log2FoldChange), x = log2FoldChange, fill = Genus)) +
    geom_vline(xintercept = 0.0, color = "gray", size = 0.5, linetype = "dashed") +
    geom_bar(stat = "identity", position = "dodge", width = 0.8) +
    scale_fill_viridis(discrete = TRUE, option = "D") +
    coord_flip() +
    labs(x = "Log2 Fold Change",
         y = "ASV",
         title = paste("Differential Abundance Analysis:", result_name)) +
    theme_minimal() +
    theme(
      axis.text.y = element_text(size = 12),
      axis.text.x = element_text(angle = 45, size=12),
      axis.title = element_text(face = "bold"),
      legend.position = "right",
      legend.title = element_text(face = "bold"),
      plot.title = element_text(size = 16, face = "bold"),
      plot.subtitle = element_text(size = 12),
      panel.grid.major.y = element_blank(),
      panel.grid.minor = element_blank()
    )
  
  ggsave(paste0("Figures/SOD/ITS/deseq_", result_name, "_barplot.pdf"), plot, width = 20, height = 12)
  
  print(plot)
}
lapply(all_results, process_result)
```
### DESEQ analysis-treatuntreat
```{r, diff abundance-deseq2 treatuntreat}
set.seed(1)

sod.deseq <- phyloseq_to_deseq2(sod_5asv.ps, ~ Treatuntreat_EP)

sod.dds <- estimateSizeFactors(sod.deseq, type = 'poscounts')

sod.dds <- DESeq(sod.dds)

all_results <- resultsNames(sod.dds)

all_results <- all_results[all_results != "Intercept"]

process_result <- function(result_name) {
  res.dt <- as.data.table(results(sod.dds, name = result_name), keep.rownames = "ASV")
  
  sig.dt <- res.dt[padj <= 0.05]
  
  sig.dt <- cbind(as(sig.dt, "data.frame"), as(tax_table(sod_5asv.ps)[sig.dt$ASV, ], "matrix"))
  
  sig.dt <- sig.dt %>% mutate(across(everything(), ~ifelse(is.na(.), "Unclassified", .)))
  
  write.csv(sig.dt, file = paste0("Results/SOD/ITS/deseq_", result_name, ".csv"), row.names = TRUE)
  
  plot <- ggplot(sig.dt, aes(y = reorder(ASV, log2FoldChange), x = log2FoldChange, fill = Genus)) +
    geom_vline(xintercept = 0.0, color = "gray", size = 0.5, linetype = "dashed") +
    geom_bar(stat = "identity", position = "dodge", width = 0.8) +
    scale_fill_viridis(discrete = TRUE, option = "D") +
    coord_flip() +
    labs(x = "Log2 Fold Change",
         y = "ASV",
         title = paste("Differential Abundance Analysis:", result_name)) +
    theme_minimal() +
    theme(
      axis.text.y = element_text(size = 12),
      axis.text.x = element_text(angle = 45, size=12),
      axis.title = element_text(face = "bold"),
      legend.position = "right",
      legend.title = element_text(face = "bold"),
      plot.title = element_text(size = 16, face = "bold"),
      plot.subtitle = element_text(size = 12),
      panel.grid.major.y = element_blank(),
      panel.grid.minor = element_blank()
    )
  
  ggsave(paste0("Figures/SOD/ITS/deseq_", result_name, "_barplot.pdf"), plot, width = 20, height = 12)
  
  print(plot)
}
lapply(all_results, process_result)
```


### DESEQ analysis-yearsincetrt
```{r, diff abundance-deseq2 yearsincetrt}
set.seed(1)
sod_5asv.ps@sam_data$YrSinceTrt_BR <- factor(sod_5asv.ps@sam_data$YrSinceTrt_BR, levels = c("5to8","3to4","10to12","untreated","1to2"))

sod.deseq <- phyloseq_to_deseq2(sod_5asv.ps, ~ YrSinceTrt_BR)

sod.dds <- estimateSizeFactors(sod.deseq, type = 'poscounts')

sod.dds <- DESeq(sod.dds)

all_results <- resultsNames(sod.dds)

all_results <- all_results[all_results != "Intercept"]

process_result <- function(result_name) {
  res.dt <- as.data.table(results(sod.dds, name = result_name), keep.rownames = "ASV")
  
  sig.dt <- res.dt[padj <= 0.05]
  
  sig.dt <- cbind(as(sig.dt, "data.frame"), as(tax_table(sod_5asv.ps)[sig.dt$ASV, ], "matrix"))
  
  sig.dt <- sig.dt %>% mutate(across(everything(), ~ifelse(is.na(.), "Unclassified", .)))
  
  write.csv(sig.dt, file = paste0("Results/SOD/ITS/deseq_", result_name, ".csv"), row.names = TRUE)
  
  plot <- ggplot(sig.dt, aes(y = reorder(ASV, log2FoldChange), x = log2FoldChange, fill = Genus)) +
    geom_vline(xintercept = 0.0, color = "gray", size = 0.5, linetype = "dashed") +
    geom_bar(stat = "identity", position = "dodge", width = 0.8) +
    scale_fill_viridis(discrete = TRUE, option = "D") +
    coord_flip() +
    labs(x = "Log2 Fold Change",
         y = "ASV",
         title = paste("Differential Abundance Analysis:", result_name)) +
    theme_minimal() +
    theme(
      axis.text.y = element_text(size = 12),
      axis.text.x = element_text(angle = 45, size=12),
      axis.title = element_text(face = "bold"),
      legend.position = "right",
      legend.title = element_text(face = "bold"),
      plot.title = element_text(size = 16, face = "bold"),
      plot.subtitle = element_text(size = 12),
      panel.grid.major.y = element_blank(),
      panel.grid.minor = element_blank()
    )
  
  ggsave(paste0("Figures/SOD/ITS/deseq_", result_name, "_barplot.pdf"), plot, width = 20, height = 12)
  
  print(plot)
}
lapply(all_results, process_result)
```

### For nicer visuals, let's make heat trees (specifically the comparison between untreated and treated samples), and we'll start with Wilcoxon rank sum-treatment

### Convert taxmap to phyloseq obj
```{r, convert to phyloseq obj}
remove_boot_columns <- function(x) {
  x <- x[, !grepl("boot", colnames(x))]
  return(x)
}

sod_5asv_tax <- tax_table(sod_5asv.ps) %>%
  as.data.frame() %>%
  remove_boot_columns() %>%
  as.matrix() %>%
  tax_table()

tax_table(sod_5asv.ps) <- sod_5asv_tax

obj_ITS<-parse_phyloseq(sod_5asv.ps)
obj_ITS_rarefy <-parse_phyloseq(rarefied.ps)

obj_ITS$data$tax_abund <- calc_taxon_abund(obj_ITS, data = "otu_table", cols = obj_ITS$data$sample_data$sample_id)
obj_ITS_rarefy$data$tax_abund <- calc_taxon_abund(obj_ITS_rarefy , data = "otu_table", cols = obj_ITS_rarefy$data$sample_data$sample_id)

obj_ITS$data$abund_prop <- calc_obs_props(obj_ITS, data = "otu_table", cols = obj_ITS$data$sample_data$sample_id)

obj_ITS$data$tax_prop <- calc_taxon_abund(obj_ITS, data = "abund_prop", cols = obj_ITS$data$sample_data$sample_id)

obj_ITS$data$tax_prop$all_samples <- rowMeans(obj_ITS$data$tax_prop[,obj_ITS$data$sample_data$sample_id])
```


### We'll then do DESEQ analysis on untreated vs. untreated samples
```{r, diff abundance trees-deseq}
# set.seed(1)
# obj_ITS$data$diff_table_deseq <- metacoder::calc_diff_abund_deseq2(obj_ITS,
#                                       data = "tax_abund",
#                                       cols = obj_ITS$data$sample_data$sample_id, # What columns of sample data to
#                                       groups = obj_ITS$data$sample_data$Treatuntreat_EP) # What category each sample is assigned to
# 
# obj_ITS$data$diff_table_deseq$log2FoldChange[obj_ITS$data$diff_table_deseq$padj > 0.05] <- 0
# 
# range(obj_ITS$obj_ITS$data$diff_table_deseq$padj, finite = TRUE) 
# range(obj_ITS$data$diff_table_deseq$log2FoldChange, finite = TRUE) 
# 
# obj_ITS<- metacoder::filter_taxa(obj_ITS, taxon_names != "None", reassign_obs = FALSE)
# obj_ITS <- metacoder::filter_taxa(obj_ITS, taxon_names != "", reassign_obs = FALSE)
# obj_ITS <- metacoder::filter_taxa(obj_ITS, taxon_names != "NA", reassign_obs = FALSE)
# 
# 
# obj_ITS %>%
#   metacoder::filter_taxa(taxon_ranks == "Species", supertaxa = TRUE, reassign_obs = FALSE) %>%
#   heat_tree_matrix(data = "diff_table_deseq",
#                    node_label = taxon_names,
#                    node_size = n_obs, # number of OTUs
#                    node_color = log2FoldChange, # difference between groups
#                    node_color_trans = "linear",
#                    node_color_interval = c(-1, 1)*max(abs(range(obj_ITS$data$diff_table_deseq$log2FoldChange))), 
#                    edge_color_interval = c(-1, 1)*max(abs(range(obj_ITS$data$diff_table_deseq$log2FoldChange))),  
#                    node_color_range = diverging_palette(), # diverging colors
#                    node_size_axis_label = "ASV count",
#                    node_color_axis_label = "Log 2 Fold Change",
#                    layout = "da", initial_layout = "re",
#                    key_size = 0.4,
#                    output_file = "Figures/SOD/ITS/diff_abundance_heattree_deseq_sod_ITS_trtuntreat.pdf") # The layout algorithm that initializes node locations
```

### Make tree-deseq-Fire_Mod
```{r, diff abundance trees-Wilcoxon rank sum fire}
# set.seed(1)
# obj_ITS$data$diff_table_deseq <- metacoder::calc_diff_abund_deseq2(obj_ITS,
#                                       data = "tax_abund",
#                                       cols = obj_ITS$data$sample_data$sample_id, # What columns of sample data to
#                                       groups = obj_ITS$data$sample_data$Fire_EP) # What category each sample is assigned to
# 
# obj_ITS$data$diff_table_deseq$log2FoldChange[obj_ITS$data$diff_table_deseq$padj > 0.05] <- 0
# 
# range(obj_ITS$obj_ITS$data$diff_table_deseq$padj, finite = TRUE) 
# range(obj_ITS$data$diff_table_deseq$log2FoldChange, finite = TRUE) 
# 
# obj_ITS<- metacoder::filter_taxa(obj_ITS, taxon_names != "None", reassign_obs = FALSE)
# obj_ITS <- metacoder::filter_taxa(obj_ITS, taxon_names != "", reassign_obs = FALSE)
# obj_ITS <- metacoder::filter_taxa(obj_ITS, taxon_names != "NA", reassign_obs = FALSE)
# 
# 
# obj_ITS %>%
#   metacoder::filter_taxa(taxon_ranks == "Species", supertaxa = TRUE, reassign_obs = FALSE) %>%
#   heat_tree_matrix(data = "diff_table_deseq",
#                    node_label = taxon_names,
#                    node_size = n_obs, # number of OTUs
#                    node_color = log2FoldChange, # difference between groups
#                    node_color_trans = "linear",
#                    node_color_interval = c(-28, 28), # symmetric interval
#                    edge_color_interval = c(-28, 28), # symmetric interval
#                    node_color_range = diverging_palette(), # diverging colors
#                    node_size_axis_label = "ASV count",
#                    node_color_axis_label = "Log 2 Fold Change",
#                    layout = "da", initial_layout = "re",
#                    key_size = 0.4,
#                    output_file = "Figures/SOD/ITS/diff_abundance_heattree_deseq_sod_ITS_firemod.pdf") # The layout algorithm that initializes node locations
```

### We'll then do DESEQ analysis on YearSinceTrt_BR
```{r, diff abundance trees-deseq-yrsincetrt_BR}
# set.seed(1)
# obj_ITS$data$diff_table_deseq <- metacoder::calc_diff_abund_deseq2(obj_ITS,
#                                       data = "tax_abund",
#                                       cols = obj_ITS$data$sample_data$sample_id, # What columns of sample data to
#                                       groups = obj_ITS$data$sample_data$YrSinceTrt_BR) # What category each sample is assigned to
# 
# obj_ITS$data$diff_table_deseq$log2FoldChange[obj_ITS$data$diff_table_deseq$padj > 0.05] <- 0
# 
# range(obj_ITS$obj_ITS$data$diff_table_deseq$padj, finite = TRUE) 
# range(obj_ITS$data$diff_table_deseq$log2FoldChange, finite = TRUE) 
# 
# obj_ITS<- metacoder::filter_taxa(obj_ITS, taxon_names != "None", reassign_obs = FALSE)
# obj_ITS <- metacoder::filter_taxa(obj_ITS, taxon_names != "", reassign_obs = FALSE)
# obj_ITS <- metacoder::filter_taxa(obj_ITS, taxon_names != "NA", reassign_obs = FALSE)
# 
# 
# obj_ITS %>%
#   metacoder::filter_taxa(taxon_ranks == "Species", supertaxa = TRUE, reassign_obs = FALSE) %>%
#   heat_tree_matrix(data = "diff_table_deseq",
#                    node_label = taxon_names,
#                    node_size = n_obs, # number of OTUs
#                    node_color = log2FoldChange, # difference between groups
#                    node_color_trans = "linear",
#                    node_color_interval = c(-38, 38), # symmetric interval
#                    edge_color_interval = c(-38, 38), # symmetric interval
#                    node_color_range = diverging_palette(), # diverging colors
#                    node_size_axis_label = "ASV count",
#                    node_color_axis_label = "Log 2 Fold Change",
#                    layout = "da", initial_layout = "re",
#                    key_size = 0.4,
#                    output_file = "Figures/SOD/ITS/diff_abundance_heattree_deseq_sod_ITS_YrsincetrtBr.pdf") # The layout algorithm that initializes node locations
```

### Software used
```{r software used}
sessioninfo::session_info()
```