---
title: "ITS SOD Experiments"
output: html_document
date: "2024-05-21"
---

```{r setup, include=TRUE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(root.dir = "~/Documents/Postdoc/SOD_microbiome_analysis")
library(phyloseq); packageVersion("phyloseq")
library(ggplot2); packageVersion("ggplot2")
library(readr); packageVersion("readr")
library(purrr); packageVersion("purrr")
library(furrr); packageVersion("furrr")
library(dplyr); packageVersion("dplyr")
library(stringr); packageVersion("stringr")
library(metacoder); packageVersion("metacoder")
library(Biostrings); packageVersion("Biostrings")
library(vegan);packageVersion("vegan")
library(agricolae);packageVersion("agricolae")
library(magrittr);packageVersion("magrittr")
library(data.table);packageVersion("data.table")
library(ggbeeswarm);packageVersion("ggbeeswarm")
library(DESeq2);packageVersion("DESeq2")
library(cowplot);packageVersion("cowplot")
library(doParallel);packageVersion("doParallel")
library(viridis);packageVersion("viridis")
library(ggpubr); packageVersion("ggpubr")
theme_set(theme_cowplot())
library(geosphere); packageVersion("geosphere")
```

### Load data and filter out wildfire samples
```{r subset data}
load("Data/ITS/filtered_data/sodexp.RData")
load("Data/ITS/filtered_data/sod_5asv.RData")

filtered_read_counts<-read.csv("Data/ITS/filtered_data/ITS_subset_track_reads_sod.csv")
summary(filtered_read_counts)

new_names <- sapply(strsplit(sod_5asv.ps@sam_data$Sample_nameITS, "-"), function(x) {
  substr(x[1], 3, nchar(x[1]))
})

sod.ps@sam_data$Sample_nameITS <- new_names
sod_5asv.ps@sam_data$Sample_nameITS <- new_names
sod_nowildfire.ps <- subset_samples(sod.ps, Fire_Mod != "wildfire")
sod_5asv_nowildfire.ps <- subset_samples(sod_5asv.ps, Fire_Mod != "wildfire")

save(sod_nowildfire.ps, file="Data/ITS/filtered_data/sodexp_nowildfire.RData")
save(sod_5asv_nowildfire.ps, file="Data/ITS/filtered_data/sod_5asv_nowildfire.RData")


sod_nowildfire.ps@sam_data$Treatuntreat_EP <- factor(sod_nowildfire.ps@sam_data$Treatuntreat_EP, 
                                            levels = c("untreated", "treated"), 
                                                  labels=c("Untreated", "Treated"))

sod_nowildfire.ps@sam_data$YrSinceTrt_BR <- factor(sod_nowildfire.ps@sam_data$YrSinceTrt_BR, 
                                            levels = c("untreated", "1to2", "3to4", "5to8", "10to12"), 
                                                  labels=c("Untreated", "1 to 2", "3 to 4", "5 to 8", "10 to 12"))

sod_5asv_nowildfire.ps@sam_data$Treatuntreat_EP <- factor(sod_5asv_nowildfire.ps@sam_data$Treatuntreat_EP, 
                                            levels = c("untreated", "treated"), 
                                                  labels=c("Untreated", "Treated"))

sod_5asv_nowildfire.ps@sam_data$YrSinceTrt_BR <- factor(sod_5asv_nowildfire.ps@sam_data$YrSinceTrt_BR, 
                                            levels = c("untreated", "1to2", "3to4", "5to8", "10to12"), 
                                                  labels=c("Untreated", "1 to 2", "3 to 4", "5 to 8", "10 to 12"))
```

### Bar plot agglomerated by species-using filtered dataset and facet wrapped by treatment
```{r barplot-Relative abundance-treatment-order}
dataITS <- sod_5asv_nowildfire.ps %>%
  transform_sample_counts(function(x) {x / sum(x)}) %>% 
  tax_glom(taxrank = "Phylum") %>%   
  psmelt() %>% 
  #mutate(across(Kingdom:Phylum ~ ifelse(is.na(.), "Unclassified", .))) %>%  
  arrange(desc(Abundance))

abund_plot <- ggplot(dataITS, aes(x = Sample_nameITS, y = Abundance, fill = Phylum)) + 
  geom_bar(stat = "identity", position = "stack", color = "black", size = 0.2) +
  scale_fill_viridis(option= "H", discrete = TRUE) +
  scale_color_viridis(option="H", discrete = TRUE) +
  theme_minimal() +
  labs(
    y = "Relative Abundance",
    fill = "Phylum"
  ) +
  facet_grid(~factor(Treatuntreat_EP), scales = "free_x") +
  theme(
    axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5, size = 16),
    axis.text.y = element_text(size = 20),  # Increase y-axis label size
    axis.title.x = element_blank(),  # Remove the x-axis title if needed
    axis.title.y = element_text(size = 20), 
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    legend.position = "bottom",
    legend.text = element_text(size = 20),
    legend.title = element_text(size = 20, face = "bold"),
    legend.key.size = unit(1, "lines"),
    strip.text = element_text(size = 24, face= "bold"),
    strip.background = element_blank()
  ) +
  guides(
    fill = guide_legend(
      reverse = FALSE,
      keywidth = 1,
      keyheight = 1,
      title.position = "top",
      title.hjust = 0.5,
      label.theme = element_text(size = 20)
    )
  )

abund_plot <- abund_plot + 
  scale_x_discrete(labels = function(x) gsub("Sample_nameITS", "", x))  # Remove unwanted text

abund_plot

# Save the plot
ggsave("Figures/ITS/relabundance_plot_treatment_relabund_phylum_sod_nowildfire.pdf", abund_plot, width = 24, height = 10)
```

### Make a similar plot, but color by order instead
```{r barplot-Relative abundance-treatment-Order}
dataITS <- sod_5asv_nowildfire.ps %>%
  tax_glom(taxrank = "Order") %>% 
  transform_sample_counts(function(x) {x / sum(x)}) %>% 
  psmelt() %>% 
  #mutate(across(Kingdom:Species, ~ ifelse(is.na(.), "Unclassified", .))) %>%  
  arrange(desc(Abundance))

abund_plot <- ggplot(dataITS, aes(x = Sample_nameITS, y = Abundance, fill = Order)) + 
  geom_bar(stat = "identity", position = "stack", color = "black", size = 0.2) +
  scale_fill_viridis(option= "H", discrete = TRUE) +
  scale_color_viridis(option="H", discrete = TRUE) +
  theme_minimal() +
  labs(
    y = "Relative Abundance",
    fill = "Order"
  ) +
  facet_grid(~factor(Treatuntreat_EP), scales = "free_x") +
  theme(
    axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5, size = 16),
    axis.text.y = element_text(size = 20),  # Increase y-axis label size
    axis.title.x = element_blank(),  # Remove the x-axis title if needed
    axis.title.y = element_text(size = 20), 
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    legend.position = "bottom",
    legend.text = element_text(size = 20),
    legend.title = element_text(size = 20, face = "bold"),
    legend.key.size = unit(1, "lines"),
    strip.text = element_text(size = 24, face= "bold"),
    strip.background = element_blank()
  ) +
  guides(
    fill = guide_legend(
      reverse = FALSE,
      keywidth = 1,
      keyheight = 1,
      title.position = "top",
      title.hjust = 0.5,
      label.theme = element_text(size = 20)
    )
  )

abund_plot <- abund_plot + 
  scale_x_discrete(labels = function(x) gsub("Sample_nameITS", "", x))  # Remove unwanted text

abund_plot

# Save the plot
ggsave("Figures/ITS/relabundance_plot_treatment_relabund_order_sod_nowildfire.pdf", abund_plot, width = 24, height =16)
```

### Let's summarize with boxplot-treatment
```{r, boxplot summary}
dataITS <- sod_5asv_nowildfire.ps %>%
  tax_glom(taxrank = "Phylum") %>% 
  transform_sample_counts(function(x) {x / sum(x)}) %>% 
  psmelt() %>% 
  filter(Abundance > 0.05)

combined_plot <- ggplot(dataITS, aes(x = Treatuntreat_EP, y = Abundance, fill = Phylum)) +
  geom_boxplot(outlier.shape = NA, alpha = 0.5) +  # Boxplot with no outliers and some transparency
  geom_jitter(color = "black", width = 0.2, size = 2) +  # Jittered black dots
  scale_fill_viridis_d(option="H") +
  theme_minimal() +
  labs(
    y = "Relative Abundance"
  ) +
  theme(
    axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5, size = 20, face = "bold"),
    axis.text.y = element_text(size = 24),
    axis.title.x = element_blank(),
    axis.title.y = element_text(size = 24),
    legend.position = "none",  # Remove the legend
    strip.text = element_text(size = 24, face= "bold"),
    strip.background = element_blank()
  ) +
  facet_wrap(~ Phylum, scales = "free_y") +
  stat_compare_means(method = "anova", label.y = 40)

# Save the plot
ggsave("Figures/ITS/relabundance_plot_treatment_combined_facet_boxplot_phylum_nowildfire.pdf", combined_plot, width = 30, height = 16)
```

### Bar plot agglomerated by order-using filtered dataset and faceted by yr since trt
```{r barplot-Relative abundance-yrtrt}
dataITS <- sod_5asv_nowildfire.ps %>%
  transform_sample_counts(function(x) {x / sum(x)}) %>% 
  tax_glom(taxrank = "Order") %>% 
  psmelt() %>% 
  #mutate(across(Kingdom:Species, ~ ifelse(is.na(.), "Unclassified", .))) %>%  
  arrange(desc(Abundance))

abund_plot <- ggplot(dataITS, aes(x = Sample_nameITS, y = Abundance, fill = Order)) + 
  geom_bar(stat = "identity", position = "stack", color = "black", size = 0.2) +
  scale_fill_viridis(option= "H", discrete = TRUE) +
  scale_color_viridis(option="H", discrete = TRUE) +
  theme_minimal() +
  labs(
    y = "Relative Abundance",
    fill = "Order"
  ) +
  facet_grid(~factor(YrSinceTrt_BR), scales = "free_x") +
  theme(
    axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5, size = 16),
    axis.text.y = element_text(size = 20),  # Increase y-axis label size
    axis.title.x = element_blank(),  # Remove the x-axis title if needed
    axis.title.y = element_text(size = 20), 
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    legend.position = "bottom",
    legend.text = element_text(size = 20),
    legend.title = element_text(size = 20, face = "bold"),
    legend.key.size = unit(1, "lines"),
    strip.text = element_text(size = 24, face= "bold"),
    strip.background = element_blank()
  ) +
  guides(
    fill = guide_legend(
      reverse = FALSE,
      keywidth = 1,
      keyheight = 1,
      title.position = "top",
      title.hjust = 0.5,
      label.theme = element_text(size = 20)
    )
  )

abund_plot <- abund_plot + 
  scale_x_discrete(labels = function(x) gsub("Sample_nameITS", "", x))  # Remove unwanted text

abund_plot

# Save the plot
ggsave("Figures/ITS/relabundance_plot_yrsincetrt_relabund_order_sod_nowildfire.pdf", abund_plot, width = 30, height = 16)
```

### Let's summarize with boxplot-year since trt
```{r, boxplot summary 2}
dataITS <- sod_5asv_nowildfire.ps %>%
  tax_glom(taxrank = "Phylum") %>% 
  transform_sample_counts(function(x) {x / sum(x)}) %>% 
  psmelt() %>% 
  filter(Abundance > 0.05)

combined_plot <- ggplot(dataITS, aes(x = YrSinceTrt_BR, y = Abundance, fill = Phylum)) +
  geom_boxplot(outlier.shape = NA, alpha = 0.5) +  # Boxplot with no outliers and some transparency
  geom_jitter(color = "black", width = 0.2, size = 2) +  # Jittered black dots
  scale_fill_viridis_d(option="H") +
  theme_minimal() +
  labs(
    y = "Relative Abundance"
  ) +
  theme(
    axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5, size = 20, face = "bold"),
    axis.text.y = element_text(size = 24),
    axis.title.x = element_blank(),
    axis.title.y = element_text(size = 24),
    legend.position = "none",  # Remove the legend
    strip.text = element_text(size = 24, face= "bold"),
    strip.background = element_blank()
  ) +
  facet_wrap(~ Phylum, scales = "free_y") +
  stat_compare_means(method = "anova", label.y = 40)

ggsave("Figures/ITS/relabundance_plot_yrsincetrt_combined_facet_boxplot_phylum_nowildfire.pdf", combined_plot, width = 30, height = 16)
```


### Let's look top taxa at order level
```{r Top taxa order}
top40 <- names(sort(taxa_sums(sod_5asv_nowildfire.ps), decreasing=TRUE))[1:40]
ps.top40 <- transform_sample_counts(sod_5asv_nowildfire.ps, function(OTU) OTU/sum(OTU))
#ps.top40@tax_table[is.na(ps.top40@tax_table)] <- "Unclassified"
ps.top40 <- tax_glom(ps.top40, taxrank = "Order")
ps.top40 <- prune_taxa(top40, ps.top40)

pdf("Figures/ITS/top40_genus_sod_nowildfire.pdf", width = 34, height = 20)
plot_bar(ps.top40, x="Sample_nameITS", fill="Order") + facet_wrap(~Treatuntreat_EP, scales="free_x")
dev.off()
```

### Let's look top taxa at family level
```{r Top taxa family}
top40 <- names(sort(taxa_sums(sod_5asv_nowildfire.ps), decreasing=TRUE))[1:40]
ps.top40 <- transform_sample_counts(sod_5asv_nowildfire.ps, function(OTU) OTU/sum(OTU))
#ps.top40@tax_table[is.na(ps.top40@tax_table)] <- "Unclassified"
ps.top40 <- tax_glom(ps.top40, taxrank = "Family")
ps.top40 <- prune_taxa(top40, ps.top40)

pdf("Figures/ITS/top40_family_sod_nowildfire.pdf", width = 34, height = 20)
plot_bar(ps.top40, x="Sample_nameITS", fill="Family") + facet_wrap(~Treatuntreat_EP, scales="free_x")
dev.off()
```

### Let's look top taxa at genus level
```{r Top taxa genus}
top40 <- names(sort(taxa_sums(sod_5asv_nowildfire.ps), decreasing=TRUE))[1:40]
ps.top40 <- transform_sample_counts(sod_5asv_nowildfire.ps, function(OTU) OTU/sum(OTU))
#ps.top40@tax_table[is.na(ps.top40@tax_table)] <- "Unclassified"
ps.top40 <- tax_glom(ps.top40, taxrank = "Genus")
ps.top40 <- prune_taxa(top40, ps.top40)

pdf("Figures/ITS/top40_genus_sod_nowildfire.pdf", width = 34, height = 20)
plot_bar(ps.top40, x="Sample_nameITS", fill="Genus") + facet_wrap(~Treatuntreat_EP, scales="free_x")
dev.off()
```

```{r make data tables-top taxa SOD treatment}
data_raw <-
  sod_5asv_nowildfire.ps %>%
  tax_glom("Phylum") %>%
  psmelt() %>%
  as_tibble()

data_summed_raw <- data_raw %>%
  filter(Abundance > 0) %>%
  group_by(Treatuntreat_EP, Kingdom, Phylum) %>%
  summarise(Raw_Abundance = sum(Abundance),
            Prevalence = n_distinct(Sample)) %>%
  ungroup()

data_rel <-
  sod_5asv_nowildfire.ps %>%
  tax_glom("Phylum") %>%
  transform_sample_counts(function(x) {x/sum(x)}) %>%
  psmelt() %>%
  as_tibble()

data_summed_rel <- data_rel %>%
  filter(Abundance > 0) %>%
  group_by(Treatuntreat_EP, Kingdom, Phylum) %>%
  summarise(Relative_Abundance = sum(Abundance)) %>%
  ungroup() %>%
  group_by(Treatuntreat_EP) %>%
  mutate(Percentage = 100 * (Relative_Abundance / sum(Relative_Abundance))) %>%
  ungroup()

data_summed <- data_summed_raw %>%
  left_join(data_summed_rel, by = c("Treatuntreat_EP", "Kingdom", "Phylum")) %>%
  arrange(Treatuntreat_EP, desc(Percentage))  # Sort by treatment, then percentage descending
  
write.csv(data_summed, file="Results/ITS/mostabundanttaxa_abundance_sod_phylum_treatment_nowildfire.csv", row.names = TRUE)

data_raw1 <-
  sod_5asv_nowildfire.ps %>%
  tax_glom("Genus") %>%
  psmelt() %>%
  as_tibble()

data_summed_raw1 <- data_raw1 %>%
  filter(Abundance > 0) %>%
  group_by(Treatuntreat_EP, Kingdom, Phylum, Class, Order, Family, Genus) %>%
  summarise(Raw_Abundance = sum(Abundance),
            Prevalence = n_distinct(Sample)) %>%
  ungroup()

data_rel1 <-
  sod_5asv_nowildfire.ps %>%
  tax_glom("Genus") %>%
  transform_sample_counts(function(x) {x/sum(x)}) %>%
  psmelt() %>%
  as_tibble()

data_summed_rel1 <- data_rel1 %>%
  filter(Abundance > 0) %>%
  group_by(Treatuntreat_EP, Kingdom, Phylum, Class, Order, Family, Genus) %>%
  summarise(Relative_Abundance = sum(Abundance)) %>%
  ungroup() %>%
  group_by(Treatuntreat_EP) %>%
  mutate(Percentage = 100 * (Relative_Abundance / sum(Relative_Abundance))) %>%
  ungroup()

data_summed1 <- data_summed_raw1 %>%
  left_join(data_summed_rel1, by = c("Treatuntreat_EP", "Kingdom", "Phylum", "Class", "Order", "Family", "Genus")) %>%
  arrange(Treatuntreat_EP, desc(Percentage))  # Sort
  
write.csv(data_summed1, file="Results/ITS/mostabundanttaxa_abundance_sod_genus_treatment_nowildfire.csv", row.names = TRUE)

data_raw2 <-
  sod_5asv_nowildfire.ps %>%
  tax_glom("Species") %>%
  psmelt() %>%
  as_tibble()

data_summed_raw2 <- data_raw2 %>%
  filter(Abundance > 0) %>%
  group_by(Treatuntreat_EP, Kingdom, Phylum, Class, Order, Family, Genus, Species) %>%
  summarise(Raw_Abundance = sum(Abundance),
            Prevalence = n_distinct(Sample)) %>%
  ungroup()

data_rel2 <-
  sod_5asv_nowildfire.ps %>%
  tax_glom("Species") %>%
  transform_sample_counts(function(x) {x/sum(x)}) %>%
  psmelt() %>%
  as_tibble()

data_summed_rel2 <- data_rel2 %>%
  filter(Abundance > 0) %>%
  group_by(Treatuntreat_EP, Kingdom, Phylum, Class, Order, Family, Genus, Species) %>%
  summarise(Relative_Abundance = sum(Abundance)) %>%
  ungroup() %>%
  group_by(Treatuntreat_EP) %>%
  mutate(Percentage = 100 * (Relative_Abundance / sum(Relative_Abundance))) %>%
  ungroup()

data_summed2 <- data_summed_raw2 %>%
  left_join(data_summed_rel2, by = c("Treatuntreat_EP", "Kingdom", "Phylum", "Class", "Order", "Family", "Genus", "Species")) %>%
  arrange(Treatuntreat_EP, desc(Percentage))  # Sort
  
write.csv(data_summed2, file="Results/ITS/mostabundanttaxa_abundance_sod_species_treatment_nowildfire.csv", row.names = TRUE)
```

### Let's make a table of top taxa-faceted by Years Since SOD treatment
```{r make data tables-top taxa yrs since trt}
data_raw3 <-
  sod_5asv_nowildfire.ps %>%
  tax_glom("Phylum") %>%
  psmelt() %>%
  as_tibble()

data_summed_raw3 <- data_raw3 %>%
  filter(Abundance > 0) %>%
  group_by(YrSinceTrt_BR, Kingdom, Phylum) %>%
  summarise(Raw_Abundance = sum(Abundance),
            Prevalence = n_distinct(Sample)) %>%
  ungroup()

data_rel3 <-
  sod_5asv_nowildfire.ps %>%
  tax_glom("Phylum") %>%
  transform_sample_counts(function(x) {x/sum(x)}) %>%
  psmelt() %>%
  as_tibble()

data_summed_rel3 <- data_rel3 %>%
  filter(Abundance > 0) %>%
  group_by(YrSinceTrt_BR, Kingdom, Phylum) %>%
  summarise(Relative_Abundance = sum(Abundance)) %>%
  ungroup() %>%
  group_by(YrSinceTrt_BR) %>%
  mutate(Percentage = 100 * (Relative_Abundance / sum(Relative_Abundance))) %>%
  ungroup()

data_summed3 <- data_summed_raw3 %>%
  left_join(data_summed_rel3, by = c("YrSinceTrt_BR", "Kingdom", "Phylum")) %>%
  arrange(YrSinceTrt_BR, desc(Percentage))  # Sort by years, then percentage descending
  
write.csv(data_summed3, file="Results/ITS/mostabundanttaxa_abundance_sod_phylum_yrsincetrt_nowildfire.csv", row.names = TRUE)

data_raw4 <-
  sod_5asv_nowildfire.ps %>%
  tax_glom("Genus") %>%
  psmelt() %>%
  as_tibble()

data_summed_raw4 <- data_raw4 %>%
  filter(Abundance > 0) %>%
  group_by(YrSinceTrt_BR, Kingdom, Phylum, Class, Order, Family, Genus) %>%
  summarise(Raw_Abundance = sum(Abundance),
            Prevalence = n_distinct(Sample)) %>%
  ungroup()

data_rel4 <-
  sod_5asv_nowildfire.ps %>%
  tax_glom("Genus") %>%
  transform_sample_counts(function(x) {x/sum(x)}) %>%
  psmelt() %>%
  as_tibble()

data_summed_rel4 <- data_rel4 %>%
  filter(Abundance > 0) %>%
  group_by(YrSinceTrt_BR, Kingdom, Phylum, Class, Order, Family, Genus) %>%
  summarise(Relative_Abundance = sum(Abundance)) %>%
  ungroup() %>%
  group_by(YrSinceTrt_BR) %>%
  mutate(Percentage = 100 * (Relative_Abundance / sum(Relative_Abundance))) %>%
  ungroup()

data_summed4 <- data_summed_raw4 %>%
  left_join(data_summed_rel4, by = c("YrSinceTrt_BR", "Kingdom", "Phylum", "Class", "Order", "Family", "Genus")) %>%
  arrange(YrSinceTrt_BR, desc(Percentage))  # Sort
  
write.csv(data_summed4, file="Results/ITS/mostabundanttaxa_abundance_sod_genus_yrsincetrt_nowildfire.csv", row.names = TRUE)

data_raw5 <-
  sod_5asv_nowildfire.ps %>%
  tax_glom("Species") %>%
  psmelt() %>%
  as_tibble()

data_summed_raw5 <- data_raw5 %>%
  filter(Abundance > 0) %>%
  group_by(YrSinceTrt_BR, Kingdom, Phylum, Class, Order, Family, Genus, Species) %>%
  summarise(Raw_Abundance = sum(Abundance),
            Prevalence = n_distinct(Sample)) %>%
  ungroup()

data_rel5 <-
  sod_5asv_nowildfire.ps %>%
  tax_glom("Species") %>%
  transform_sample_counts(function(x) {x/sum(x)}) %>%
  psmelt() %>%
  as_tibble()

data_summed_rel5 <- data_rel5 %>%
  filter(Abundance > 0) %>%
  group_by(YrSinceTrt_BR, Kingdom, Phylum, Class, Order, Family, Genus, Species) %>%
  summarise(Relative_Abundance = sum(Abundance)) %>%
  ungroup() %>%
  group_by(YrSinceTrt_BR) %>%
  mutate(Percentage = 100 * (Relative_Abundance / sum(Relative_Abundance))) %>%
  ungroup()

data_summed5 <- data_summed_raw5 %>%
  left_join(data_summed_rel5, by = c("YrSinceTrt_BR", "Kingdom", "Phylum", "Class", "Order", "Family", "Genus", "Species")) %>%
  arrange(YrSinceTrt_BR, desc(Percentage))  # Sort
  
write.csv(data_summed5, file="Results/ITS/mostabundanttaxa_abundance_sod_species_yrsincetrt_nowildfire.csv", row.names = TRUE)
```

### Let's look at spacial placement of different plots
```{r Map exploration}
pdf("Figures/ITS/location_of_plots_nowildfire.pdf", width = 20, height = 12)
map_plot <- ggplot(sod_5asv_nowildfire.ps@sam_data, aes(x = Longitude, y = Latitude)) +
  geom_text(aes(color = Fire_Mod, label = Sample_nameITS), size = 2.5, 
            check_overlap = FALSE) +
  labs(x = "Longitude", y = "Latitude", title = "Sampling Locations") +
  theme_minimal() +
  theme(
    plot.title = element_text(size = 20), 
    axis.title = element_text(size = 16),
    axis.text = element_text(color = "black"),
    legend.title = element_text(size = 14),
    legend.text = element_text(size = 12)
  )
print(map_plot)
dev.off()
```

### Let's normalize our data in preparation for beta diversity analyses
```{r Normalization}
#sod_5asv_nowildfire.ps<- phyloseq::prune_taxa(taxa_sums(sod_nowildfire.ps) > 5, sod_nowildfire.ps)
sdt = data.table::data.table(as(sample_data(sod_5asv_nowildfire.ps), "data.frame"),
                 TotalReads = sample_sums(sod_5asv_nowildfire.ps), keep.rownames = TRUE)
min_reads<-min(sdt$TotalReads)

rarefied.ps <- rarefy_even_depth(sod_5asv_nowildfire.ps, sample.size = min_reads, rngseed = 100)

strat_rel.ps<-transform_sample_counts(sod_5asv_nowildfire.ps, function(x) x / sum(x))
```

### Alpha diversity-use non-normalized or pruned data
Let's start with t-test as point of comparison with anova analysis
```{r alpha diversity analyses-t-test}
set.seed(1)
AD <- estimate_richness(sod_nowildfire.ps, measures = c("Observed", "InvSimpson", "Shannon"))
rownames(AD) <- gsub("\\.", "-", rownames(AD))

AD <- cbind(AD, sample_data(sod_nowildfire.ps)[row.names(AD), c("Treatuntreat_EP")])

AD <- data.table(AD, keep.rownames = TRUE)

AD <- melt(AD, id.vars = c('rn', c("Treatuntreat_EP")))

factors <- c("Treatuntreat_EP")
ttest_list <- list()

for (factor_name in factors) {
  ttest <- data.table(alpha_index = character(), pvalue = numeric())
  
  for (alphad in c("InvSimpson", "Shannon", "Observed")) {
    ttest_res <- t.test(value ~ get(factor_name), data = AD[variable == alphad], var.equal = TRUE)
    p_value <- ttest_res$p.value
    ttest <- rbind(ttest, data.table(alpha_index = alphad, pvalue = p_value))
  }
  
  ttest_list[[factor_name]] <- ttest
}

for (factor_name in factors) {
  cat(paste("T-Test Results for Factor:", factor_name), "\n")
  print(knitr::kable(ttest_list[[factor_name]], format = "markdown"))
  cat("\n")
}
``` 

### Let's now proceed with ANOVA analyses and then compile our alpha diversity stats
```{r compiled alpha diversity table}
set.seed(1)
anova_list <- list()

AD <- estimate_richness(sod_nowildfire.ps, measures = c("Observed", "InvSimpson", "Shannon"))
rownames(AD) <- gsub("\\.", "-", rownames(AD))

AD <- cbind(AD, sample_data(sod_nowildfire.ps)[row.names(AD), c("Treatuntreat_EP", "YrSinceTrt_BR", "YrSinceTrt_Range")])

AD <- data.table(AD, keep.rownames = TRUE)

AD <- melt(AD, id.vars = c('rn', c("Treatuntreat_EP", "YrSinceTrt_BR", "YrSinceTrt_Range")))

factors2 <- c("Treatuntreat_EP", "YrSinceTrt_BR", "YrSinceTrt_Range")

plotted_factors <- character()
 
final_anova_table <- data.table(factor_name = character(), alpha_index = character(), pvalue = numeric(), df_between = numeric(), df_within = numeric(), ss_between = numeric(), ss_within = numeric(), ms_between = numeric(), ms_within = numeric(), F_value = numeric())

for (factor_name in factors2) {
  anova <- data.table(factor_name = character(), alpha_index = character(), pvalue = numeric(), df_between = numeric(), df_within = numeric(), ss_between = numeric(), ss_within = numeric(), ms_between = numeric(), ms_within = numeric(), F_value = numeric())
  
  for (alphad in c("Observed","InvSimpson", "Shannon")) {
    anova_res <- aov(value ~ get(factor_name), data = AD[variable == alphad])
    anova_summary <- summary(anova_res)
    p_value <- anova_summary[[1]]$`Pr(>F)`[1]
    df_between <- anova_summary[[1]]$"Df"[1]
    df_within <- anova_summary[[1]]$"Df"[2]
    ss_between <- anova_summary[[1]]$"Sum Sq"[1]
    ss_within <- anova_summary[[1]]$"Sum Sq"[2]
    ms_between <- anova_summary[[1]]$"Mean Sq"[1]
    ms_within <- anova_summary[[1]]$"Mean Sq"[2]
    F_value <- anova_summary[[1]]$"F value"[1]
    
    anova <- rbind(anova, data.table(factor_name = factor_name, alpha_index = alphad, pvalue = p_value, df_between = df_between, df_within = df_within, ss_between = ss_between, ss_within = ss_within, ms_between = ms_between, ms_within = ms_within, F_value = F_value))
  }
  
  final_anova_table <- rbind(final_anova_table, anova)
}

print(knitr::kable(final_anova_table, format = "markdown"))
write.csv(final_anova_table, "Results/ITS/final_anova_table_sod_nowildfire.csv", row.names = FALSE)
```

### Let's make faceted boxplots for our alpha diversity analyses
```{r alpha diversity faceted plots}
set.seed(1)
AD <- estimate_richness(sod_nowildfire.ps, measures = c("Observed", "InvSimpson", "Shannon"))
rownames(AD) <- gsub("\\.", "-", rownames(AD))

AD <- cbind(AD, sample_data(sod_nowildfire.ps)[rownames(AD), c("Treatuntreat_EP", "YrSinceTrt_BR", "YrSinceTrt_Range")])

variables <- c("Treatuntreat_EP", "YrSinceTrt_BR", "YrSinceTrt_Range")

plots <- list()

for (variable in variables) {
  p <- plot_richness(sod_nowildfire.ps, x = variable, measures = c("Shannon", "InvSimpson", "Observed")) + 
    geom_boxplot()
  plots[[variable]] <- p
}

plots_arrange <- gridExtra::grid.arrange(
  plots[["Treatuntreat_EP"]], plots[["YrSinceTrt_BR"]],plots[["YrSinceTrt_Range"]],ncol=1)

plots_arrange_with_text <- cowplot::ggdraw() +
  cowplot::draw_plot(plots_arrange) +
  cowplot::draw_plot("Treatuntreat_EP", x = 0.28, y = 0.98) +
  cowplot::draw_plot("YrSinceTrt_BR", x = 0.28, y = 0.98) +
  cowplot::draw_plot("YrSinceTrt_Range", x = 0.28, y = 0.98) 

ggsave(plots_arrange_with_text, path = "Figures/ITS", filename = "alpha_diversity_each_factor_sod_nowildfire.pdf", 
       width = 16, height = 24)
```

### Let's perform TUKEY HSD tests
```{r alpha diversity compilation with tukey summary}
set.seed(1)
variables <- c("Treatuntreat_EP", "YrSinceTrt_BR", "YrSinceTrt_Range")
measures <- c("Observed", "InvSimpson", "Shannon")

tukey_results <- list()

for (variable in variables) {
  for (measure in measures) {
    data_subset <- AD[, c(variable, measure)]
    formula <- as.formula(paste(measure, "~ ."))
    anova_res <- aov(formula, data = data_subset)
    tukey_result <- HSD.test(anova_res, variable)
    print(tukey_result)
  }
}
```

### Let's visualize alpha diversity results with rarefaction curve-treatment
```{r, rarefaction curve trt }
sample_data <- data.frame(sample_data(sod_nowildfire.ps))
otu_data <- as(otu_table(sod_nowildfire.ps), "matrix")

sample_data$Treatuntreat_EP <- factor(sample_data$Treatuntreat_EP, levels = c("Untreated", "Treated"))
colors <- ifelse(sample_data$Treatuntreat_EP == "Untreated", "blue","red")

raremax <- min(rowSums(otu_data))

pdf("Figures/ITS/rarefaction_curve_trt_nowildfire.pdf", width = 10, height = 8)
rarecurve(otu_data, step = 10, sample = raremax, col = colors, cex = 0.6, label = FALSE, xlab = "Sequencing Depth", ylab = "Number of ASVs")
legend("topright", legend = levels(sample_data$Fire_Mod), col = c("blue", "red"), lty = 1, title = "Treatuntreat_EP")
dev.off()
```

### Let's visualize alpha diversity results with rarefaction curve-year since trt
```{r, rarefaction curve yr since trt}
sample_data <- data.frame(sample_data(sod_nowildfire.ps))
otu_data <- as(otu_table(sod_nowildfire.ps), "matrix")

colors <- ifelse(sample_data$YrSinceTrt_BR== "Untreated", "lightblue",
                 ifelse(sample_data$YrSinceTrt_BR == "1 to 2", "red", 
                        ifelse(sample_data$YrSinceTrt_BR == "3 to 4", "green", 
                               ifelse(sample_data$YrSinceTrt_BR == "5 to 8", "orange","purple"))))

raremax <- min(rowSums(otu_data))

pdf("Figures/ITS/rarefaction_curve_yrsincetrt_nowildfire.pdf", width = 10, height = 8)
rarecurve(otu_data, step = 10, sample = raremax, col = colors, cex = 0.6, label = FALSE, xlab = "Sequencing Depth", ylab = "Number of ASVs")
legend("topright", legend = levels(sample_data$YrSinceTrt_BR), col = c("lightblue", "red", "green", "orange", "purple"), lty = 1, title = "YrSinceTrt_BR")
dev.off()
```

### Beta diversity-PERMANOVA analyses-ASV counts transformed (rel. abundance)
```{r PERMANOVA ANALYSES-relative abundance, include=TRUE}
set.seed(1)
mapping <- as.data.frame(sample_data(strat_rel.ps))

mapping2=mapping[, c("Treatuntreat_EP", "YrSinceTrt_BR", "YrSinceTrt_Range")]

ASV_tables_bray <- phyloseq::distance(strat_rel.ps, method = "bray")

df_metadata <- data.frame(sample_data(strat_rel.ps))

df_metadata<-df_metadata[, c("Treatuntreat_EP", "YrSinceTrt_BR","YrSinceTrt_Range")]

pvals <- sapply(colnames(mapping2), function(variable){
  form1<-as.formula(paste("ASV_tables_bray", variable, sep="~"))
  res <- adonis2(form1, data = df_metadata, permutations = 9999)
  pval <- res$`Pr(>F)`[[1]] # we take the first p value which corresponds to the variable of interest
  return(pval)
})

pval_adj <- p.adjust(pvals, method = "fdr")

save.pval.adonis0.05_braycurtis <- pval_adj[pval_adj < 0.05]
save.pval.adonis0.05_braycurtis

print(pval_adj)
```

### Compile PERMANOVA stats-main effects
```{r compile PERMANOVA stats}
set.seed(1)
res <- adonis2(ASV_tables_bray~Treatuntreat_EP+YrSinceTrt_BR, data = df_metadata, permutations = 9999, by="terms")
res_data <- data.frame(res)
colnames(res_data)[colnames(res_data) == "Pr..F."] <- "pr_f"
res_data <- res_data[-c(nrow(res_data)-1, nrow(res_data)),]

res_data$FDR_adj_p <- p.adjust(res_data$pr_f, method = "fdr")
res_data$Bonferroni_adj_p <- p.adjust(res_data$pr_f, method = "bonferroni")
res_data$Factor <- rownames(res_data)

res_data <- res_data[, c("Factor", "Df", "SumOfSqs", "R2", "F", "pr_f", "FDR_adj_p", "Bonferroni_adj_p")]
rownames(res_data) <- NULL

beta_disp_results <- lapply(c("Treatuntreat_EP", "YrSinceTrt_BR", "YrSinceTrt_Range"), function(variable) {
  res_betadisper <- betadisper(d = ASV_tables_bray, group = df_metadata[[variable]])
  perm_test_res <- permutest(res_betadisper)
  pval_betadisper <- perm_test_res$tab$`Pr(>F)`[[1]]
  
  data.frame(Factor = variable, P_Value_BetaDispersion = pval_betadisper)
})

beta_disp_results_df <- do.call(rbind, beta_disp_results)
final_results <- merge(res_data, beta_disp_results_df, by = "Factor", all.x = TRUE)
final_results <- final_results[, c("Factor", "Df", "SumOfSqs", "R2", "F", "pr_f", "FDR_adj_p", "Bonferroni_adj_p", "P_Value_BetaDispersion")]

print(final_results, row.names = FALSE)
write.csv(final_results, file = "Results/ITS/PERMANOVA_with_betadisper_maineffects_nowildfire.csv", row.names = FALSE)
```

### Compile PERMANOVA stats-interactions
```{r compile PERMANOVA stats interaction}
set.seed(1)
res <- adonis2(ASV_tables_bray~Treatuntreat_EP*YrSinceTrt_BR, data = df_metadata, permutations = 9999, by="terms")
res_data <- data.frame(res)
colnames(res_data)[colnames(res_data) == "Pr..F."] <- "pr_f"
res_data <- res_data[-c(nrow(res_data)-1, nrow(res_data)),]

res_data$FDR_adj_p <- p.adjust(res_data$pr_f, method = "fdr")
res_data$Bonferroni_adj_p <- p.adjust(res_data$pr_f, method = "bonferroni")
res_data$Factor <- rownames(res_data)
res_data <- res_data[, c("Factor", "Df", "SumOfSqs", "R2", "F", "pr_f", "FDR_adj_p", "Bonferroni_adj_p")]
rownames(res_data) <- NULL

beta_disp_results <- lapply(c("Treatuntreat_EP", "YrSinceTrt_BR","YrSinceTrt_Range"), function(variable) {
  res_betadisper <- betadisper(d = ASV_tables_bray, group = df_metadata[[variable]])
  perm_test_res <- permutest(res_betadisper)
  pval_betadisper <- perm_test_res$tab$`Pr(>F)`[[1]]
  
  data.frame(Factor = variable, P_Value_BetaDispersion = pval_betadisper)
})

beta_disp_results_df <- do.call(rbind, beta_disp_results)
final_results <- merge(res_data, beta_disp_results_df, by = "Factor", all.x = TRUE)

final_results <- final_results[, c("Factor", "Df", "SumOfSqs", "R2", "F", "pr_f", "FDR_adj_p", "Bonferroni_adj_p", "P_Value_BetaDispersion")]

print(final_results, row.names = FALSE)

write.csv(final_results, file = "Results/ITS/PERMANOVA_with_betadisper_interactions_nowildfire.csv", row.names = FALSE)
```

### Let's now examine beta diversity through ordination plots
We'll will examine outputs for rarefied data as well as data adjusted by relative abundance
We'll now make ordination plots on data for which we've calculated relative abundance
```{r Beta diversity-Rel. abundance-treated and untreated samples}
set.seed(1)
ps_NMDS = ordinate(strat_rel.ps, 
                   method="NMDS", 
                   distance="bray")

plot<- plot_ordination(strat_rel.ps, 
                         ps_NMDS,
                         color= "Treatuntreat_EP") +
  ggtitle("NMDS ordination plot - relative abundance") +
  labs(color = "Factor") + 
  theme(axis.text=element_text(size=14), axis.title=element_text(size=14), title =  element_text(size=13), legend.text = element_text(size = 18))

plot_ellipses<-plot +
   stat_ellipse(type = "norm", linetype = 2) +
  theme_bw()

plot_ellipses

write.csv(ps_NMDS[["points"]], file="Results/ITS/ITS_NMDS_coordinates_persample_nowildfire.csv",row.names = TRUE)
```

### Now, we'll look at ordination plot rel abundance-with YrSinceTrt_BR factor
```{r Beta diversity-Rel. abundance-YrSinceTrt_BR}
set.seed(1)
ps_NMDS = ordinate(strat_rel.ps, 
                   method="NMDS",
                   distance="bray")

plot<- plot_ordination(strat_rel.ps, 
                         ps_NMDS,
                         labs(color = "Factor"),
                         color= "YrSinceTrt_BR") +
  ggtitle("NMDS ordination plot - relative abundance") +
  theme(axis.text=element_text(size=14), axis.title=element_text(size=14), title =  element_text(size=13), legend.text = element_text(size = 18))


plot_ellipses<-plot +
   stat_ellipse(type = "norm", linetype = 2) +
  theme_bw()

plot_ellipses
```

### Now, we'll look at ordination plot rel abundance-with YrSinceTrt_Range factor
```{r Beta diversity-Rel. abundance-YrSinceTrt_Range}
set.seed(1)
ps_NMDS = ordinate(strat_rel.ps, 
                   method="NMDS",
                   distance="bray")

plot<- plot_ordination(strat_rel.ps, 
                         ps_NMDS,
                         labs(color = "Factor"),
                         color= "YrSinceTrt_Range") +
  ggtitle("NMDS ordination plot - relative abundance") +
  theme(axis.text=element_text(size=14), axis.title=element_text(size=14), title =  element_text(size=13), legend.text = element_text(size = 18))


plot_ellipses<-plot +
   stat_ellipse(type = "norm", linetype = 2) +
  theme_bw()

plot_ellipses
```
### Are communities more similar if geographically closeby? Some plots, like wildfire plots are close together, for example
Very preliminary experiment
```{r Beta diversity-Rel. abundance-YearSam_EP}
otu_table <- otu_table(strat_rel.ps)
sample_data <- sample_data(strat_rel.ps)
otu_table_matrix <- t(as(otu_table, "matrix"))
lon_lat <- cbind(sample_data$Longitude, sample_data$Latitude)

geo_dist <- distm(lon_lat, fun = distHaversine)
comm_dist <- vegdist(otu_table_matrix, method = "bray", na.rm = TRUE)

comm_dist_matrix <- as.matrix(comm_dist)

sample_ids <- rownames(sample_data)
sample_ids_in_otu <- rownames(otu_table)

common_samples <- intersect(sample_ids_in_otu, sample_ids)
common_indices <- which(sample_ids %in% common_samples)

geo_dist_aligned <- geo_dist[common_indices, common_indices]
comm_dist_aligned <- comm_dist_matrix[common_indices, common_indices]

print(dim(geo_dist_aligned))
print(dim(comm_dist_aligned))

mantel_result <- mantel(geo_dist_aligned, comm_dist_aligned, method = "spearman", permutations = 999)
print(mantel_result)

otu_table <- otu_table(strat_rel.ps)
sample_data <- sample_data(strat_rel.ps)

otu_table_matrix <- as(otu_table, "matrix")
comm_dist <- vegdist(otu_table_matrix, method = "bray")

nmds_result <- metaMDS(comm_dist, k = 2, trymax = 100)

nmds_scores <- as.data.frame(scores(nmds_result))
nmds_scores$SampleID <- rownames(nmds_scores)

sample_ids <- rownames(sample_data(strat_rel.ps))  
sample_coords <- data.frame(SampleID = sample_ids,
                             Longitude = sample_data(strat_rel.ps)$Longitude,
                             Latitude = sample_data(strat_rel.ps)$Latitude, 
                             Treatuntreat=sample_data(strat_rel.ps)$Treatuntreat_EP,
                             Fire_Mod=sample_data(strat_rel.ps)$Fire_Mod
                            )

nmds_scores <- merge(nmds_scores, sample_coords, by = "SampleID")


nmds_plot <- ggplot(nmds_scores, aes(x = NMDS1, y = NMDS2)) +
  geom_point(aes(color = Latitude, size = Longitude), alpha = 0.7) +  
  labs(title = "NMDS Plot with Latitude and Longitude",
       x = "NMDS Dimension 1",
       y = "NMDS Dimension 2") +
  theme_minimal() +
  scale_color_gradient(low = "blue", high = "red") +  
  scale_size_continuous(range = c(2, 6))  

print(nmds_plot)

nmds_plot <- ggplot(nmds_scores, aes(x = NMDS1, y = NMDS2)) +
  geom_point(aes(color = Latitude, size = Longitude), alpha = 0.7) +
  facet_wrap(~ Fire_Mod) +
  labs(title = "NMDS Plot with Latitude, Longitude, and Treatment",
       x = "NMDS Dimension 1",
       y = "NMDS Dimension 2") +
  theme_minimal() +
  scale_color_gradient(low = "blue", high = "red") +
  scale_size_continuous(range = c(2, 6))

print(nmds_plot)

```

### Software used
```{r software used}
sessioninfo::session_info()
```