---
title: "RPS10 SOD Preparation Steps"
output: html_document
date: "2024-06-11"
---

```{r setup}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(root.dir = "~/SOD_DADS_microbiome")

library(phyloseq); packageVersion("phyloseq")
library(ggplot2); packageVersion("ggplot2")
library(readr); packageVersion("readr")
library(purrr); packageVersion("purrr")
library(furrr); packageVersion("furrr")
library(dplyr); packageVersion("dplyr")
library(stringr); packageVersion("stringr")
library(metacoder); packageVersion("metacoder")
library(data.table); packageVersion("data.table")
library(decontam); packageVersion("decontam")
```

### Load data in order to make phyloseq and taxmap objects
```{r load files, include=TRUE}
#Load DADA2 files
seqtab.nochim <- read.table("Data/RPS10/dada2_outputs/rps10_seqtab_nochim.out",sep="\t",stringsAsFactors=F,header=T)

taxa <- read.table("Data/RPS10/dada2_outputs/rps10_taxa.out",sep="\t",stringsAsFactors=F,header=T)

taxa_filt <- taxa %>%
  mutate(across(starts_with("tax."), ~ ifelse(get(str_replace(cur_column(), "tax.", "boot.")) < 60, NA, .))) %>%
  set_names(str_replace(names(.), "tax.", ""))

samdf <- read.csv("Data/RPS10/metadata_dbs/Metadata_rps10_short_mod.csv")
```

### Exploratory analysis-track reads throughout DADA2 analysis   
Examine number of reads throughout the pipeline
```{r track reads through DADA2 analysis, include=TRUE}
track <- read.table("Data/RPS10/dada2_outputs/rps10_trackreads.out", sep="\t", stringsAsFactors = F, header=T)
```

### Format various dataframes so that we can convert to phyloseq obj
```{r format dataframes, include=TRUE}
# For ease of analysis, convert the DADA2 outputs into a Phyloseq object, and filter out any necessary samples
all(rownames(seqtab.nochim) %in% samdf$Sample_nameRps10)
rownames(samdf) <- samdf$Sample_nameRps10

samdata.ps <- sample_data(samdf)

taxa.ps <- tax_table(as.matrix(taxa_filt))

seqtab.nochim.ps <- otu_table(seqtab.nochim, taxa_are_rows=FALSE)

ps <- phyloseq(seqtab.nochim.ps, samdata.ps, taxa.ps)

#Add refseq slot
dna <- Biostrings::DNAStringSet(taxa_names(ps))
names(dna) <- taxa_names(ps)
ps <- merge_phyloseq(ps, dna)
taxa_names(ps) <- paste0("ASV", seq(ntaxa(ps)))
```


### Decontamination step using decontam package-this was most necessary for RPS10 analysis.
```{r decontam, include=TRUE}
#Inspect library sizes
df <- as.data.frame(sample_data(ps)) # Put sample_data into a ggplot-friendly data.frame
df$LibrarySize <- sample_sums(ps)
df <- df[order(df$LibrarySize),]
df$Index <- seq(nrow(df))
ggplot(data=df, aes(x=Index, y=LibrarySize, color=Sample_or_Control)) + geom_point()


sample_data(ps)$is.neg <- sample_data(ps)$Sample_or_Control == "Control"
contamdf.prev <- isContaminant(ps, method="prevalence", neg="is.neg")
table(contamdf.prev$contaminant)
head(which(contamdf.prev$contaminant))

contamdf.prev05 <- isContaminant(ps, method="prevalence", neg="is.neg", threshold=0.5)
table(contamdf.prev05$contaminant)
which(contamdf.prev05$contaminant)

# There don't appear to be any contaminants so we won't proceed with additional filtering described in decontam examples. 

#Let's save 
ps.pa <- transform_sample_counts(ps, function(abund) 1*(abund>0))
ps.pa.neg <- prune_samples(sample_data(ps.pa)$Sample_or_Control == "control", ps.pa)
ps.pa.pos <- prune_samples(sample_data(ps.pa)$Sample_or_Control == "sample", ps.pa)

# Make data.frame of prevalence in positive and negative samples
df.pa <- data.frame(pa.pos=taxa_sums(ps.pa.pos), pa.neg=taxa_sums(ps.pa.neg),
                     contaminant=contamdf.prev05$contaminant)
ggplot(data=df.pa, aes(x=pa.neg, y=pa.pos, color=contaminant)) + geom_point() +
  xlab("Prevalence (Negative Controls)") + ylab("Prevalence (True Samples)")

# Remove contaminants
ps.decontam <- prune_taxa(!contamdf.prev05$contaminant, ps)
save(ps.decontam, file = "Data/RPS10/filtered_data/RPS10_phyloseq.decontam.RData")
```


###Take a first glance at the data  
Helpful for deciding how to filter data. Prevalence is how many samples have the ASV, and total abundance is how many times ASV is present across all samples. 
```{r examine ASV Prevalance, include=TRUE}
#Prevalence-number of samples in which a taxon appears at least once
#Total abundance
prevdf = apply(X = phyloseq::otu_table(ps.decontam),
               MARGIN = ifelse(taxa_are_rows(ps.decontam), yes = 1, no = 2),
               FUN = function(x){sum(x > 0)})

# Add taxonomy and total read counts to this data.frame
prevdf = data.frame(Prevalence = prevdf,
                    TotalAbundance = taxa_sums(ps.decontam),
                    tax_table(ps.decontam))
```

Now I will focus on sod experiment and prepare dataset for further analysis by subsetting these samples from the rest of the sample data. I will also remove allow abundance ASVs not present in at least one sample and not present at least five times
```{r Subset dataset, include=TRUE}
sod.ps<-subset_samples(ps.decontam, Experiment=="basetree_100plots")
sod.ps<- prune_taxa(taxa_sums(sod.ps) > 0, sod.ps)

readsums<-sort(sample_sums(sod.ps))
median(readsums)

#Let's filter an ASVs not present at least 5 times
sod_5asv.ps <- prune_taxa(taxa_sums(sod.ps) > 5, sod.ps)
sod_5asv.ps<- prune_taxa(taxa_sums(sod_5asv.ps) > 0, sod_5asv.ps)

save(sod.ps, file="Data/RPS10/filtered_data/sodexp.RData")
save(sod_5asv.ps, file="Data/RPS10/filtered_data/sod_5asv.RData")

readsums<-sort(sample_sums(sod.ps))
median(readsums)
summary(readsums)
```



### Check prevalence after filtering
```{r check prevalence dataframes after filterig}
#Double check that the filtering makes sense by making new prevalence dataframe
prevdf_check = apply(X = phyloseq::otu_table(sod_5asv.ps),
               MARGIN = ifelse(taxa_are_rows(sod_5asv.ps), yes = 1, no = 2),
               FUN = function(x){sum(x > 0)})

# Add taxonomy and total read counts to this data.frame
prevdf_check = data.frame(Prevalence = prevdf_check,
                    TotalAbundance = taxa_sums(sod_5asv.ps),
                    tax_table(sod_5asv.ps))

```

### Subset read counts so we are just looking at sod samples
```{r subset tracked reads}
samples_to_subset <- samdf %>%
  filter(Experiment=="basetree_100plots") %>% 
  pull(Sample_nameRps10)
# Filter the track dataframe

#track$sample <- rownames(track)
subset_track_reads <- track[rownames(track) %in% samples_to_subset, ]
write.csv(subset_track_reads, file="Data/RPS10/filtered_data/RPS10_subset_track_reads_sod.csv",row.names = TRUE)
# Print summary of the filtered track dataframe
summary(subset_track_reads)
```

### Now I will check how many ASV sequences were classified as P. ramorum
```{r check which ASVs were assigned as P. ramorum}
tax_table <- as.data.frame(sod.ps@tax_table)

ramorum_taxa <- apply(tax_table(sod.ps), 1, function(x) any(grepl("Phytophthora_ramorum", x, ignore.case = TRUE)))
print(sum(ramorum_taxa))

ramorum_asvs <- rownames(tax_table(sod.ps))[ramorum_taxa]


otu_table <- as.data.frame(as(otu_table(sod.ps), "matrix"))

ramorum_counts <- otu_table[, ramorum_asvs, drop = FALSE]
print(ramorum_counts)
```

```{r Filter out samples with few reads}
### Unfortunately, some samples have very few reads. To proceed with analyses, I'll remove any samples with less than 2500 reads. 
#Unfortunately, a subset of samples have less than 3k reads. I will remove these samples for now. 
sod_prunereads.ps = prune_samples(sample_sums(sod.ps)>=2500, sod.ps)
sod_prunereads.ps <- prune_taxa(taxa_sums(sod_prunereads.ps ) > 0, sod_prunereads.ps)
checkcounts=sort(sample_sums(sod_prunereads.ps))

sod_prunereads.ps@sam_data$Treatuntreat_EP<-factor(sod_prunereads.ps@sam_data$Treatuntreat_EP, levels = c(0, 1), labels = c("untreated", "treated"))
sod_prunereads.ps@sam_data$YearSam_EP <- as.factor(sod_prunereads.ps@sam_data$YearSam_EP)
sod_prunereads.ps@sam_data$YrSinceTrt_BR <- as.factor(sod_prunereads.ps@sam_data$YrSinceTrt_BR)
sod_prunereads.ps@sam_data$YrSinceTrt_BR <- factor(sod_prunereads.ps@sam_data$YrSinceTrt_BR, levels = c("untreated", "1to2", "3to4", "5to8", "10to12"))
sod_prunereads.ps@sam_data$Fire_Mod <- factor(sod_prunereads.ps@sam_data$Fire_Mod, levels = c("untreated", "wildfire", "SODtrt"))
sod_prunereads.ps@sam_data$VegBin <- as.factor(sod_prunereads.ps@sam_data$VegBin)
sod_prunereads.ps@sam_data$SoilBin <- as.factor(sod_prunereads.ps@sam_data$SoilBin)

save(sod_prunereads.ps, file="Data/RPS10/filtered_data/sod_pruned.RData")

readsums<-sort(sample_sums(sod_prunereads.ps))
median(readsums)

#Prune ASVs not present in least 5 times (across one more more samples)
sod_prunereads_5asv.ps<- phyloseq::prune_taxa(taxa_sums(sod_prunereads.ps) > 5, sod_prunereads.ps)
sod_prunereads_5asv.ps<- prune_taxa(taxa_sums(sod_prunereads_5asv.ps) > 0, sod_prunereads_5asv.ps)
save(sod_prunereads_5asv.ps, file="Data/RPS10/filtered_data/sod_prunereads_5asv.RData")
```

### Create prevalance plots (Rel abundance vs. total abundance) facetted by genus
```{r prevalence plots organized by phylum}
prevdf1 = subset(prevdf_check, Genus %in% get_taxa_unique(sod_prunereads_5asv.ps, "Genus"))
ggplot(prevdf1, aes(TotalAbundance, Prevalence / nsamples(sod_prunereads_5asv.ps),color=Genus)) +
  # Include a guess for parameter
  geom_hline(yintercept = 0.005, alpha = 0.5, linetype = 2) +  geom_point(size = 2, alpha = 0.7) +
  scale_x_log10() +  xlab("Total Abundance") + ylab("Prevalence [Frac. Samples]") +
  facet_wrap(~Genus) + theme(legend.position="none")
```

### Let's make histograms of read counts before and after filtering
```{r Reads sequenced by experiment}
#Before filtering reads
sdt = data.table::data.table(as(sample_data(sod.ps), "data.frame"),
                 TotalReads = sample_sums(sod.ps), keep.rownames = TRUE)
pseqDepth = ggplot(sdt, aes(TotalReads)) + geom_histogram(color="black", fill="lightblue") + ggtitle("Histogram of sequencing depth-before filtering samples with low read depths") + labs(x="Depth", y="Frequency")
pseqDepth

#After filtering reads
sdt_filt = data.table::data.table(as(sample_data(sod_prunereads_5asv.ps), "data.frame"),
                 TotalReads = sample_sums(sod_prunereads_5asv.ps), keep.rownames = TRUE)
pseqDepth_filt = ggplot(sdt_filt, aes(TotalReads)) + geom_histogram(color="black", fill="lightblue") + ggtitle("Histogram of sequencing depth after filtering samples with low read depths") + labs(x="Depth", y="Frequency")
pseqDepth_filt
```

### Software used
```{r, software used}
sessionInfo()
```
