---
title: "RPS10 SOD Preparation Steps"
output: html_document
date: "2024-06-11"
---

```{r setup}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(root.dir = "~/Documents/Postdoc/SOD_microbiome_analysis")
library(phyloseq); packageVersion("phyloseq")
library(ggplot2); packageVersion("ggplot2")
library(readr); packageVersion("readr")
library(purrr); packageVersion("purrr")
library(furrr); packageVersion("furrr")
library(dplyr); packageVersion("dplyr")
library(stringr); packageVersion("stringr")
library(metacoder); packageVersion("metacoder")
library(data.table); packageVersion("data.table")
library(decontam); packageVersion("decontam")
library(vegan); packageVersion("vegan")
```

### Load data in order to make phyloseq and taxmap objects
```{r load files, include=TRUE}
#Load DADA2 files
seqtab.nochim <- read.table("Data/RPS10/dada2_outputs/rps10_seqtab_nochim.out",sep="\t",stringsAsFactors=F,header=T)

taxa <- read.table("Data/RPS10/dada2_outputs/rps10_taxa.out",sep="\t",stringsAsFactors=F,header=T)

taxa_filt <- taxa %>%
  mutate(across(starts_with("tax."), ~ ifelse(get(str_replace(cur_column(), "tax.", "boot.")) < 0, NA, .))) %>%
  set_names(str_replace(names(.), "tax.", ""))

samdf <- read.csv("Data/RPS10/metadata_dbs/Metadata_rps10_short_mod.csv")
```

### Exploratory analysis-track reads throughout DADA2 analysis   
```{r track reads through DADA2 analysis, include=TRUE}
track <- read.table("Data/RPS10/dada2_outputs/rps10_trackreads.out", sep="\t", stringsAsFactors = F, header=T)
```

### Format various dataframes so that we can convert to phyloseq obj
```{r format dataframes, include=TRUE}
all(rownames(seqtab.nochim) %in% samdf$Sample_nameRps10)
rownames(samdf) <- samdf$Sample_nameRps10

samdata.ps <- sample_data(samdf)

taxa.ps <- tax_table(as.matrix(taxa_filt))

seqtab.nochim.ps <- otu_table(seqtab.nochim, taxa_are_rows=FALSE)

ps <- phyloseq(seqtab.nochim.ps, samdata.ps, taxa.ps)

#Add refseq slot
dna <- Biostrings::DNAStringSet(taxa_names(ps))
names(dna) <- taxa_names(ps)
ps <- merge_phyloseq(ps, dna)
taxa_names(ps) <- paste0("ASV", seq(ntaxa(ps)))
```

### Make ASV matrix to better examine results
```{r load files standard workflow rps10}
taxa_df <- as.data.frame(taxa_filt)
taxa_df$sequence <- rownames(taxa_df)
taxa_df <- taxa_df[, c("sequence", setdiff(names(taxa_df), "sequence"))]
rownames(taxa_df) <- NULL

seqtab.nochim <- t(seqtab.nochim)
seqtab.nochim_df <- as.data.frame(seqtab.nochim)
seqtab.nochim_df$sequence <- rownames(seqtab.nochim_df)
seqtab.nochim_df <- seqtab.nochim_df[, c("sequence", setdiff(names(seqtab.nochim_df), "sequence"))]
rownames(seqtab.nochim_df) <- NULL

#Combine matrices
combined_df <- merge(seqtab.nochim_df, taxa_df, by = "sequence", all = TRUE)

taxonomic_cols <- c("Kingdom", "Phylum", "Class", "Order", 
                    "Family", "Genus", "Species", "Reference")

bootstrap_cols <- c("boot.Kingdom", "boot.Phylum", "boot.Class", "boot.Order", 
                    "boot.Family", "boot.Genus", "boot.Species", "boot.Reference")

dada2_tax <- mapply(function(tax, boot, level) {
  paste(tax, boot, level, sep = "--")
}, combined_df[, taxonomic_cols], combined_df[, bootstrap_cols], 
   c("Kingdom", "Phylum", "Class", "Order", "Family", "Genus", "Species", "Reference"))

combined_df$dada2_tax <- apply(dada2_tax, 1, paste, collapse = ";")

#Remove all other tax and boot columns
#combined_df <- combined_df[, !(names(combined_df) %in% c(taxonomic_cols, bootstrap_cols))]
combined_df <- combined_df[, c("sequence", "dada2_tax", setdiff(names(combined_df), c("sequence", "dada2_tax")))]

#Save file
write.csv(combined_df, "Data/RPS10/filtered_data/rps10_abundance_matrix_0.csv")
```
      
### Decontamination step using decontam package-this was most necessary for RPS10 analysis.
```{r decontam, include=TRUE}
df <- as.data.frame(sample_data(ps)) # Put sample_data into a ggplot-friendly data.frame
df$LibrarySize <- sample_sums(ps)
df <- df[order(df$LibrarySize),]
df$Index <- seq(nrow(df))
ggplot(data=df, aes(x=Index, y=LibrarySize, color=Sample_or_Control)) + geom_point()


sample_data(ps)$is.neg <- sample_data(ps)$Sample_or_Control == "Control"
contamdf.prev <- isContaminant(ps, method="prevalence", neg="is.neg")
table(contamdf.prev$contaminant)
head(which(contamdf.prev$contaminant))

contamdf.prev05 <- isContaminant(ps, method="prevalence", neg="is.neg", threshold=0.5)
table(contamdf.prev05$contaminant)
which(contamdf.prev05$contaminant)

ps.pa <- transform_sample_counts(ps, function(abund) 1*(abund>0))
ps.pa.neg <- prune_samples(sample_data(ps.pa)$Sample_or_Control == "control", ps.pa)
ps.pa.pos <- prune_samples(sample_data(ps.pa)$Sample_or_Control == "sample", ps.pa)

df.pa <- data.frame(pa.pos=taxa_sums(ps.pa.pos), pa.neg=taxa_sums(ps.pa.neg),
                     contaminant=contamdf.prev05$contaminant)
ggplot(data=df.pa, aes(x=pa.neg, y=pa.pos, color=contaminant)) + geom_point() +
  xlab("Prevalence (Negative Controls)") + ylab("Prevalence (True Samples)")

ps.decontam <- prune_taxa(!contamdf.prev05$contaminant, ps)
save(ps.decontam, file = "Data/RPS10/filtered_data/RPS10_phyloseq.decontam.RData")
```


###Take a first glance at the data  
Helpful for deciding how to filter data. Prevalence is how many samples have the ASV, and total abundance is how many times ASV is present across all samples. 
```{r examine ASV Prevalance, include=TRUE}

prevdf = apply(X = phyloseq::otu_table(ps.decontam),
               MARGIN = ifelse(taxa_are_rows(ps.decontam), yes = 1, no = 2),
               FUN = function(x){sum(x > 0)})

prevdf = data.frame(Prevalence = prevdf,
                    TotalAbundance = taxa_sums(ps.decontam),
                    tax_table(ps.decontam))
```

### Subset relevant SOD samples from rest of the dataset  
I will also remove allow abundance ASVs not present in at least one sample and not present at least five times
```{r Subset dataset, include=TRUE}
sod.ps<-subset_samples(ps.decontam, Experiment=="basetree_100plots")
sod.ps<- prune_taxa(taxa_sums(sod.ps) > 0, sod.ps)

readsums<-sort(sample_sums(sod.ps))
median(readsums)

sod_5asv.ps <- prune_taxa(taxa_sums(sod.ps) > 5, sod.ps)
sod_5asv.ps<- prune_taxa(taxa_sums(sod_5asv.ps) > 0, sod_5asv.ps)

save(sod.ps, file="Data/RPS10/filtered_data/sodexp.RData")
save(sod_5asv.ps, file="Data/RPS10/filtered_data/sod_5asv.RData")

readsums<-sort(sample_sums(sod.ps))
median(readsums)
summary(readsums)
```

### Check prevalence after filtering
```{r check prevalence after filtering}
prevdf_check = apply(X = phyloseq::otu_table(sod_5asv.ps),
               MARGIN = ifelse(taxa_are_rows(sod_5asv.ps), yes = 1, no = 2),
               FUN = function(x){sum(x > 0)})

prevdf_check = data.frame(Prevalence = prevdf_check,
                    TotalAbundance = taxa_sums(sod_5asv.ps),
                    tax_table(sod_5asv.ps))
```

### Subset tracked reads so we are just looking at read distributions of SOD samples
```{r subset samples 2}
samples_to_subset <- samdf %>%
  filter(Experiment=="basetree_100plots") %>% 
  pull(Sample_nameRps10)

#track$sample <- rownames(track)
subset_track_reads <- track[rownames(track) %in% samples_to_subset, ]
write.csv(subset_track_reads, file="Data/RPS10/filtered_data/RPS10_subset_track_reads_sod.csv",row.names = TRUE)

summary(subset_track_reads)
```

### Now I will check how many ASV sequences were classified as P.ramorum
```{r check which ASVs were assigned as P. ramorum}
tax_table <- as.data.frame(sod.ps@tax_table)

ramorum_taxa <- apply(tax_table(sod.ps), 1, function(x) any(grepl("Phytophthora_ramorum", x, ignore.case = TRUE)))
print(sum(ramorum_taxa))

ramorum_asvs <- rownames(tax_table(sod.ps))[ramorum_taxa]


otu_table <- as.data.frame(as(otu_table(sod.ps), "matrix"))

ramorum_counts <- otu_table[, ramorum_asvs, drop = FALSE]
print(ramorum_counts)
```

### Remove samples with fewer than 2500 reads  
```{r Filter out samples with few reads}
### Unfortunately, some samples have very few reads. To proceed with analyses, I'll remove any samples with less than 2500 reads. 
#Unfortunately, a subset of samples have less than 3k reads. I will remove these samples for now. 
sod_prunereads.ps = prune_samples(sample_sums(sod.ps)>=2500, sod.ps)
sod_prunereads.ps <- prune_taxa(taxa_sums(sod_prunereads.ps ) > 0, sod_prunereads.ps)
checkcounts=sort(sample_sums(sod_prunereads.ps))

sod_prunereads.ps@sam_data$Treatuntreat_EP<-factor(sod_prunereads.ps@sam_data$Treatuntreat_EP, levels = c(0, 1), labels = c("untreated", "treated"))
sod_prunereads.ps@sam_data$YearSam_EP <- as.factor(sod_prunereads.ps@sam_data$YearSam_EP)

sod_prunereads.ps@sam_data$YrSinceTrt_BR <- as.factor(sod_prunereads.ps@sam_data$YrSinceTrt_BR)
sod_prunereads.ps@sam_data$YrSinceTrt_BR <- factor(sod_prunereads.ps@sam_data$YrSinceTrt_BR, levels = c("untreated", "1to2", "3to4", "5to8", "10to12"))

sod_prunereads.ps@sam_data$YrSinceTrt_Range <- as.factor(sod_prunereads.ps@sam_data$YrSinceTrt_Range)
sod_prunereads.ps@sam_data$YrSinceTrt_Range <- factor(sod_prunereads.ps@sam_data$YrSinceTrt_Range, levels = c("untreated", "<3", ">3"))

sod_prunereads.ps@sam_data$Fire_Mod <- factor(sod_prunereads.ps@sam_data$Fire_Mod, levels = c("untreated", "wildfire", "SODtrt"))
sod_prunereads.ps@sam_data$VegBin <- as.factor(sod_prunereads.ps@sam_data$VegBin)
sod_prunereads.ps@sam_data$SoilBin <- as.factor(sod_prunereads.ps@sam_data$SoilBin)

save(sod_prunereads.ps, file="Data/RPS10/filtered_data/sod_pruned.RData")

readsums<-sort(sample_sums(sod_prunereads.ps))
median(readsums)

sod_prunereads_5asv.ps<- phyloseq::prune_taxa(taxa_sums(sod_prunereads.ps) > 5, sod_prunereads.ps)
sod_prunereads_5asv.ps<- prune_taxa(taxa_sums(sod_prunereads_5asv.ps) > 0, sod_prunereads_5asv.ps)
save(sod_prunereads_5asv.ps, file="Data/RPS10/filtered_data/sod_prunereads_5asv.RData")
```

### Subset read counts so we are just looking at sod samples after filtering out low abund reads
```{r subset tracked reads}
samdf_filt <- data.frame(sod_prunereads.ps@sam_data)

if ("Sample_nameRps10" %in% names(samdf_filt)) {
  # Filter the data frame and pull the samples
  samples_to_subset <- samdf_filt %>%
    filter(Experiment == "basetree_100plots") %>%
    pull(Sample_nameRps10)
  
  track <- as.data.frame(track)
  track$sample <- rownames(track)
  
  subset_track_reads <- track[rownames(track) %in% samples_to_subset, ]
  
  write.csv(subset_track_reads, file = "Data/RPS10/filtered_data/RPS10_subset_track_reads_filt.csv", row.names = TRUE)

  summary(subset_track_reads)
} else {
  stop("Column 'Sample_nameRps10' does not exist in samdf_filt.")
}

```

### Check read counts
```{r load sod data}
read_counts<-read.csv("Data/RPS10/filtered_data/RPS10_subset_track_reads_sod.csv")
summary(read_counts)

filtered_read_counts<-read.csv("Data/RPS10/filtered_data/RPS10_subset_track_reads_filt.csv")
summary(filtered_read_counts)
```

### Check prevalence after filtering out low read samples and ASVs not present at least 5 times
```{r check prevalence after filtering 2}
prevdf_check = apply(X = phyloseq::otu_table(sod_prunereads_5asv.ps),
               MARGIN = ifelse(taxa_are_rows(sod_prunereads_5asv.ps), yes = 1, no = 2),
               FUN = function(x){sum(x > 0)})

prevdf_check = data.frame(Prevalence = prevdf_check,
                    TotalAbundance = taxa_sums(sod_prunereads_5asv.ps),
                    tax_table(sod_prunereads_5asv.ps))
```

### Create prevalance plots (Rel abundance vs. total abundance) faceted by Genus
```{r prevalence plots faceted by genus}
prevdf1 = subset(prevdf_check, Genus %in% get_taxa_unique(sod_prunereads_5asv.ps, "Genus"))
ggplot(prevdf1, aes(TotalAbundance, Prevalence / nsamples(sod_prunereads_5asv.ps),color=Genus)) +
  geom_hline(yintercept = 0.005, alpha = 0.5, linetype = 2) +  geom_point(size = 2, alpha = 0.7) +
  scale_x_log10() +  xlab("Total Abundance") + ylab("Prevalence [Frac. Samples]") +
  facet_wrap(~Genus) + theme(legend.position="none")
```

### Create prevalance plots (Rel abundance vs. total abundance) faceted by Genus. ASVs not present at least 5 times were retained
```{r prevalence plots organized by genus-data not filtered}
prevdf_check_nofilt = apply(X = phyloseq::otu_table(sod_prunereads.ps),
               MARGIN = ifelse(taxa_are_rows(sod_prunereads.ps), yes = 1, no = 2),
               FUN = function(x){sum(x > 0)})

prevdf_check_nofilt = data.frame(Prevalence = prevdf_check_nofilt,
                    TotalAbundance = taxa_sums(sod_prunereads.ps),
                    tax_table(sod_prunereads.ps))
```

### Create prevalance plots (Rel abundance vs. total abundance) faceted by Genus. ASVs not present at least 5 times were retained.
```{r prevalence plots organized by genus}
prevdf1 = subset(prevdf_check_nofilt, Genus %in% get_taxa_unique(sod_prunereads.ps, "Genus"))

pdf("Figures/RPS10/exploratoryplot_abundance.pdf", width = 10, height = 8)
ggplot(prevdf1, aes(TotalAbundance, Prevalence / nsamples(sod_prunereads.ps),color=Genus)) +
  # Include a guess for parameter
  geom_hline(yintercept = 0.005, alpha = 0.5, linetype = 2) +  geom_point(size = 2, alpha = 0.7) +
  scale_x_log10() +  xlab("Total Abundance") + ylab("Prevalence [Frac. Samples]") +
  facet_wrap(~Genus) + theme(legend.position="none")
dev.off()
```

### Let's make histograms of read counts before and after filtering out samples with few reads
```{r Reads sequenced by experiment}
#Before filtering samples with fewer than 2500 reads
sdt = data.table::data.table(as(sample_data(sod.ps), "data.frame"),
                 TotalReads = sample_sums(sod.ps), keep.rownames = TRUE)
pseqDepth = ggplot(sdt, aes(TotalReads)) + geom_histogram(color="black", fill="lightblue") + ggtitle("Histogram of sequencing depth-before filtering samples with low read depths") + labs(x="Depth", y="Frequency")
pseqDepth

#After filtering out samples with few reads
sdt_filt = data.table::data.table(as(sample_data(sod_prunereads.ps), "data.frame"),
                 TotalReads = sample_sums(sod_prunereads.ps), keep.rownames = TRUE)
pseqDepth_filt = ggplot(sdt_filt, aes(TotalReads)) + geom_histogram(color="black", fill="lightblue") + ggtitle("Histogram of sequencing depth after filtering samples with low read depths") + labs(x="Depth", y="Frequency")
pseqDepth_filt
```

### Let's make rarefaction curves 
```{r Rarefaction curves}
#Before filtering out ASVs not present at least 5 times
otu_data <- as(otu_table(sod_prunereads.ps), "matrix")

S <- specnumber(otu_data) # observed number of species
raremax <- min(rowSums(otu_data))

pdf("Figures/RPS10/rarefaction_curve.pdf", width = 10, height = 8)
rarecurve(otu_data, step = 10, sample = raremax, col = "blue", cex = 0.6, label = FALSE, xlab = "Sequencing Depth", ylab = "Number of ASVs")
dev.off()

pdf("Figures/RPS10/speciesaccum_curve.pdf", width = 10, height = 8)
asv_accum <- vegan::specaccum(otu_data, method = "exact")
plot(asv_accum, ci.type = "poly", col = "blue", lwd = 2, ci.lty = 0, ci.col = "lightblue", xlab = "Number of samples", ylab = "Number of Filtered ASVs")
dev.off()
```

### Software used
```{r, software used}
sessionInfo()
```