---
title: "RPS10 SOD Experiments"
output: html_document
date: "2024-05-21"
---

```{r setup, include=TRUE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(root.dir = "~/Documents/Postdoc/SOD_microbiome_analysis")
library(phyloseq); packageVersion("phyloseq")
library(ggplot2); packageVersion("ggplot2")
library(readr); packageVersion("readr")
library(purrr); packageVersion("purrr")
library(furrr); packageVersion("furrr")
library(dplyr); packageVersion("dplyr")
library(stringr); packageVersion("stringr")
library(metacoder); packageVersion("metacoder")
library(Biostrings); packageVersion("Biostrings")
library(vegan);packageVersion("vegan")
library(agricolae);packageVersion("agricolae")
library(magrittr);packageVersion("magrittr")
library(data.table);packageVersion("data.table")
library(ggbeeswarm);packageVersion("ggbeeswarm")
library(DESeq2);packageVersion("DESeq2")
library(cowplot);packageVersion("cowplot")
library(doParallel);packageVersion("doParallel")
library(viridis);packageVersion("viridis")
library(ggpubr); packageVersion("ggpubr")
library(geosphere); packageVersion("geosphere")
theme_set(theme_cowplot())
```

### Load data
```{r subset data}
load("Data/RPS10/filtered_data/sod_pruned.RData")
load("Data/RPS10/filtered_data/sod_prunereads_5asv.RData")

filtered_read_counts<-read.csv("Data/RPS10/filtered_data/RPS10_subset_track_reads_sod.csv")
summary(filtered_read_counts)

new_names <- sapply(strsplit(sod_prunereads_5asv.ps@sam_data$Sample_nameRps10, "-"), function(x) {
  substr(x[1], 3, nchar(x[1]))
})

sod_prunereads_5asv.ps@sam_data$Sample_nameRps10 <- new_names

sod_prunereads.ps@sam_data$Fire_Mod <- factor(sod_prunereads.ps@sam_data$Fire_Mod, 
                                              levels = c("untreated","SODtrt","wildfire"), 
                                                  labels=c("Untreated","Treated","Wildfire"))

sod_prunereads_5asv.ps@sam_data$Fire_Mod <- factor(sod_prunereads_5asv.ps@sam_data$Fire_Mod, 
                                              levels = c("untreated","SODtrt","wildfire"), 
                                                  labels=c("Untreated","Treated","Wildfire"))
```
### Bar plot agglomerated by order-facetted by fire factor
```{r barplot-Relative abundance-fire}
RPS10data <- sod_prunereads_5asv.ps %>%
  transform_sample_counts(function(x) {x / sum(x)}) %>% 
  tax_glom(taxrank = "Order") %>% 
  psmelt() %>% 
  #mutate(across(Kingdom:Species, ~ ifelse(is.na(.), "Unclassified", .))) %>%  
  arrange(desc(Abundance))

abund_plot <- ggplot(RPS10data, aes(x = Sample_nameRps10, y = Abundance, fill = Order)) + 
  geom_bar(stat = "identity", position = "stack", color = "black", size = 0.2) +
  scale_fill_viridis_d() +
  scale_color_viridis_d() +
  theme_minimal() +
  labs(
    y = "Relative Abundance",
    fill = "Order"
  ) +
  facet_grid(~factor(Fire_Mod), scales = "free_x") +
  theme(
    axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5, size = 16),
    axis.text.y = element_text(size = 20),  # Increase y-axis label size
    axis.title.x = element_blank(),  # Remove the x-axis title if needed
    axis.title.y = element_text(size = 20), 
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    legend.position = "bottom",
    legend.text = element_text(size = 20),
    legend.title = element_text(size = 20, face = "bold"),
    legend.key.size = unit(1, "lines"),
    strip.text = element_text(size = 24, face= "bold"),
    strip.background = element_blank()
  ) +
  guides(
    fill = guide_legend(
      reverse = FALSE,
      keywidth = 1,
      keyheight = 1,
      title.position = "top",
      title.hjust = 0.5,
      label.theme = element_text(size = 20)
    )
  )

abund_plot <- abund_plot + 
  scale_x_discrete(labels = function(x) gsub("Sample_nameRps10", "", x))  # Remove unwanted text

abund_plot

# Save the plot
ggsave("Figures/RPS10/relabundance_plot_fire_relabund_order_wildfire_samples.pdf", abund_plot, width = 24, height = 10)
```

### Let's summarize with boxplot 
```{r, boxplot summary}
dataRPS10 <- sod_prunereads_5asv.ps %>%
  tax_glom(taxrank = "Genus") %>% 
  transform_sample_counts(function(x) {x / sum(x)}) %>% 
  psmelt() %>% 
  filter(Abundance > 0.05)

combined_plot <- ggplot(dataRPS10, aes(x = Fire_Mod, y = Abundance, fill = Genus)) +
  geom_boxplot(outlier.shape = NA, alpha = 0.5) +  # Boxplot with no outliers and some transparency
  geom_jitter(color = "black", width = 0.2, size = 2) +  # Jittered black dots
  scale_fill_viridis_d(option="H") +
  theme_minimal() +
  labs(
    y = "Relative Abundance"
  ) +
  theme(
    axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5, size = 20, face = "bold"),
    axis.text.y = element_text(size = 24),
    axis.title.x = element_blank(),
    axis.title.y = element_text(size = 24),
    legend.position = "none",  # Remove the legend
    strip.text = element_text(size = 24, face= "bold"),
    strip.background = element_blank()
  ) +
  facet_wrap(~ Genus, scales = "free_y") +
  stat_compare_means(method = "anova", label.y = 40)

# Save the plot
ggsave("Figures/RPS10/relabundance_plot_combined_facet_boxplot_genus_wildfire_samples.pdf", combined_plot, width = 30, height = 16)
```

### Let's look top taxa at family level
```{r Top taxa family}
top40 <- names(sort(taxa_sums(sod_prunereads_5asv.ps), decreasing=TRUE))[1:40]
ps.top40 <- transform_sample_counts(sod_prunereads_5asv.ps, function(OTU) OTU/sum(OTU))
#ps.top40@tax_table[is.na(ps.top40@tax_table)] <- "Unclassified"
ps.top40 <- tax_glom(ps.top40, taxrank = "Family")
ps.top40 <- prune_taxa(top40, ps.top40)

pdf("Figures/RPS10/top40_family_sod_wildfire_samples.pdf", width = 34, height = 20)
plot_bar(ps.top40, x="Sample_nameRps10", fill="Family") + facet_wrap(~Treatuntreat_EP, scales="free_x")
dev.off()
```


### Let's look top taxa at genus level
```{r Top taxa genus}
top40 <- names(sort(taxa_sums(sod_prunereads_5asv.ps), decreasing=TRUE))[1:40]
ps.top40 <- transform_sample_counts(sod_prunereads_5asv.ps, function(OTU) OTU/sum(OTU))
#ps.top40@tax_table[is.na(ps.top40@tax_table)] <- "Unclassified"
ps.top40 <- tax_glom(ps.top40, taxrank = "Genus")
ps.top40 <- prune_taxa(top40, ps.top40)

pdf("Figures/RPS10/top40_genus_sod_wildfire_samples.pdf", width = 34, height = 20)
plot_bar(ps.top40, x="Sample_nameRps10", fill="Genus") + facet_wrap(~Treatuntreat_EP, scales="free_x")
dev.off()
```

### Make data table of top taxa
```{r make data tables-top taxa}
data <-
  sod_prunereads_5asv.ps %>%
  tax_glom("Species") %>%
  psmelt() %>%
  as_tibble()

data_summed <- data %>%
  filter(Abundance > 0) %>%
  group_by(Fire_Mod, Kingdom, Phylum, Class, Order, Family, Genus, Species) %>%
  summarise(Abundance = sum(Abundance),
            Prevalence = n_distinct(Sample)) %>%
  ungroup() %>%
  mutate(Total_Abundance = sum(Abundance)) %>%
  mutate(Percentage = 100*(Abundance / Total_Abundance))
  
write.csv(data_summed, file="Results/RPS10/mostabundanttaxa_abundance_species_wildfire_samples.csv",row.names = TRUE)

data1 <-
  sod_prunereads_5asv.ps %>%
  tax_glom("Genus") %>%
  psmelt() %>%
  as_tibble()

data_summed1 <- data1 %>%
  filter(Abundance > 0) %>%
  group_by(Fire_Mod, Phylum, Class, Order, Family, Genus) %>%
  summarise(Abundance = sum(Abundance),
            Prevalence = n_distinct(Sample)) %>%
  ungroup() %>%
  mutate(Total_Abundance = sum(Abundance)) %>%
  mutate(Percentage = 100*(Abundance / Total_Abundance))
  
write.csv(data_summed1, file="Results/RPS10/mostabundanttaxa_abundance_genus_wildfire_samples.csv",row.names = TRUE)
```

### Let's look at spacial placement of different plots
```{r Map exploration}
pdf("Figures/RPS10/location_of_plots_wildfire_samples.pdf", width = 20, height = 12)
map_plot <- ggplot(sod_prunereads_5asv.ps@sam_data, aes(x = Longitude, y = Latitude)) +
  geom_text(aes(color = Fire_Mod, label = Sample_nameRps10), size = 2.5, 
            check_overlap = TRUE) +
  labs(x = "Longitude", y = "Latitude", title = "Sampling Locations") +
  theme_minimal() +
  theme(
    plot.title = element_text(size = 20), 
    axis.title = element_text(size = 16),
    axis.text = element_text(color = "black"),
    legend.title = element_text(size = 14),
    legend.text = element_text(size = 12)
  )
print(map_plot)
dev.off()
```

### Let's normalize our data in preparation for beta diversity analyses
```{r Normalization}
sod_prunereads_5asv.ps<- phyloseq::prune_taxa(taxa_sums(sod_prunereads.ps) > 5, sod_prunereads.ps)
sdt = data.table::data.table(as(sample_data(sod_prunereads_5asv.ps), "data.frame"),
                 TotalReads = sample_sums(sod_prunereads_5asv.ps), keep.rownames = TRUE)
min_reads<-min(sdt$TotalReads)

rarefied.ps <- rarefy_even_depth(sod_prunereads_5asv.ps, sample.size = min_reads, rngseed = 100)

strat_rel.ps<-transform_sample_counts(sod_prunereads_5asv.ps, function(x) x / sum(x))
```

### Alpha diversity-use non-normalized or pruned data
Let's start with t-test as point of comparison with anova analysis
```{r alpha diversity analyses-t-test}
set.seed(1)
AD <- estimate_richness(sod_prunereads.ps, measures = c("Observed", "InvSimpson", "Shannon"))
rownames(AD) <- gsub("\\.", "-", rownames(AD))

AD <- cbind(AD, sample_data(sod_prunereads.ps)[row.names(AD), c("Treatuntreat_EP", "Lineage_EP", "SoilBin", "VegBin")])

AD <- data.table(AD, keep.rownames = TRUE)

AD <- melt(AD, id.vars = c('rn', c("Treatuntreat_EP", "Lineage_EP", "SoilBin", "VegBin")))

factors <- c("Treatuntreat_EP", "Lineage_EP", "SoilBin", "VegBin")
ttest_list <- list()

for (factor_name in factors) {
  ttest <- data.table(alpha_index = character(), pvalue = numeric())
  for (alphad in c("InvSimpson", "Shannon", "Observed")) {
    ttest_res <- t.test(value ~ get(factor_name), data = AD[variable == alphad], var.equal = TRUE)
    p_value <- ttest_res$p.value
    ttest <- rbind(ttest, data.table(alpha_index = alphad, pvalue = p_value))
  }
  
  ttest_list[[factor_name]] <- ttest
}

for (factor_name in factors) {
  cat(paste("T-Test Results for Factor:", factor_name), "\n")
  print(knitr::kable(ttest_list[[factor_name]], format = "markdown"))
  cat("\n")
}
``` 

### Let's now proceed with ANOVA analyses and then compile our alpha diversity stats
```{r compiled alpha diversity table}
set.seed(1)
anova_list <- list()

AD <- estimate_richness(sod_prunereads.ps, measures = c("Observed", "InvSimpson", "Shannon"))
rownames(AD) <- gsub("\\.", "-", rownames(AD))

AD <- cbind(AD, sample_data(sod_prunereads.ps)[row.names(AD), c("Treatuntreat_EP", "Fire_Mod", "YrSinceTrt_BR")])

AD <- data.table(AD, keep.rownames = TRUE)

AD <- melt(AD, id.vars = c('rn', c("Treatuntreat_EP", "Fire_Mod", "YrSinceTrt_BR")))

factors2 <- c("Treatuntreat_EP", "Fire_Mod", "YrSinceTrt_BR")

plotted_factors <- character()
 
final_anova_table <- data.table(factor_name = character(), alpha_index = character(), pvalue = numeric(), df_between = numeric(), df_within = numeric(), ss_between = numeric(), ss_within = numeric(), ms_between = numeric(), ms_within = numeric(), F_value = numeric())

for (factor_name in factors2) {
  anova <- data.table(factor_name = character(), alpha_index = character(), pvalue = numeric(), df_between = numeric(), df_within = numeric(), ss_between = numeric(), ss_within = numeric(), ms_between = numeric(), ms_within = numeric(), F_value = numeric())
  
  for (alphad in c("Observed","InvSimpson", "Shannon")) {
    anova_res <- aov(value ~ get(factor_name), data = AD[variable == alphad])
    anova_summary <- summary(anova_res)
    p_value <- anova_summary[[1]]$`Pr(>F)`[1]
    df_between <- anova_summary[[1]]$"Df"[1]
    df_within <- anova_summary[[1]]$"Df"[2]
    ss_between <- anova_summary[[1]]$"Sum Sq"[1]
    ss_within <- anova_summary[[1]]$"Sum Sq"[2]
    ms_between <- anova_summary[[1]]$"Mean Sq"[1]
    ms_within <- anova_summary[[1]]$"Mean Sq"[2]
    F_value <- anova_summary[[1]]$"F value"[1]
    
    anova <- rbind(anova, data.table(factor_name = factor_name, alpha_index = alphad, pvalue = p_value, df_between = df_between, df_within = df_within, ss_between = ss_between, ss_within = ss_within, ms_between = ms_between, ms_within = ms_within, F_value = F_value))
  }
  
  final_anova_table <- rbind(final_anova_table, anova)
}

print(knitr::kable(final_anova_table, format = "markdown"))
write.csv(final_anova_table, "Results/RPS10/final_anova_table_wildfire_samples.csv", row.names = FALSE)
```

### Let's make faceted boxplots for our alpha diversity analyses
```{r alpha diversity faceted plots}
set.seed(1)
AD <- estimate_richness(sod_prunereads.ps, measures = c("Observed", "InvSimpson", "Shannon"))
rownames(AD) <- gsub("\\.", "-", rownames(AD))

AD <- cbind(AD, sample_data(sod_prunereads_5asv.ps)[rownames(AD), c("Treatuntreat_EP", "Fire_Mod", "YrSinceTrt_BR")])

variables <- c("Treatuntreat_EP", "Fire_Mod", "YrSinceTrt_BR")

plots <- list()

for (variable in variables) {
  p <- plot_richness(sod_prunereads.ps, x = variable, measures = c("Shannon", "InvSimpson", "Observed")) + 
    geom_boxplot()
  plots[[variable]] <- p
}

plots_arrange <- gridExtra::grid.arrange(
  plots[["Treatuntreat_EP"]], plots[["Fire_Mod"]], plots[["YrSinceTrt_BR"]],ncol=1)

plots_arrange_with_text <- cowplot::ggdraw() +
  cowplot::draw_plot(plots_arrange) +
  cowplot::draw_plot("Treatuntreat_EP", x = 0.28, y = 0.98) +
  cowplot::draw_plot("Fire_EP", x = 0.28, y = 0.98) +
  cowplot::draw_plot("YrSinceTrt_BR", x = 0.28, y = 0.98) 

ggsave(plots_arrange_with_text, path = "Figures/RPS10", filename = "alpha_diversity_each_factor_sod_wildfire_samples.pdf", 
       width = 16, height = 24)
```

### Let's perform TUKEY HSD tests
```{r alpha diversity compilation with tukey summary}
# Define your variables of interest
set.seed(1)
variables <- c("Treatuntreat_EP", "Fire_Mod", "YrSinceTrt_BR")
measures <- c("Observed", "InvSimpson", "Shannon")

tukey_results <- list()

for (variable in variables) {
  for (measure in measures) {
    data_subset <- AD[, c(variable, measure)]
    formula <- as.formula(paste(measure, "~ ."))
    anova_res <- aov(formula, data = data_subset)
    tukey_result <- HSD.test(anova_res, variable)
    print(tukey_result)
  }
}
```

### Let's visualize alpha diversity results with rarefaction curve
```{r, rarefaction curve}
sample_data <- data.frame(sample_data(sod_prunereads.ps))
otu_data <- as(otu_table(sod_prunereads.ps), "matrix")
sample_data$Fire_Mod <- factor(sample_data$Fire_Mod, levels = c("Untreated", "Treated", "Wildfire"))

colors <- ifelse(sample_data$Fire_Mod == "Untreated", "blue",
                 ifelse(sample_data$Fire_Mod == "Treated", "red", "green"))

raremax <- min(rowSums(otu_data))

pdf("Figures/RPS10/rarefaction_curve_wildfire_samples.pdf", width = 10, height = 8)
rarecurve(otu_data, step = 10, sample = raremax, col = colors, cex = 0.6, label = FALSE, xlab = "Sequencing Depth", ylab = "Number of ASVs")
legend("topright", legend = levels(sample_data$Fire_Mod), col = c("blue", "red", "green"), lty = 1, title = "Treatuntreat_EP")
dev.off()

sod_prunereads.ps@sam_data$Fire_Mod 
```

### Beta diversity-PERMANOVA analyses-ASV counts transformed (rel. abundance)
```{r PERMANOVA ANALYSES-relative abundance, include=TRUE}
set.seed(1)
mapping <- as.data.frame(sample_data(strat_rel.ps))

mapping2=mapping[, c("Treatuntreat_EP", "Fire_Mod", "YrSinceTrt_BR")]

ASV_tables_bray <- phyloseq::distance(strat_rel.ps, method = "bray")

df_metadata <- data.frame(sample_data(strat_rel.ps))

df_metadata<-df_metadata[, c("Treatuntreat_EP", "Fire_Mod", "YrSinceTrt_BR")]

pvals <- sapply(colnames(mapping2), function(variable){
  form1<-as.formula(paste("ASV_tables_bray", variable, sep="~"))
  res <- adonis2(form1, data = df_metadata, permutations = 9999)
  pval <- res$`Pr(>F)`[[1]] 
  return(pval)
})

pval_adj <- p.adjust(pvals, method = "fdr")
save.pval.adonis0.05_braycurtis <- pval_adj[pval_adj < 0.05]
save.pval.adonis0.05_braycurtis

print(pval_adj)
```

### Compile PERMANOVA stats-main effects
```{r compile interaction stats}
set.seed(1)
res <- adonis2(ASV_tables_bray ~ Fire_Mod, data = df_metadata, permutations = 9999, by="terms")
res_data <- data.frame(res)
colnames(res_data)[colnames(res_data) == "Pr..F."] <- "pr_f"
res_data <- res_data[-c(nrow(res_data)-1, nrow(res_data)),]

res_data$FDR_adj_p <- p.adjust(res_data$pr_f, method = "fdr")
res_data$Bonferroni_adj_p <- p.adjust(res_data$pr_f, method = "bonferroni")
res_data$Factor <- rownames(res_data)

res_data <- res_data[, c("Factor", "Df", "SumOfSqs", "R2", "F", "pr_f", "FDR_adj_p", "Bonferroni_adj_p")]
rownames(res_data) <- NULL

beta_disp_results <- lapply(c("Treatuntreat_EP", "Fire_Mod"), function(variable) {
  res_betadisper <- betadisper(d = ASV_tables_bray, group = df_metadata[[variable]])
  perm_test_res <- permutest(res_betadisper)
  pval_betadisper <- perm_test_res$tab$`Pr(>F)`[[1]]
  
  data.frame(Factor = variable, P_Value_BetaDispersion = pval_betadisper)
})

beta_disp_results_df <- do.call(rbind, beta_disp_results)

final_results <- merge(res_data, beta_disp_results_df, by = "Factor", all.x = TRUE)

final_results <- final_results[, c("Factor", "Df", "SumOfSqs", "R2", "F", "pr_f", "FDR_adj_p", "Bonferroni_adj_p", "P_Value_BetaDispersion")]

print(final_results, row.names = FALSE)

write.csv(final_results, file = "Results/RPS10/PERMANOVA_with_betadisper_maineffects_wildfire_samples.csv", row.names = FALSE)
```

### Let's now examine beta diversity through ordination plots
We'll will examine outputs for rarefied data as well as data adjusted by relative abundance
We'll now make ordination plots on data for which we've calculated relative abundance
```{r Beta diversity-Rel. abundance-treated and untreated samples}
set.seed(1)
ps_NMDS = ordinate(strat_rel.ps, 
                   method="NMDS", 
                   distance="bray")

plot<- plot_ordination(strat_rel.ps, 
                         ps_NMDS,
                         color= "Treatuntreat_EP") +
  ggtitle("NMDS ordination plot - relative abundance") +
  labs(color = "Factor") + 
  theme(axis.text=element_text(size=14), axis.title=element_text(size=14), title =  element_text(size=13), legend.text = element_text(size = 18))

plot_ellipses<-plot +
   stat_ellipse(type = "norm", linetype = 2) +
  theme_bw()

plot_ellipses

write.csv(ps_NMDS[["points"]], file="Results/RPS10/RPS10_NMDS_coordinates_persample_wildfire_samples.csv",row.names = TRUE)
```

### Now, we'll look at ordination plot (rel abundance dataset), with Fire_Mod factor
```{r Beta diversity-Rel. abundance-Fire_Mod}
set.seed(1)
ps_NMDS = ordinate(strat_rel.ps, 
                   method="NMDS",
                   distance="bray")

plot<- plot_ordination(strat_rel.ps, 
                         ps_NMDS,
                         labs(color = "Factor"),
                         color= "Fire_Mod") +
  ggtitle("NMDS ordination plot - relative abundance") +
  theme(axis.text=element_text(size=14), axis.title=element_text(size=14), title =  element_text(size=13), legend.text = element_text(size = 18))


plot_ellipses<-plot +
   stat_ellipse(type = "norm", linetype = 2) +
  theme_bw()

plot_ellipses
```

### Now, we'll look at ordination plot rel abundance-with YrSinceTrt_BR factor
```{r Beta diversity-Rel. abundance-YrSinceTrt_BR}
set.seed(1)
ps_NMDS = ordinate(strat_rel.ps, 
                   method="NMDS",
                   distance="bray")

plot<- plot_ordination(strat_rel.ps, 
                         ps_NMDS,
                         labs(color = "Factor"),
                         color= "YrSinceTrt_BR") +
  ggtitle("NMDS ordination plot - relative abundance") +
  theme(axis.text=element_text(size=14), axis.title=element_text(size=14), title =  element_text(size=13), legend.text = element_text(size = 18))


plot_ellipses<-plot +
   stat_ellipse(type = "norm", linetype = 2) +
  theme_bw()

plot_ellipses
```
### Software used
```{r software used}
sessioninfo::session_info()
```